---
title: "Welcome to Mass. Numbers"
description: |
  Welcome to the Mass. Numbers blog. Mass. Numbers takes
  a quantiative look as Massachusetts politics and policy.
author:
  - name: Brent Benson
    url: https://massnumbers.us/brentbenson
date: 07-05-2022
citation: false
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)

library(tidyverse)
library(sf)
library(tmap)
```

```{r city-town-map-setup, fig.cap="Partisan Voter Index (PVI) ratings for MA municipalities"}
ma_towns_dems <- read_csv("../../_data/2020/ma_city_town_demographics_pvi_2020.csv") %>%
    mutate(city_town = str_replace(city_town, "Manchester-by-the-Sea", "Manchester")) %>%
    rename(white_pct = race_white_pct,
           black_pct = race_black_pct,
           hispanic_pct = race_hispanic_pct,
           asian_pct = race_asian_pct) %>%
    select(city_town_fips, city_town, PVI, PVI_N, total_population,
           white_pct, black_pct, hispanic_pct, asian_pct)

ma_city_town <- read_sf("../../_gis/townssurvey_shp/TOWNSSURVEY_POLYM.shp", as_tibble=TRUE) %>%
    st_make_valid() %>%
    mutate(city_town = str_to_title(TOWN)) %>%
    left_join(ma_towns_dems, by="city_town")
```

```{r city-town-map}
tmap_mode(mode="plot")

city_town_map <- (
    tm_shape(ma_city_town) +
    tm_borders(lwd=1) + 
    tm_fill(col="PVI_N",
            palette=c("#FF4949", "white", "#0F4392"),
            midpoint=0)
    )
city_town_map    
```

## Welcome

The Mass. Numbers blog began in November of 2012 with [a
post](https://massnumbers.blogspot.com/2012/11/warren-realizes-largest-gains-in.html)
comparing Elizabeth Warren's win over Scott Brown to Martha Coakley's
2010 loss, observing that Warren's largest gains came in working class
cities like Holyoke, Lawrence, Chelsea, and Brockton.

Some of the things that enabled posts like these were [easily
accessible fine-grained election
results](https://electionstats.state.ma.us/) and a nice tool for
creating interactive maps: Google Fusion Tables. As you might have
seen by clicking on the first post, Google stopped supporting Google
Fusion Tables in 2019.

This new version of the Mass. Numbers blog is enabled by some other
new tools. The technologies of
[Rmarkdown](https://rmarkdown.rstudio.com/),
[Distill](https://rstudio.github.io/distill/blog.html), and [GitHub
Pages](https://pages.github.com/) allow for creating data-driven blog
posts that can be compiled into nice looking web pages, but also allow
for including all of the data and analysis that appear in the post.
This is exciting because it reduces the number of steps between
analysis and publishing a result, but also because it promotes
transparency, open data journalism, and reproducible data science.

If you read a post on this blog and are curious about how the analysis
was done, you can look at the [linked GitHub
repository](https://github.com/massnumbers/blog) and see the `.Rmd`
source of the post and the referenced data, checked in and versioned
in the repository.

One other shout-out goes to the [tmap](https://r-tmap.github.io/tmap/)
mapping library that does Google Fusion Tables one better by making it
easy to create data driven maps, both static maps and interactive
maps, with a simple configuration option. The fact that I can create
the maps completely with code allows for a great deal of automation,
reducing the amount of manual work to create a map, and resulting in
something that is documented and reproducible.

I am guessing that most readers of this blog are interested in the
political analysis and data visualizations, but I felt that it made
sense to explain the evolution of the technology behind this writing
and to help others who are interested in reproducing the results, or
want to take the analysis in another direction.

I am interested in your feedback. You can send me feedback through my
[`@bwbensonjr`](https://twitter.com/bwbensonjr) twitter, or using the
email address with username `brent` and the domain of this website.
