---
title: "Will Boston neighborhoods play a role in the 10th Suffolk special election?"
description: |
  The three candidates bring a lot to the table including their home base.
author: Brent Benson
date: 2023-04-28
citation: false
preview: https://www.massnumbers.us/images/ma_10th_suffolk_neighborhoods.png
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)

library(metathis)
library(tidyverse)
library(gt)
library(gtExtras)
library(sf)
library(units)
library(tmap)
library(formattable)

tmap_mode(mode="view")
```

```{r post-meta}
meta() %>%
    meta_description("The three 10th Suffolk special election candidates bring a lot to the table including their home base.") %>%
    meta_viewport() %>%
    meta_social(
        title = "How important are neighborhoods in Boston state legislative elections?",
        url = "https://www.massnumbers.us/posts/2023-04-28-suffolk-10th-special/",
        image = "https://www.massnumbers.us/images/ma_10th_suffolk_neighborhoods.png",
        image_alt = "The neighborhoods of the 10th Suffolk State Rep district",
        og_type = "article",
        og_author = "Brent Benson",
        twitter_card_type = "summary_large_image",
        twitter_creator = "@bwbensonjr"
    )
```

Resignations by Massachusetts State Representatives Jon Santiago and
Ed Coppinger have resulted in Special State Primary elections on May
2, 2023 to elect their successors in the South End-based 9th Suffolk
and West Roxbury-dominated 10th Suffolk districts. There are five
Democratic candidates who submitted signatures and qualified for the
two primary ballots.

Amparo "Chary" Ortiz has withdrawn from the 9th Suffolk race leaving
John Moran as the presumptive winner of the South Boston-based
seat. There are three strong candidates for 10th Suffolk seat
Democratic primary, each one bringing different different assets to
the race, including their three distinct neighborhoods.

```{r suffolk-10th-candidates}

suff_10_cands <- tribble(
    ~Candidate, ~Profession, ~Neighborhood,
    "Bill MacGregor", "Loan Officer", "West Roxbury",
    "Robert Orthman", "Attorney", " Roslindale",
    "Celia Segel", "Health Policy Director", "Jamaica Plain")

suff_10_cands %>%
    gt() %>%
    cols_label(Candidate = md("**Candidate**"),
               Profession = "",
               Neighborhood = md("**Neighborhood**")) %>%
    # tab_options(column_labels.hidden=TRUE) %>%
    # tab_header("10th Suffolk State Rep Candidates")
    tab_caption("10th Suffolk State Rep Candidates")

```

## Boston Neighborhoods and their State Rep Districts

```{r prepare-data}
bos_nh <- read_sf("../../_gis/Boston_Neighborhoods.geojson",
                  as_tibble=TRUE) %>%
    st_make_valid() %>%
    st_transform(6491) %>%
    relocate(Name)

summarize_neighborhoods <- function(df) {
    df %>%
        pull(Neighborhood) %>%
        str_c(collapse=",")
}

wps_bos <- read_sf(file.path("../../_gis",
                                "ma_wards_precincts_w_subs.geojson"),
                   as_tibble=TRUE) %>%
    filter(city_town == "Boston") %>%
    select(-city_town_abbrev) %>%
    st_transform(6491)

wps_bos_nh <- wps_bos %>%
    st_intersection(bos_nh) %>%
    mutate(area = st_area(geometry)) %>%
    filter(area > set_units(10000, m^2)) %>%
    select(city_town, Ward, Pct, Neighborhood=Name, area) %>%
    as_tibble() %>%
    group_by(city_town, Ward, Pct) %>%
    mutate(area_pct = area/sum(area)) %>%
    ungroup() %>%
    nest(neighborhood = c(Neighborhood, area, area_pct, geometry)) %>%
    mutate(neighborhoods = map_chr(neighborhood, summarize_neighborhoods),
           num_neighborhoods = map_dbl(neighborhood, nrow))

wps_bos_nh_sr <- wps_bos %>%
    st_intersection(bos_nh) %>%
    mutate(area = st_area(geometry)) %>%
    filter(area > set_units(10000, m^2)) %>%
    select(Neighborhood=Name, State_Rep, city_town, Ward, Pct)
```

```{r boston-neighborhoods-precincts, fig.height=5, layout="l-body-outset"}
(tm_shape(wps_bos_nh_sr) +
 tm_fill(col="Neighborhood", alpha=0.6, popup.vars=TRUE) +
 tm_borders(lwd=1) +
 tm_basemap("OpenStreetMap"))
```

There are 160 State Representatives in the Massachusetts House of
Representatives with 19 Suffolk County-based representatives, each who
represents about 44,000 citizens from Boston or adjacent
communities. Only two of the State Rep districts&mdash;East Boston's
1st Suffolk and Dorchester's 13th Suffolk&mdash;are contained in a
single Boston neighborhood. Every other Suffolk-based State Rep
district crosses neighborhood boundaries.

How important are the neighborhood boundaries to Boston state
legislative races? The 10th Suffolk race is an extreme example where
all three candidates are from different neighborhoods and one
particular neighborhood, West Roxbury, contains over half of the
likely voters.

## 10th Suffolk Neighborhoods

```{r suffolk-10-data}
wps_suff_10 <- read_sf(file.path("../../_gis",
                                 "ma_wards_precincts_w_subs.geojson"),
                       as_tibble=TRUE) %>%
    filter(State_Rep == "Tenth Suffolk") %>%
    select(-city_town_abbrev) %>%
    st_transform(6491)

wpsn_suff_10 <- wps_suff_10 %>%
    st_intersection((bos_nh %>% select(Neighborhood = Name))) %>%
    mutate(area = st_area(geometry)) %>%
    filter(area > set_units(15000, m^2)) %>%
    group_by(city_town, Ward, Pct) %>%
    mutate(num_parts = n(),
           precinct_part = row_number(),
           area_pct = area/sum(area),
           name = if_else(num_parts > 1,
                          str_glue("{name} (part {precinct_part})"),
                          name)) %>%
    ungroup() %>%
    mutate(area_pct = map_dbl(area_pct, ~.x)) %>%
    rbind((wps_suff_10 %>%
           filter(city_town == "Brookline") %>%
           mutate(area = st_area(geometry),
                  area_pct = 1,
                  Neighborhood = "Brookline",
                  num_parts = 1,
                  precinct_part = 1))) %>%
    arrange(name)
```

```{r suffolk-10th-map, layout="l-body-outset"}
(tm_shape(wpsn_suff_10) +
 tm_fill(col="Neighborhood", alpha=0.6, popup.vars=TRUE) +
 tm_borders(lwd=2, lty="dashed") +
 tm_shape(wps_suff_10) +
 tm_borders(lwd=2) +
 tm_basemap("OpenStreetMap"))
```

I will use the votes from the 2022 State Democratic Primary to
estimate the percentage of votes that we might see from each
neighborhood in the May 2nd special primary.

The 10th Suffolk State Representative district has 22 Boston precincts
and a single precinct in Brookline. Two of the Boston precincts are
split between Roslindale and West Roxbury, complicating the estimated
vote breakdown, but we can use the relative size of the neighborhood
parts and adjust the vote totals accordingly (94% to 6%). Boston Ward 20 Precinct
1 is mostly in Roslindale with a small dip into West Roxbury, while
Ward 20 Precinct 7 is more evenly split with 57% of the votes coming
from Roslindale and 43% coming from West Roxbury.

```{r suffolk-10-neighborhood-votes}
compass_changes <- c(
    "N. " = "North ",
    "E. " = "East ",
    "S. " = "South ",
    "W. " = "West "
)

ma_dem_primary_nd <- read_csv("../../_data/2023/ma_2022_dem_primary_pct_results.csv") %>%
    select(city_town = `City/Town`,
           Ward,
           Pct,
           total_votes_cast = `Total Votes Cast`) %>%
    mutate(city_town = str_replace_all(city_town, compass_changes))

ma_dem_primary <- read_csv("../../_data/2023/ma_districts_precincts_2022.csv") %>%
    left_join(ma_dem_primary_nd,
              by=c("city_town", "Ward", "Pct"))

suff_10_dem_primary <- ma_dem_primary %>%
    filter(State_Rep == "Tenth Suffolk") %>%
    select(city_town, Ward, Pct, total_votes_cast)

wpsn_suff_10_dem_primary <- wpsn_suff_10 %>%
    left_join(suff_10_dem_primary, by=c("city_town", "Ward", "Pct")) %>%
    mutate(total_votes_adj = total_votes_cast * area_pct)

wpsn_suff_10_dem_primary %>%
    st_drop_geometry() %>%
    select(name, Neighborhood, total_votes_adj) %>%
    gt() %>%
    tab_style(style=cell_text(weight="bold"),
              locations=cells_column_labels()) %>%
    fmt_number(columns=total_votes_adj, decimals=0) %>%
    cols_label(name="Precinct",
               total_votes_adj="2022 Dem. Primary Votes") %>%
    tab_caption("10th Suffolk Precincts")

```

The final breakdown shows West Roxbury with 57% of the vote,
Roslindale with 25%, Jamaica Plain with 13%, and the single
Brookline precinct giving 5% of the vote.

```{r suff-10-vote-table}
suff_10_nbhd_votes <- wpsn_suff_10_dem_primary %>%
    st_drop_geometry() %>%
    group_by(Neighborhood) %>%
    summarize(total_votes = sum(total_votes_adj)) %>%
    mutate(vote_pct = total_votes/sum(total_votes))

suff_10_nbhd_votes %>%
    gt() %>%
    tab_style(style=cell_text(weight="bold"),
              locations=cells_column_labels()) %>%
    fmt_percent(columns=vote_pct, decimals=0) %>%
    fmt_number(columns=total_votes, decimals=0) %>%
    cols_label(total_votes = "2022 Dem. Primary Votes",
               vote_pct = "% of Votes") %>%
    tab_caption("Likely % of votes by neighborhood")
```

## Neighborhood Advantage

The obvious neighborhood advantage would seem to go to Bill MacGregor
with his home base of Roxbury and his endorsement by departing State
Representative and West Roxbury stalwart Ed Coppinger.

The story for Robert Orthman is more complicated as he is originally
from West Roxbury while currently making his home in
Roslindale. Orthman has also received the mainstream endorsement of
the Boston Globe, while also being endorsed by some progressive and
activist groups.

Celia Segel is at a relative disadvantage with only 13% of the votes
likely to come from Jamaica Plain, but gotten high profile
endorsements from Emily's List, the MTA, and sitting legislators like
Cindy Creem, Becca Rausch, and Julian Cyr.

It is very hard to understand the dynamics and predict an outcome in a
three-way race without high-quality polling, which presumably the
candidates or associated groups have done. As is often said, the poll
that really matters is the one at the ballot box on Tuesday. In the
meantime, we can speculate on how much the influence the home
neighborhood of each candidate will play in the final outcome for the
10th Suffolk legislative seat.

