---
title: "A look at MA State Senate Redistricting"
description: |
    There are now six majority-minority state senate districts in
    Massachusetts but only three give minority voters a robust chance of
    electing their candidate of choice
author: Brent Benson
date: 2022-07-22
citation: false
preview: https://www.massnumbers.us/images/middlesex_and_suffolk_redist.png
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
options(digits=1, knitr.kable.NA="")
library(kableExtra)
library(formattable)
library(tidyverse)
library(sf)
library(tmap)
library(readxl)
library(viridis)
library(tidycensus)
library(tigris)
library(units)
library(nominatimlite)
```

```{r senate-districts}
sen_dist <- read_excel("../../_data/2020/MA_State_Senate_Race_2022.xlsx", sheet="Districts")

sen_cand <- read_csv("../../_data/2020/ma_state_senate_cands_2022_contested_addr.csv") %>%
    rename(Inc = Incumbent) %>%
    mutate(address = str_c(street_address, city_town, "MA", sep=", ")) %>%
    select(-street_address)

sen_geom <- read_sf("../../_gis/SENATE2021/SENATE2021_POLY.shp", as_tibble=TRUE) %>%
    st_make_valid() %>%
    select(Number=SENDISTNUM) %>%
    st_transform(st_crs(6491)) %>%
    left_join(sen_dist, by="Number") %>%
    relocate(District)

sen_geom_12 <- read_sf("../../_gis/senate2012/SENATE2012_POLY.shp", as_tibble=TRUE) %>%
    st_make_valid() %>%
    st_transform(st_crs(6491))

tmap_mode(mode="view")
one_m_sq <- set_units(1, m^2)
two_k_m_sq <- set_units(2000, m^2)
five_k_m_sq <- set_units(5000, m^2)
seven_five_k_m_sq <- set_units(75000, m^2)
```

```{r precinct-setup}
ma_towns <- read_sf("../../_gis/townssurvey_shp/TOWNSSURVEY_POLYM.shp",
                    as_tibble=TRUE) %>%
    st_make_valid() %>%
    mutate(city_town = str_to_title(TOWN)) %>%
    select(city_town) %>%
    st_transform(st_crs(6491))

ward_precinct_name <- function(city_town, ward, precinct) {
    str_c(city_town,
          if_else(ward == "-", "", str_c("Ward ", ward)),
          str_c("Precinct ", precinct),
          sep=" ")
}

add_voting_percents <- function(df) {
    df %>%
        mutate(cvap_white_pct = cvap_2019_white/cvap_2019_total,
               cvap_hispanic_pct = cvap_2019_hispanic/cvap_2019_total,
               cvap_black_pct = cvap_2019_black/cvap_2019_total,
               cvap_asian_pct = cvap_2019_asian/cvap_2019_total,
               cvap_pred_group = case_when(((cvap_black_pct >= 0.4) &
                                       (cvap_hispanic_pct >= 0.4)) ~ "Black_and_Hispanic",
                                  (cvap_black_pct >= 0.4) ~ "Black",
                                  (cvap_hispanic_pct >= 0.4) ~ "Hispanic",
                                  (cvap_asian_pct >= 0.4) ~ "Asian",
                                  TRUE ~ "White"),
               vap_white_pct = vap_white_2020/vap_total_2020,
               vap_hispanic_pct = vap_hispanic_2020/vap_total_2020,
               vap_black_pct = vap_black_2020/vap_total_2020,
               vap_asian_pct = vap_asian_2020/vap_total_2020,
               vap_pred_group = case_when(((vap_black_pct >= 0.4) &
                                       (vap_hispanic_pct >= 0.4)) ~ "Black_and_Hispanic",
                                  (vap_black_pct >= 0.4) ~ "Black",
                                  (vap_hispanic_pct >= 0.4) ~ "Hispanic",
                                  (vap_asian_pct >= 0.4) ~ "Asian",
                                  TRUE ~ "White"))
}

sen_dist_blks <- read_csv("../../_data/2020/ma_blocks_to_sen_dist_2021.csv",
                          col_types=cols("c", "c", "i"))

sen_vtd_demos <- read_csv("../../_data/2020/ma_sen_census_2020_daves.csv",
                          col_types=cols(vtd_fips="c"))

vtd_names <- read_csv("../../_data/2020/ma_vtds_city_town_fips.csv", col_types=cols(.default="c")) %>%
    select(vtd_fips, city_town, Ward, Pct)

vtd_full <- sen_vtd_demos %>%
    select(-c(district_id, district_num, partial_pct)) %>%
    group_by(vtd_fips) %>%
    summarize_all(sum)

vtd_demos <- vtd_names %>%
    left_join(vtd_full, by="vtd_fips") %>%
    add_voting_percents()

vtd_geom <- voting_districts(state=25, cb=TRUE) %>%
    select(vtd_fips = GEOID20) %>%
    st_transform(6491) %>%
    left_join(vtd_demos, by="vtd_fips") %>%
    mutate(name = ward_precinct_name(city_town, Ward, Pct)) %>%
    relocate(name)

block_geom <- blocks(state=25, year=2020) %>%
    select(block_fips = GEOID20) %>%
    st_transform(6491)

vtd_blks <- read_csv("../../_data/2020/ma_blocks_to_vtds.csv",
                     col_types=cols(.default="c")) %>%
    left_join(vtd_names, by="vtd_fips") %>%
    inner_join(block_geom, by="block_fips")

blks_sen_dist_vtd <- vtd_blks %>%
    left_join(sen_dist_blks, by="block_fips")

sen_vtds <- blks_sen_dist_vtd %>%
    nest(blocks = c(block_fips, geometry)) %>%
    group_by(vtd_fips) %>%
    mutate(partial_pct = n() > 1) %>%
    ungroup() %>%
    mutate(geometry = map(blocks, ~st_union(.x$geometry))) %>%
    unnest(geometry) %>%
    select(-blocks)

sen_vtd_comp <- sen_vtds %>%
    left_join(sen_vtd_demos, by=c("vtd_fips", "district_id", "district_num")) %>%
    add_voting_percents() %>%
    st_as_sf()

```

## Redistricting Concepts

The Massachusetts State Senate lines have been redrawn by the
legislature after the 2020 Census as required by the constitution.
For a guide to the redistricting principles used by the state
legislatures, check out the [Executive Summary of the Redistricting
Red
Book](https://www.ncsl.org/Portals/1/Documents/Redistricting/Exec_Summary_2020.pdf),
published by the National Conference of State Legislatures.

The primary responsibility given to redistricters by the constitution
is to ensure equally-sized districts. For state legislatures, the
courts have generally approved of redistricting plans where the
largest and smallest districts are within 5 percent of the target
population. The equal-size mandate results in the shifting of
district lines as populations ebb and flow through the Commonwealth.

The next mandate for the redistricters comes from the Voting Rights
Act (VRA) and relates to race, color, and minority language groups.
While it is illegal to use race and ethnicity as factor in
redistricting in the general case, it is also illegal to dilute
concentrations of a minority voting block, resulting in that voting
block not being able to elect its candidate of choice.

While the short-hand concept of a majority-minority district is often used when
evaluating redistricting plans, it is usually an oversimplification which does
not addequately address minority voting block dilution. In particular, minority
groups do not usually vote as a monolith. A district with 25% black voters and
30% hispanic voters is technically majority-minority, but gives neither the
black voters nor the Hispanic voters the ability to elect their preferred
candidate. (In practice, redistricting bodies can use techniques like
ecological regression to analyze and account for historical voting patterns and
tendencies.)

The newly drawn 2020-2021 Massachusetts State Senate districts
increase the number of majority-minority districts from
three to six, but only three of the districts contain
a minority voting block with enough power to elect their candidate of
choice. It is also clear from looking at the data that there aren't
enough geographically concentrated minority blocks in Massachusetts
to do any better at this time.

## Concentrations of Minority Voting Blocks

Let's first look at the 2,152 voting precincts in Massachusetts by flagging 
any precinct that has a citizen votaing age population (CVAP) for a
minority group that is over 40% of the total citizen voting age population.

We use citizen voting age population because that is the actual
number of people who are eligible to vote, a better determinant of
voting power than total population. When we look at the population
numbers used to equally divide the districts, we will use total
population, as required by the constitution.

```{r vtd-minority-precints}
vtd_demos %>%
    count(cvap_pred_group) %>%
    mutate(Percent = percent(n/sum(n), 0)) %>%
    rename(`Predominant Group`=cvap_pred_group,
           `Precinct Count`=n) %>%
    arrange(desc(`Precinct Count`)) %>%
    kbl(caption="Predominant voting group by voting precinct") %>%
    kable_styling()
```

If we look at the municipality of the 204 voting precincts that have a
critical mass of minority voters by municipality (103+94+5+3) it guides us
towards which precincts might be grouped together into a single district
to prevent dilution of that minority group's voting power.

```{r vtd-group-by-city}
cvap_towns <- vtd_demos %>%
    filter(cvap_pred_group != "White") %>%
    count(city_town, cvap_pred_group)

cvap_towns %>%
    rename(`City/Town`=city_town,
           `Predominant Group`=cvap_pred_group,
           `Precinct Count`=n) %>%
    arrange(desc(`Precinct Count`)) %>%
    kbl(caption="Number of plausibly minority precincts by city/town") %>%
    kable_styling()
```

A statewide map of the precincts with concentrated minority
representation shows the limited number and concentration of such
districts.

```{r vtd-minority-map, layout="l-page", fig.cap="*Predominant group for all MA precincts*"}
color_list <- viridis(4)
group_colors <- c(Black_and_Hispanic=color_list[4],
                  Black=color_list[1],
                  Hispanic=color_list[2],
                  Asian=color_list[3],
                  White=NA)
popup_vars <- c("cvap_pred_group", "cvap_white_pct", "cvap_black_pct",
                "cvap_hispanic_pct", "cvap_asian_pct")

cvap_town_outlines <- ma_towns %>% filter(city_town %in% unique(cvap_towns$city_town))

pred_group_map <- (tm_shape(vtd_geom) +
                       tm_borders(lwd=0.05) +
                       tm_fill(col="cvap_pred_group",
                               alpha=1,
                               title="",
                               palette=group_colors) +
                       tm_shape(cvap_town_outlines) +
                       tm_borders(lwd=0.6) +
                       tm_text(text="city_town", size=0.5, xmod=-1.25) +
                       tm_shape(ma_towns) +
                       tm_borders(lwd=0.7))

tmap_mode(mode="plot")
pred_group_map
```
A closer look at four of the clusters around Springfield, Lawrence, Boston, and
Brockton show the biggest concentration of minority voting blocks. All but
the Brockton/Randolph cluster have enough concentration to plausibly
elect a minority candidate of choice. 

```{r cvap-detail, layout="l-body-outset", fig.cap="*Cities with large minority voting blocks*"}
cvap_detail_map <- function(city_towns, fig_title, legend_show) {
    (tm_shape((vtd_geom %>% filter(city_town %in% city_towns))) +
     tm_borders(lwd=0.5, col="black") +
     tm_fill(col="cvap_pred_group",
             alpha=1,
             title="",
             palette=group_colors) +
     tm_shape((ma_towns %>% filter(city_town %in% city_towns))) +
                       tm_borders(lwd=1) +
                       tm_text(text="city_town", size=0.75, xmod=4) +
     tm_layout(title=fig_title, title.size=0.7, legend.show=legend_show))
}
springfield_cvap_map <- cvap_detail_map(c("Springfield"), "Basis of Hampden", TRUE)
essex_1_cvap_map <- cvap_detail_map(c("Haverhill", "Methuen", "Lawrence"),
                                     "Basis of 1st Essex",
                                     FALSE)
boston_cvap_map <- cvap_detail_map(c("Boston"),
                                   "Basis of 1st/2nd Suffolk",
                                   FALSE)
brockt_rand_cvap_map <- cvap_detail_map(c("Brockton", "Randolph"),
                                        "Basis of 2nd Norfolk and Plymouth",
                                        FALSE)
tmap_arrange(springfield_cvap_map,
             essex_1_cvap_map,
             boston_cvap_map,
             brockt_rand_cvap_map,
             ncol=2)
```

```{r proto-dist-pops}
## springfield_cvap <- ma_vtd_cvap %>% filter(city_town %in% c("Springfield"))
## essex_1_cvap <- ma_vtd_cvap %>% filter(city_town %in% c("Haverhill", "Methuen", "Lawrence"))
## boston_cvap <- ma_vtd_cvap %>% filter(city_town %in% c("Boston"))
## brockt_rand_cvap <- ma_vtd_cvap %>% filter(city_town %in% c("Brockton", "Randolph"))
```

## Districts with Minority Power

The Hampden, First Essex, and Second Suffolk Districts contain a large enough
percentage of a particular minority group to allow for election of a 
canidate of choice by that minority group.

```{r district-map-setup}
color_list_1 <- viridis(5)
group_colors_1 <- c(Black_and_Hispanic=color_list_1[4],
                    Black=color_list_1[1],
                    Hispanic=color_list_1[2],
                    Asian=color_list_1[3],
                    White=color_list_1[5])
popup_vars_1 <- c("cvap_pred_group", "cvap_white_pct", "cvap_black_pct",
                  "cvap_hispanic_pct", "cvap_asian_pct")

senate_geometry <- function(dist_number) {
    sen_geom %>% filter(Number == dist_number)
}

senate_precincts <- function(n) {
    sen_vtd_comp %>%
        filter(district_num == n)
}

senate_map <- function(sen_dist, sen_pcts) {
    (tm_shape(sen_pcts) +
     tm_borders(lwd=0.75) +
     tm_fill(col="cvap_pred_group",
             alpha=0.5,
             title="",
             palette=group_colors_1,
             popup.vars=popup_vars_1) +
    tm_shape(sen_dist) + tm_borders(lwd=3) +
    tm_basemap("OpenStreetMap"))
}
```

## The Hampden District

```{r sen-hampden-map}
sen_3 <- senate_geometry(3)
sen_pcts_3 <- senate_precincts(3)
sen_3_map <- senate_map(sen_3, sen_pcts_3)
tmap_leaflet(sen_3_map)
```

```{r sen-hampden}
summarize_dist <- function(sen_pcts) {
    sen_pcts %>%
        summarize(total_pop = sum(total_pop_2020),
                  cvap_total = sum(cvap_2019_total),
                  cvap_hispanic = sum(cvap_2019_hispanic),
                  cvap_black = sum(cvap_2019_black),
                  cvap_asian = sum(cvap_2019_asian),
                  cvap_white = sum(cvap_2019_white),
                  cvap_hispanic_pct = percent(cvap_hispanic/cvap_total, 0),
                  cvap_black_pct = percent(cvap_black/cvap_total, 0),
                  cvap_asian_pct = percent(cvap_asian/cvap_total, 0),
                  cvap_white_pct = percent(cvap_white/cvap_total, 0)) %>%
         st_drop_geometry()
}

summarize_dist(sen_pcts_3) %>%
    select(total_pop, cvap_total, cvap_hispanic_pct,
           cvap_black_pct, cvap_asian_pct, cvap_white_pct) %>%
    kbl(caption="Hampden District Demographics",
        format.args=list(scientific=FALSE, big.mark=",")) %>%
    kable_styling()
```

The Hampden State Senate District contains 90 percent of Springfield's
precincts and 60 percent of Chicopee's precincts. The district
has about as many Hispanic voters as non-Hispanic white voters,
giving that voting block a chance at electing their preferred
candidate.

The current version of the Hampden district has a similar makeup
and in 2020 Springfield City Councilor Alex Gomez defeated
incumbent State Senator James Welch in the Democratic primary.

A simple ecological regression model that uses Gomez's margin over
Welch as the response variable and the CVAP Hispanic percent and
whether the voter is in Springfield as interacting explanatory
variables ends up explaining 90 percent of the variation in the
margin. Hispanic voters were much more likely to vote for Gomez,
with knowing about Gomez as a Springfield voter also playing
a factor.

## The First Essex District

```{r first-essex-redist, fig.cap="*First Essex District*"}
sen_19 <- senate_geometry(19)
sen_pcts_19 <- senate_precincts(19)
sen_19_map <- senate_map(sen_19, sen_pcts_19)
tmap_leaflet(sen_19_map)
```

```{r sen-first-essex}
summarize_dist(sen_pcts_19) %>%
    select(total_pop, cvap_total, cvap_hispanic_pct,
           cvap_black_pct, cvap_asian_pct, cvap_white_pct) %>%
    kbl(format.args=list(scientific=FALSE, big.mark=",")) %>%
    kable_styling()
```

The First Essex State Senate District contains all of Lawrence, all of
Methuen, and approximately 5 precincts of Haverhill (give or take a
few census blocks). The previous lines of the district did not include
Lawrence and instead included a large number of largely white
surrounding communities. The new lines result in a voting age
population that is 36% White, 56% Hispanic, 10% Black, and 3%
Asian. The First Essex incumbent, Diana DiZoglio (D-Methuen), is
running for Auditor, leaving an open race for a competitive field of
candidates from Lawrence and Methuen (check out the upcoming
Mass. Numbers State Sentate primary overview).

## The Second Suffolk District

```{r second-suffolk-redist, fig.cap="*Second Suffolk District*"}
sen_29 <- senate_geometry(29)
sen_pcts_29 <- senate_precincts(29)
sen_29_map <- senate_map(sen_29, sen_pcts_29)
tmap_leaflet(sen_29_map)
```

```{r sen-second-suffolk}
summarize_dist(sen_pcts_29) %>%
    select(total_pop, cvap_total, cvap_hispanic_pct,
           cvap_black_pct, cvap_asian_pct, cvap_white_pct) %>%
    kbl(format.args=list(scientific=FALSE, big.mark=",")) %>%
    kable_styling()
```


## The Middlesex and Suffolk District

```{r middlesex-and-suffolk-redist, layout="l-body-outset", fig.cap="*Middlesex and Suffolk District*"}
sen_26 <- senate_geometry(26)
sen_pcts_26 <- senate_precincts(26)
sen_26_map <- senate_map(sen_26, sen_pcts_26)
tmap_leaflet(sen_26_map)
```

The Middelsex and Suffolk state senate district represented by Sal
DiDomenico (D-Everett) has been redrawn adding diverse sections of
Cambridgeport and East Cambridge, while removing parts of Boston's
West End, Allston, and Brighton neighborhoods. The district ends up
with a demographic mix of 46% White, 24% Hispanic, 12% Black, and 15%
Asian. The previous incarnation of this district would have been a
majority minority district with the latest demographic information,
but the new lines reinforce its majority minority demographic.

While DiDomenico does not have primary or general election challenger,
Lydia Edwards, the incumbent of the current First Suffolk and
Middlesex District which contains some of the same Cambridgeport and
East Cambridge neighborhoods as the new Middlesex and Suffolk
district, is running unchallenged in the new Third Suffolk State
Senate District. The Third Suffolk is not majority minority, but will
continue to be represented by Edwards, the only black woman in the
Massachusetts State Senate, and one of only three people of color in
the state senate.

## The Second Plymouth and Norfolk District

```{r spn-redist, layout="l-body-outset", fig.cap="*Second Plymouth and Norfolk District*"}
sen_34 <- senate_geometry(34)
sen_pcts_34 <- senate_precincts(34)
sen_34_map <- senate_map(sen_34, sen_pcts_34)
tmap_leaflet(sen_34_map)
```

The Second Plymouth and Norfolk District was a late-comer to the
majority minority district additions during the latest redistricting
process. The first draft of the senate redistricting plan left the
Brockton-based district almost unchanged and grouped with the mostly
white surrounding suburb of Hanover, Plympton and Easton. [Pressure
from redistricting advocacy
groups](https://www.bostonglobe.com/2021/10/19/metro/under-pressure-state-senate-redraws-map-adding-new-majority-minority-district-brockton/)
led to a plan which extended the district through Avon into parts of
Randolph that moved the demographics of the district into majority
minority, while moving the white suburbs into adjacent districts.

The newly configured district has attracted an experienced Democratic
challenger in Randolph Town Councilor Katrina Huff-Larmond who could
increase the number of black women in the senate above one, Lydia
Edward (D-Boston), if she were to win the primary on September 16th.

## The remaining majority minority districts


```{r sen-2nd-suffolk-map}
sen_29 <- senate_geometry(29)
sen_pcts_29 <- senate_precincts(29)
sen_29_map <- senate_map(sen_29, sen_pcts_29)
tmap_leaflet(sen_29_map)
```

```{r sen-1st-suffolk-map}
sen_30 <- senate_geometry(30)
sen_pcts_30 <- senate_precincts(30)
sen_30_map <- senate_map(sen_30, sen_pcts_30)
tmap_leaflet(sen_30_map)
```


