---
title: "A look at MA State Senate Redistricting"
description: |
    There are now six majority-minority state senate districts in
    Massachusetts but only three give minority voters a robust chance of
    electing their candidate of choice
author: Brent Benson
date: 2022-08-22
citation: false
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
options(digits=1, knitr.kable.NA="")
library(kableExtra)
library(formattable)
library(tidyverse)
library(sf)
library(tmap)
library(readxl)
library(viridis)
library(tidycensus)
library(tigris)
library(units)
library(nominatimlite)
tmap_options(check.and.fix=TRUE)
```

```{r senate-districts}
sen_dist <- read_excel("../../_data/2020/MA_State_Senate_Race_2022.xlsx", sheet="Districts")

sen_cand <- read_csv("../../_data/2020/ma_state_senate_cands_2022_contested_addr.csv") %>%
    rename(Inc = Incumbent) %>%
    mutate(address = str_c(street_address, city_town, "MA", sep=", ")) %>%
    select(-street_address)

sen_geom <- read_sf("../../_gis/SENATE2021/SENATE2021_POLY.shp", as_tibble=TRUE) %>%
    st_make_valid() %>%
    select(Number=SENDISTNUM) %>%
    st_transform(st_crs(6491)) %>%
    left_join(sen_dist, by="Number") %>%
    relocate(District)

sen_geom_12 <- read_sf("../../_gis/senate2012/SENATE2012_POLY.shp", as_tibble=TRUE) %>%
    st_make_valid() %>%
    st_transform(st_crs(6491))

tmap_mode(mode="view")
one_m_sq <- set_units(1, m^2)
two_k_m_sq <- set_units(2000, m^2)
five_k_m_sq <- set_units(5000, m^2)
seven_five_k_m_sq <- set_units(75000, m^2)
```

```{r precinct-setup}
ma_towns <- read_sf("../../_gis/townssurvey_shp/TOWNSSURVEY_POLYM.shp",
                    as_tibble=TRUE) %>%
    st_make_valid() %>%
    mutate(city_town = str_to_title(TOWN)) %>%
    select(city_town) %>%
    st_transform(st_crs(6491))

ward_precinct_name <- function(city_town, ward, precinct) {
    str_c(city_town,
          if_else(ward == "-", "", str_c("Ward ", ward)),
          str_c("Precinct ", precinct),
          sep=" ")
}

add_voting_percents <- function(df) {
    df %>%
        mutate(cvap_white_pct = cvap_2019_white/cvap_2019_total,
               cvap_hispanic_pct = cvap_2019_hispanic/cvap_2019_total,
               cvap_black_pct = cvap_2019_black/cvap_2019_total,
               cvap_asian_pct = cvap_2019_asian/cvap_2019_total,
               cvap_pred_group = case_when(((cvap_black_pct >= 0.4) &
                                       (cvap_hispanic_pct >= 0.4)) ~ "Black_and_Hispanic",
                                  (cvap_black_pct >= 0.4) ~ "Black",
                                  (cvap_hispanic_pct >= 0.4) ~ "Hispanic",
                                  (cvap_asian_pct >= 0.4) ~ "Asian",
                                  TRUE ~ "White"),
               vap_white_pct = vap_white_2020/vap_total_2020,
               vap_hispanic_pct = vap_hispanic_2020/vap_total_2020,
               vap_black_pct = vap_black_2020/vap_total_2020,
               vap_asian_pct = vap_asian_2020/vap_total_2020,
               vap_pred_group = case_when(((vap_black_pct >= 0.4) &
                                       (vap_hispanic_pct >= 0.4)) ~ "Black_and_Hispanic",
                                  (vap_black_pct >= 0.4) ~ "Black",
                                  (vap_hispanic_pct >= 0.4) ~ "Hispanic",
                                  (vap_asian_pct >= 0.4) ~ "Asian",
                                  TRUE ~ "White"))
}

sen_dist_blks <- read_csv("../../_data/2020/ma_blocks_to_sen_dist_2021.csv",
                          col_types=cols("c", "c", "i"))

sen_vtd_demos <- read_csv("../../_data/2020/ma_sen_census_2020_daves.csv",
                          col_types=cols(vtd_fips="c"))

vtd_names <- read_csv("../../_data/2020/ma_vtds_city_town_fips.csv", col_types=cols(.default="c")) %>%
    select(vtd_fips, city_town, Ward, Pct)

vtd_full <- sen_vtd_demos %>%
    select(-c(district_id, district_num, partial_pct)) %>%
    group_by(vtd_fips) %>%
    summarize_all(sum)

vtd_demos <- vtd_names %>%
    left_join(vtd_full, by="vtd_fips") %>%
    add_voting_percents()

vtd_geom <- voting_districts(state=25, cb=TRUE) %>%
    select(vtd_fips = GEOID20) %>%
    st_transform(6491) %>%
    left_join(vtd_demos, by="vtd_fips") %>%
    mutate(name = ward_precinct_name(city_town, Ward, Pct)) %>%
    relocate(name)

block_geom <- blocks(state=25, year=2020) %>%
    select(block_fips = GEOID20) %>%
    st_transform(6491)

vtd_blks <- read_csv("../../_data/2020/ma_blocks_to_vtds.csv",
                     col_types=cols(.default="c")) %>%
    left_join(vtd_names, by="vtd_fips") %>%
    inner_join(block_geom, by="block_fips")

blks_sen_dist_vtd <- vtd_blks %>%
    left_join(sen_dist_blks, by="block_fips")

sen_vtds <- blks_sen_dist_vtd %>%
    nest(blocks = c(block_fips, geometry)) %>%
    group_by(vtd_fips) %>%
    mutate(partial_pct = n() > 1) %>%
    ungroup() %>%
    mutate(geometry = map(blocks, ~st_union(.x$geometry))) %>%
    unnest(geometry) %>%
    select(-blocks)

sen_dist_names <- sen_dist %>%
    select(district_num = Number,
           district = District)

sen_vtd_comp <- sen_dist_names %>%
    right_join(sen_vtds, by="district_num") %>%
    left_join(sen_vtd_demos, by=c("vtd_fips", "district_id", "district_num", "partial_pct")) %>%
    add_voting_percents() %>%
    mutate(name = ward_precinct_name(city_town, Ward, Pct)) %>%
    relocate(name) %>%
    st_as_sf()
```

## Overview

This deep-dive on the 2020-2021 Massachusetts State Senate
redistricting focuses on the Voting Rights Act's requirement to create
districts that disallow the dilution of minority voting blocks. By
looking at the predominant voting group in each Massachusetts precinct
we see the opportunity and limits of what can be achieved in
for Massachusetts State Senate districts.

While there are six state senate districts that are technically
majority-minority in the newly created districts, only three
provide a particular minority voting block with the power to
elect its candidate of choice, and that is the best that can be
done given Massachusetts voter demographics.

## Redistricting Concepts

The Massachusetts State Senate lines have been redrawn by the
legislature after the 2020 Census as required by the constitution.
For a guide to the redistricting principles used by the state
legislatures, a good reference is the [Executive Summary of the
Redistricting Red
Book](https://www.ncsl.org/Portals/1/Documents/Redistricting/Exec_Summary_2020.pdf),
published by the National Conference of State Legislatures.

The primary responsibility given to redistricters by the constitution
is to ensure equally-sized districts. For state legislatures, the
courts have generally approved of redistricting plans where the
largest and smallest districts are within 5 percent of the target
population. The equal-size mandate results in the shifting of
district lines as populations ebb and flow through the Commonwealth.

The next mandate for the redistricters comes from the Voting Rights
Act (VRA) and relates to race, color, and minority language groups.
While it is illegal to use race and ethnicity as factor in
redistricting in the general case, it is also illegal to dilute
concentrations of a minority voting block, resulting in that voting
block not being able to elect its candidate of choice.

While the short-hand concept of a majority-minority district is often
used when evaluating redistricting plans, it is usually an
oversimplification which does not addequately address minority voting
block dilution. In particular, minority groups do not usually vote as
a monolith. A district with 25% black voters and 30% hispanic voters
is technically majority-minority, but gives neither the black voters
nor the hispanic voters the ability to elect their preferred
candidate. (In practice, redistricting bodies can use techniques like
ecological regression to analyze and account for historical voting
patterns and tendencies to understand the voting blocks in play.)

The newly drawn 2020-2021 Massachusetts State Senate districts
increase the number of majority-minority districts from
three to six, but only three of the districts contain
a minority voting block with enough power to elect their candidate of
choice. It is also clear from looking at the data that there aren't
enough geographically concentrated minority blocks in Massachusetts
to do any better at this time.

## Concentrations of Minority Voting Blocks

Let's first look at the 2,152 voting precincts in Massachusetts by
flagging any precinct that has a citizen voting age population (CVAP)
for a minority group that is over 40% of the total citizen voting age
population.

We use citizen voting age population because that is the actual
number of people who are eligible to vote, a better determinant of
voting power than total population. We will also use voting age
population and overall population numbers when looking at specific
districts, but CVAP makes sense as the number to consider for
voting fairness.

```{r vtd-minority-precints}
vtd_demos %>%
    count(cvap_pred_group) %>%
    mutate(Percent = percent(n/sum(n), 0)) %>%
    rename(`Predominant Group`=cvap_pred_group,
           `Precinct Count`=n) %>%
    arrange(desc(`Precinct Count`)) %>%
    kbl(caption="Predominant voting group by voting precinct") %>%
    kable_styling()
```

If we look at the municipality of the 203 voting precincts that have a
critical mass of minority voters (102+94+5+3) it guides us towards
which precincts might be grouped together into a single district to
prevent dilution of that minority group's voting power.

```{r vtd-group-by-city}
cvap_towns <- vtd_demos %>%
    filter(cvap_pred_group != "White") %>%
    count(city_town, cvap_pred_group)

cvap_towns %>%
    rename(`City/Town`=city_town,
           `Predominant Group`=cvap_pred_group,
           `Precinct Count`=n) %>%
    arrange(desc(`Precinct Count`)) %>%
    kbl(caption="Number of plausibly minority precincts by city/town") %>%
    kable_styling()
```

A statewide map of the precincts with concentrated minority
representation shows the limited number and concentration of such
districts.

```{r vtd-minority-map, layout="l-page", fig.cap="*Predominant group for all MA precincts*"}
color_list <- viridis(4)
group_colors <- c(Black_and_Hispanic=color_list[4],
                  Black=color_list[1],
                  Hispanic=color_list[2],
                  Asian=color_list[3],
                  White=NA)
popup_vars <- c("cvap_pred_group", "cvap_white_pct", "cvap_black_pct",
                "cvap_hispanic_pct", "cvap_asian_pct")

cvap_town_outlines <- ma_towns %>% filter(city_town %in% unique(cvap_towns$city_town))

pred_group_map <- (tm_shape(vtd_geom) +
                       tm_borders(lwd=0.05) +
                       tm_fill(col="cvap_pred_group",
                               alpha=1,
                               title="",
                               palette=group_colors) +
                       tm_shape(cvap_town_outlines) +
                       tm_borders(lwd=0.6) +
                       tm_text(text="city_town", size=0.5, xmod=-1.25) +
                       tm_shape(ma_towns) +
                       tm_borders(lwd=0.7))

tmap_mode(mode="plot")
pred_group_map
```
A closer look at four of the clusters around Springfield, Lawrence, Boston, and
Brockton show the biggest concentration of minority voting blocks. All but
the Brockton/Randolph cluster have enough concentration to plausibly
elect a minority candidate of choice. 

```{r cvap-detail, layout="l-body-outset", fig.cap="*Cities with large minority voting blocks*"}
cvap_detail_map <- function(city_towns, fig_title, legend_show) {
    (tm_shape((vtd_geom %>% filter(city_town %in% city_towns))) +
     tm_borders(lwd=0.5, col="black") +
     tm_fill(col="cvap_pred_group",
             alpha=1,
             title="",
             palette=group_colors) +
     tm_shape((ma_towns %>% filter(city_town %in% city_towns))) +
                       tm_borders(lwd=1) +
                       tm_text(text="city_town", size=0.75, xmod=4) +
     tm_layout(title=fig_title, title.size=0.7, legend.show=legend_show))
}
springfield_cvap_map <- cvap_detail_map(c("Springfield"), "Basis of Hampden", TRUE)
essex_1_cvap_map <- cvap_detail_map(c("Haverhill", "Methuen", "Lawrence"),
                                     "Basis of 1st Essex",
                                     FALSE)
boston_cvap_map <- cvap_detail_map(c("Boston"),
                                   "Basis of 1st/2nd Suffolk",
                                   FALSE)
brockt_rand_cvap_map <- cvap_detail_map(c("Brockton", "Randolph"),
                                        "Basis of 2nd Norfolk and Plymouth",
                                        FALSE)
tmap_arrange(springfield_cvap_map,
             essex_1_cvap_map,
             boston_cvap_map,
             brockt_rand_cvap_map,
             ncol=2)
```

## Districts with Minority Power

The new Hampden, First Essex, and Second Suffolk Districts contain a
large enough percentage of a particular minority group to allow for
election of a candidate of choice by that minority group.

```{r district-map-setup}
color_list_1 <- viridis(5)
group_colors_1 <- c(Black_and_Hispanic=color_list_1[4],
                    Black=color_list_1[1],
                    Hispanic=color_list_1[2],
                    Asian=color_list_1[3],
                    White=color_list_1[5])
popup_vars_1 <- c("cvap_pred_group", "cvap_white_pct", "cvap_black_pct",
                  "cvap_hispanic_pct", "cvap_asian_pct")

senate_geometry <- function(dist_number) {
    sen_geom %>% filter(Number == dist_number)
}

senate_precincts <- function(n) {
    sen_vtd_comp %>%
        filter(district_num == n)
}

senate_map <- function(sen_dist, sen_pcts) {
    (tm_shape(sen_pcts) +
     tm_borders(lwd=0.75) +
     tm_fill(col="cvap_pred_group",
             alpha=0.5,
             title="",
             palette=group_colors_1,
             popup.vars=popup_vars_1) +
    tm_shape(sen_dist) + tm_borders(lwd=3) +
    tm_basemap("OpenStreetMap"))
}
```

## The Hampden District

```{r sen-hampden-map}
sen_3 <- senate_geometry(3)
sen_pcts_3 <- senate_precincts(3)
sen_3_map <- senate_map(sen_3, sen_pcts_3)
tmap_leaflet(sen_3_map)
```

```{r summarize-districts}
summarize_dist <- function(sen_pcts) {
    sen_pcts %>%
        st_drop_geometry() %>%
        group_by(district) %>%
        summarize(total_total = sum(total_pop_2020),
                  total_hispanic = sum(total_hispanic_2020),
                  total_black = sum(total_black_2020),
                  total_asian = sum(total_asian_2020),
                  total_white = sum(total_white_2020),
                  total_hispanic_pct = total_hispanic/total_total,
                  total_black_pct = total_black/total_total,
                  total_asian_pct = total_asian/total_total,
                  total_white_pct = total_white/total_total,
                  total_minority_pct = 1 - total_white_pct,
                  vap_total = sum(vap_total_2020),
                  vap_hispanic = sum(vap_hispanic_2020),
                  vap_black = sum(vap_black_2020),
                  vap_asian = sum(vap_asian_2020),
                  vap_white = sum(vap_white_2020),
                  vap_hispanic_pct = vap_hispanic/vap_total,
                  vap_black_pct = vap_black/vap_total,
                  vap_asian_pct = vap_asian/vap_total,
                  vap_white_pct = vap_white/vap_total,
                  vap_minority_pct = 1 - vap_white_pct,
                  cvap_total = sum(cvap_2019_total),
                  cvap_hispanic = sum(cvap_2019_hispanic),
                  cvap_black = sum(cvap_2019_black),
                  cvap_asian = sum(cvap_2019_asian),
                  cvap_white = sum(cvap_2019_white),
                  cvap_hispanic_pct = cvap_hispanic/cvap_total,
                  cvap_black_pct = cvap_black/cvap_total,
                  cvap_asian_pct = cvap_asian/cvap_total,
                  cvap_white_pct = cvap_white/cvap_total,
                  cvap_minority_pct = 1 - cvap_white_pct) %>%
         select(-ends_with(c("_hispanic", "_black", "_asian", "_white"))) %>%
         pivot_longer(cols=!district,
                      names_sep="_",
                      names_to=c("stat", "group")) %>%
         pivot_wider(id_cols=c(district, stat),
                     names_from=group,
                     values_from=value)
}

summary_table <- function(sen_summary, caption) {
    sen_summary %>%
        select(-district) %>%
        mutate(stat = str_replace_all(stat, c("total"="Total Population",
                                              "cvap"="Citizen Voting Age Population",
                                              "vap"="Voting Age Population")),
               hispanic = percent(hispanic, 0),
               black = percent(black, 0),
               asian = percent(asian, 0),
               white = percent(white, 0),
               minority = percent(minority, 0)) %>%
        kbl(caption=caption,
            format.args=list(scientific=FALSE, big.mark=",")) %>%
        kable_styling()
}
```

```{r sen-hampden}
summary_table(summarize_dist(sen_pcts_3), "Hampden District Demographics")
```

The Hampden State Senate District contains 90 percent of Springfield's
precincts and 60 percent of Chicopee's precincts. The district
has about as many hispanic voters as non-hispanic white voters,
giving that voting block a chance at electing their preferred
candidate.

The pre-2021 version of the Hampden district had a similar
characteristics. In 2020 Springfield City Councilor Alex Gomez
defeated incumbent State Senator James Welch in the Democratic
primary.

A simple precinct-level ecological regression model that uses Gomez's
margin over Welch as the response variable and the CVAP hispanic
percent and whether the voter is in Springfield as interacting
explanatory variables ends up explaining 90 percent of the variation
in the margin of the 2020 Hampden District Democratic primary
vote. Hispanic voters were much more likely to vote for Gomez,
especially those who lived in Springfield an knew of Gomez from the
City Council.

## The First Essex District

```{r first-essex-redist, fig.cap="*First Essex District*"}
sen_19 <- senate_geometry(19)
sen_pcts_19 <- senate_precincts(19)
sen_19_map <- senate_map(sen_19, sen_pcts_19)
tmap_leaflet(sen_19_map)
```

```{r sen-first-essex}
summary_table(summarize_dist(sen_pcts_19), "First Essex District Demographics")
```

The First Essex District is the first successful attempt to create a
majority hispanic Lawrence-based State Senate district by carving out
largely hispanic precincts in Haverhill to add to Lawrence's large
hispanic population.

The district contains all of Lawrence, all of Methuen, and
approximately 11 precincts of Haverhill (give or take a few census
blocks). The new lines result in a voting age population that is
36% White, 56% Hispanic, 10% Black, and 3% Asian. There is an almost
10 point difference when looking at hispanic citizen voting age
population, showing the large assimilation of immigrants, and somewhat
limiting voting power.

The First Essex incumbent, Diana DiZoglio (D-Methuen), is
running for Auditor, leaving an open race for a competitive field of
candidates from Lawrence (Pavel Payano and Doris Rodriguez) and
Methuen (Eunice Zeigler).

## The Second Suffolk District

```{r second-suffolk-redist, fig.cap="*Second Suffolk District*"}
sen_29 <- senate_geometry(29)
sen_pcts_29 <- senate_precincts(29)
sen_29_map <- senate_map(sen_29, sen_pcts_29)
tmap_leaflet(sen_29_map)
```

```{r sen-second-suffolk}
summary_table(summarize_dist(sen_pcts_29), "Second Suffolk District Demographics")
```

The State Senate redistricting challenge in Boston is to create a
district that gives candidate-of-choice power to black voters in
Boston, whose numbers are large, but not large enough to give voting
majorities in two districts. The 2020-2021 Senate redistricters
shifted majority-black precincts from the First Suffolk District
to the Second Suffolk, give a black citizen age voting population
of 47%.

There is a very competitive 2022 Democratic primary for the Second
Suffolk District featuring Miniard Culpepper, Nika Elugardo, Liz
Miranda, and Dianne Wilkerson.

## The remaining majority minority districts

While there are three more majority-minority State Senate
districts, none of contain a large enough block of minority
voters to provide candidate of choice power to any one
minority group.

## The Second Plymouth and Norfolk District

```{r spn-redist, fig.cap="*Second Plymouth and Norfolk District*"}
sen_34 <- senate_geometry(34)
sen_pcts_34 <- senate_precincts(34)
sen_34_map <- senate_map(sen_34, sen_pcts_34)
tmap_leaflet(sen_34_map)
```

```{r sen-spn-summary}
summary_table(summarize_dist(sen_pcts_34), "Second Plymouth and Norfolk District Demographics")
```

The Second Plymouth and Norfolk District was a late-comer to the
majority minority district additions during the latest redistricting
process. The first draft of the senate redistricting plan left the
Brockton-based district almost unchanged and grouped with the mostly
white surrounding suburb of Hanover, Plympton and Easton. [Pressure
from redistricting advocacy
groups](https://www.bostonglobe.com/2021/10/19/metro/under-pressure-state-senate-redraws-map-adding-new-majority-minority-district-brockton/)
led to a plan which extended the district through Avon into parts of
Randolph that moved the demographics of the district into majority
minority, while moving the white suburbs into adjacent districts.

The newly configured district has attracted an experienced Democratic
challenger in Randolph Town Councilor Katrina Huff-Larmond who could
increase the number of black women in the senate above one, Lydia
Edward (D-Boston), if she were to win the primary on September 16th.


## The First Suffolk District

```{r sen-1st-suffolk-map}
sen_30 <- senate_geometry(30)
sen_pcts_30 <- senate_precincts(30) %>%
    erase_water(area_threshold=0.6) %>%
    st_make_valid()
    
sen_30_map <- senate_map(sen_30, (sen_pcts_30 %>% filter(Ward != "1")))
tmap_leaflet(sen_30_map)
```
```{r sen-1st-suffolk-summary}
summary_table(summarize_dist(sen_pcts_30), "First Suffolk District Demographics")
```

The First Suffolk District is currently represented by Nick Collins of
South Boston. While the seat was held by Linda Dorcena Forry of
Dorchester from 2013 through 2018, the demographics of the district
are not particularly favorable for electing a black state senator. The
2020-2021 senate redistricting committe chose to shore up the Second
Suffolk District to give its black voters the ability to elect their
candidate of choice, leaving the First Suffolk with a black citizen
voting age population of around 23%.

## The Middlesex and Suffolk District

```{r middlesex-and-suffolk-redist, fig.cap="*Middlesex and Suffolk District*"}
sen_26 <- senate_geometry(26)
sen_pcts_26 <- senate_precincts(26)
sen_26_map <- senate_map(sen_26, sen_pcts_26)
tmap_leaflet(sen_26_map)
```

```{r sen-m-and-s-summary}
summary_table(summarize_dist(sen_pcts_26), "Middlesex and Suffolk District Demographics")
```

The Middelsex and Suffolk state senate district represented by Sal
DiDomenico (D-Everett) has been redrawn adding  sections of
Cambridgeport and East Cambridge, while removing parts of Boston's
West End, Allston, and Brighton neighborhoods. While the overall
population is 58% minority, the citizen voting age population is 39%
minority and the largest minority voting block are hispanics at 17%.

## Conclusion

The Massachusetts State Senate redistricting committee adhered
to the letter and the spirit of the Voting Rights Act by
doing its best to group together minority voting blocks and
create districts with true minority voting power.

