---
title: "Massachusetts shifts towards Trump"
description: |
    Demographic factors associated with the shift include rates of college attendance, socio-economic status, race, and ethnicity
author: Brent Benson
date: 2025-01-12
citation: false
preview: https://www.massnumbers.us/images/pres_vote_shifts.png
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(formattable)
library(plotly)
library(rstanarm)
library(glue)
library(gt)
library(gtExtras)
library(sf)
library(tmap)
library(tmaptools)
library(metathis)

tmap_mode("view")
```

```{r post-meta}
meta() %>%
    meta_description("Demographic factors associated with MA presidential votes shifts include rates of college attendance, socio-economic status, race, and ethnicity") |>
    meta_viewport() |>
    meta_social(
        title = "Massachusetts shifts towards Trump",
        url = "https://www.massnumbers.us/posts/2025-01-12-vote-shifts/",
        image = "https://www.massnumbers.us/images/pres_vote_shifts.png",
        image_alt = "Presidential vote shift by State Representative district",
        og_type = "article",
        og_author = "Brent Benson",
        twitter_card_type = "summary_large_image",
        twitter_creator = "@bwbensonjr"
    )
```

The Commonwealth of Massachusetts remained reliably blue in the 2024
presidential election with Kamala Harris taking 61% of the
vote. However, Massachusetts has seen increase in Donald Trump's vote
share going from approximately 32% of the vote in 2016 and 2020
to 36% in 2024.

The statewide shift has not been uniform with some areas moving
towards Trump while other areas became even more Democratic. To better
understand these shift we use statistical analysis of demographic
factors to better understand which factors are associated with the
shifts.

We will use Massachusetts State Representative districts as the unit
to measure the vote shift and demographic factors. The advantage of
using State Rep. districts is that they are approximately the same
size in terms of population (currently about 44,000 people), and they
are created with the guidance of the Voting Rights Act (VRA) with the
goal of not breaking up voting blocks along the lines of race and
minority language groups, resulting in variations that provide good
contrasts for our statistical analysis.  (There are more details on
redistricting and district formation in the [Mass. Numbers deep-dive
on State Senate
redistricting](https://www.massnumbers.us/posts/2022-07-18-senate-majority-minority/).)

## Districts with the largest shifts

It is important to emphasize that we are examining voting behavior
shifts rather than district partisanship. *All* of the districts with
the largest shifts towards Trump are majority Democratic
districts. The district with the largest shift towards Trump, the
Tenth Hampden District in Springfield, remains remarkably Democratic
with a Partisan Voter Index (PVI, an average of the last two
presidential elections compared to the U.S. as a whole) of D+24, even
after an eight point Republican shift.

The districts with the largest shifts towards Trump are:

```{r layout="l-body-outset"}

largest_trump_shifts <- tribble(
    ~District, ~`City/Town`, ~PVI, ~Representative, ~Shift,
    "Tenth Hampden", "Springfield", "D+24", "Carlos Gonzalez (D-Springfield)", "R+8",
    "Fourth Essex", "Lawrence, Methuen", "D+12", "Estela A. Reyes (D-Lawrence)", "R+7",
    "Sixteenth Essex", "Lawrence, Methuen", "D+11", " Francisco E. Paulino (D-Methuen)", "R+7",
    "Eleventh Bristol", "New Bedford", "D+10", " Christopher Hendricks (D-New Bedford)", "R+7",
    "Seventh Bristol",	"Fall River", "D+1", " Alan Silvia (D-Fall River)", "R+7",
    "Seventeenth Essex", "Andover, Lawrence", "D+17", " Frank A. Moran (D-Lawrence)", "R+6")

largest_trump_shifts |>
    gt()
```

The districts with the largest shifts away from Trump and towards Democratic candidates are:

```{r layout="l-body-outset"}
largest_dem_shifts <- tribble(
    ~District, ~`City/Town`, ~PVI, ~Representative, ~Shift,
    "Fourth Barnstable", "Chatham, Eastham, Harwich, Orleans, Provincetown, Truro, Wellfleet", "D+19", "Hadley Luddy (D-Orleans)", "D+3",
    "Second Middlesex",	"Chelmsford, Littleton, Westford", "D+17", "James Arciero (D-Westford)", "D+3",
    "Fourteenth Essex", "Amesbury, Boxford, Groveland, North Andover, West Newbury", "D+10", "Adrianne Pusateri Ramos (D-North Andover)", "D+3",
    "First Middlesex", "Ashby, Dunstable, Groton, Lunenburg, Pepperell, Townsend", "D+5", "Margaret R. Scarsdale (D-Pepperell)", "D+3",
    "Third Barnstable", "Bourne, Falmouth, Mashpee", "D+9", "David T. Vieira (R-Falmouth)", "D+3",
    "Thirty-Second Middlesex", "Malden, Melrose, Wakefield", "D+20", "Kate Lipper-Garabedian (D-Melrose)", "D+3",)

largest_dem_shifts |>
    gt()
```

```{r}
square_meters_per_square_mile = 2.59e+6

state_reps <- read_csv("https://bwbensonjr.github.io/ma-election-db/data/ma_general_election_summaries_1990_2024.csv.gz") |>
    filter(election_date == "2024-11-05", office == "State Representative") |>
    select(district, Representative=display_winner)

elec_summaries <- read_csv("https://bwbensonjr.github.io/ma-election-db/data/ma_general_election_summaries_1990_2024.csv.gz")

pres_results <- read_csv("https://bwbensonjr.github.io/mapoli/pvi/ma_state_rep_pres_pvi_2024.csv") |>
    rename(district = State_Rep)

demos <- read_csv("https://bwbensonjr.github.io/mapoli/demographics/data/ma_state_rep_demographics.csv") |>
    mutate(area_square_miles = area_m2/square_meters_per_square_mile,
           household_density = total_households/area_square_miles)

pvi_string <- function(pvi_n) {
    case_when(
        pvi_n <= -0.5 ~ str_c("R+", round(abs(pvi_n))),
        pvi_n >= 0.5 ~ str_c("D+", round(pvi_n)),
        TRUE ~ "EVEN",
    )
}

pres_demos <- pres_results |>
    left_join(state_reps, by="district") |>
    left_join(demos, by="district") |>
    mutate(`PVI Shift` = pvi_string(pvi_shift)) |>
    rename(`PVI 2024` = PVI,
           `PVI 2020` = PVI_20,
           `Some College`=ed_some_college_pct,
           White=race_white_pct,
           Hispanic=race_hispanic_pct,
           Portuguese=ancestry_portuguese_pct,
           Poverty=below_poverty_pct)

dist_geom <- read_sf("https://bwbensonjr.github.io/mapoli/gis/geojson/house2021.geojson") |>
    left_join(pres_demos, by="district")
```

```{r layout="l-body-outset", fig.height=5, fig.cap="An overview of vote shifts by State Rep. district with red indicating shifts towards Trump and blue towards Democratic presidential candidates."}
percent_zero <- function(n) {
    percent(n, 0)
}

digits_zero <- function(n) {
    digits(n, digits=0)
}

popup_vars <- c(
    "Representative",
    "PVI Shift",
    "PVI 2024",
    "PVI 2020",
    "Some College",
    "White",
    "Hispanic",
    "Portuguese",
    "Poverty"
)

popup_format = list(
    `Some College`=list(fun=percent_zero),
    White=list(fun=percent_zero),
    Hispanic=list(fun=percent_zero),
    Portuguese=list(fun=percent_zero),
    Poverty=list(fun=percent_zero)
)

tm_shape(dist_geom) +
    tm_polygons(col="pvi_shift",
                palette=c("#FF4949", "white", "#0F4392"),
                midpoint=0,
                popup.vars=popup_vars,
                popup.format=popup_format,
                alpha=0.7) +
    tm_basemap("OpenStreetMap") +
    tm_layout("Presidential vote shift by State Representative district") +
    tm_credits("An overview of vote shifts by State Rep. district with red indicating shifts towards Trump and blue towards Democratic presidential candidates.") +
    tm_view(set.view=8)
```

## Associated demographic factors

The primary demographic factors associated with shifts toward and away from Trump are:

* **College attendance** - A higher percentage of people who have attended some college is associated with a shift away from Trump.
* **Race and ethnicity** - A larger Hispanic shares of the population and a larger percentage of the population of Portuguese ancestry is correlated with a bigger shift towards Trump while a larger white share of the population is correlated with larger shifts away from Trump.
* **Socio-economic level** - districts with a higher percentage of people below the poverty line moved more towards Trump.

```{r layout="l-body-outset", fig.height=6, fig.cap="Factors associated with vote shift. Lines sloping downward indicate shifts are towards Trump and upward sloping lines are shifts toward Democrats."}


variables <- c(
    "Some College",
    "White",
    "Hispanic",
    "Portuguese",
    "Poverty"
)

ggplotly((pres_demos |>
    select(district, pvi_shift, all_of(variables)) |>
    pivot_longer(cols=any_of(variables),
                 names_to="variable",
                 values_to="value") |>
    rename(`PVI Shift`=pvi_shift) |>
    ggplot(aes(x=value, y=`PVI Shift`, group=1, district=district)) +
    geom_point() +
    geom_smooth(formula="y ~ x", span=2, se=FALSE) +
    scale_x_continuous(labels = scales::percent_format()) +
    facet_wrap(~variable) +
    theme(axis.title.x=element_blank())))
```


```{r layout="l-body-outset"}
partisan_palette <- function(x) {
     f_neg <- scales::col_numeric(
         palette = c("#FF4949", "white"),
         domain = c(-44, 0)
     )
     f_pos <- scales::col_numeric(
         palette = c("white", "#0F4392"),
         domain = c(0, 44)
     )
     ifelse(x < 0, f_neg(x), f_pos(x))
}

pres_demos |>
    arrange(pvi_shift) |>
    select(district,
           Representative,
           pvi_shift,
           PVI_N,
           PVI_N_20,
           `PVI Shift`,
           `PVI 2024`,
           `PVI 2020`,
           `Some College`,
           Poverty,
           Hispanic,
           Portuguese,
           White) |>
    mutate(`Some College` = `Some College`*100,
           White = White*100,
           Hispanic = Hispanic*100,
           Portuguese = Portuguese*100,
           Poverty = Poverty*100) |>
    gt() |>
        tab_header(title="Vote shifts and demographics by State Rep. district") |>
        cols_hide(columns=c(pvi_shift, PVI_N, PVI_N_20)) |>
        data_color(columns=pvi_shift,
                   colors=partisan_palette,
                   target_columns =`PVI Shift`) |>
        data_color(columns=PVI_N,
                   colors=partisan_palette,
                   target_columns =`PVI 2024`) |>
        data_color(columns=PVI_N_20,
                   colors=partisan_palette,
                   target_columns =`PVI 2020`) |>
        gt_plt_bar_pct(column=`Some College`, fill="#B07AA1FF", scaled=TRUE, labels=TRUE, decimal=0) |>
        gt_plt_bar_pct(column=Poverty, fill="#F28E2BFF", scaled=TRUE, labels=TRUE, decimal=0) |>
        gt_plt_bar_pct(column=Hispanic, fill="#FF9DA7FF", scaled=TRUE, labels=TRUE, decimal=0) |>
        gt_plt_bar_pct(column=Portuguese, fill="#BAB0ACFF", scaled=TRUE, labels=TRUE, decimal=0) |>
        gt_plt_bar_pct(column=White, fill="#76B7B2FF", scaled=TRUE, labels=TRUE, decimal=0)
```

