---
title: "Driscoll won Lieutenant Governor primary by having largest geographic reach"
description: |
  Lesser and Gouveia were not able to expand local support bases
author: Brent Benson
date: 2022-09-12
citation: false
preview: https://www.massnumbers.us/images/lt_gov_regionality.png
output:
  distill::distill_article:
    self_contained: false
draft: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)

library(metathis)
library(kableExtra)
library(tidyverse)
library(tigris)
library(sf)
library(units)
library(tmap)
library(formattable)

tmap_mode(mode="view")
```

```{r post-meta}
meta() %>%
    meta_description("Expansion beyond regional base was key to Driscoll LG victory") %>%
    meta_viewport() %>%
    meta_social(
        title = "Driscoll won Lieutenant Governor primary by having largest geographic reach",
        url = "https://www.massnumbers.us/posts/2022-09-10-lg-regional-candidates/",
        image = "https://www.massnumbers.us/images/lt_gov_regionality.png",
        image_alt = "The geographic range of each Lieutenant Governor Democratic primary candidate",
        og_type = "article",
        og_author = "Brent Benson",
        twitter_card_type = "summary_large_image",
        twitter_creator = "@bwbensonjr"
    )
```

```{r lg-data}
city_town_votes <- read_csv("../../_data/2020/ma_lt_gov_dem_primary_2022.csv") %>%
    mutate(total_votes = Driscoll + Gouveia + Lesser,
           max_votes = pmax(Driscoll, Gouveia, Lesser),
           Driscoll_pct = Driscoll/total_votes,
           Gouveia_pct = Gouveia/total_votes,
           Lesser_pct = Lesser/total_votes,
           winner = case_when(
               0 == max_votes ~ "Unreported",
               Driscoll == max_votes ~ "Driscoll",
               Gouveia == max_votes ~ "Gouveia",
               Lesser == max_votes ~ "Lesser"))

total_votes <- city_town_votes %>%
    select(city_town, Driscoll, Gouveia, Lesser) %>%
    pivot_longer(cols=-city_town, names_to="candidate", values_to="votes") %>%
    group_by(candidate) %>%
    summarize(votes = sum(votes)) %>%
    mutate(votes_pct = votes/sum(votes))

## Combine City/Town geometry from MassGIS with votes
city_town_geom <- read_sf("../../_gis/townssurvey_shp/TOWNSSURVEY_POLYM_GENCOAST.shp") %>%
    st_make_valid() %>%
    mutate(city_town = str_replace(str_to_title(TOWN),
                                   "Manchester",
                                   "Manchester-by-the-Sea"),
           town_center = st_centroid(geometry)) %>%
    select(city_town, town_center)

## Use the center of each City/Town for distance calculation
city_town_centers <- city_town_geom %>%
    select(city_town, town_center) %>%
    arrange(city_town) %>%
    st_drop_geometry() %>%
    st_as_sf(sf_column_name="town_center")

## Calculate distance between each pair of municipalities and
## put in a data-frame with columns city_town, city_town_2, distance

dist_matrix <- set_units(st_distance(city_town_centers, city_town_centers), mi)
colnames(dist_matrix) <- city_town_centers$city_town
dist_df <- dist_matrix %>%
    as_tibble() %>%
    mutate(city_town = city_town_centers$city_town) %>%
    relocate(city_town) %>%
    pivot_longer(cols=-city_town,
                 names_to="city_town_2",
                 values_to="distance")

lt_candidates <- city_town_centers %>%
    right_join(tibble(
        name=c("Kim Driscoll", "Tami Gouveia", "Eric Lesser"),
        position=c("Mayor", "State Representative", "State Senator"),
        candidate=c("Driscoll", "Gouveia", "Lesser"),
        city_town=c("Salem", "Acton", "Longmeadow")),
        by="city_town") %>%
    left_join(total_votes, by="candidate") %>%
    relocate(name) %>%
    arrange(desc(votes))

## Non-contiguous towns to leave out of distance calculation
outlier_towns <- c(
    "New Ashford",
    "Windsor",
    "Weston",
    "Gosnold",
    "Aquinnah"
)

## Calculate the distance between candidate home town
## and farthest won town, disregarding non-contiguous
## outliers
lt_cand_geom <- lt_candidates %>%
    left_join(dist_df, by="city_town") %>%
    left_join((city_town_votes %>%
               filter(!(city_town %in% outlier_towns)) %>%
               select(city_town_2=city_town, winner)),
              by="city_town_2") %>%
    filter(city_town != city_town_2, candidate == winner) %>%
    group_by(candidate) %>%
    filter(distance == max(distance)) %>%
    ungroup() %>%
    rename(farthest_win = city_town_2,
           win_distance = distance) %>%
    select(-winner) %>%
    left_join((city_town_centers %>%
               as_tibble() %>%
               rename(farthest_win = city_town,
                      farthest_center = town_center)),
              by="farthest_win") 

## Create a line between candidate home town
## and farthest won municipality.
lt_lines <- lt_cand_geom %>%
    as_tibble() %>%
    rowwise() %>%
    mutate(distance_line = st_cast(st_combine(c(town_center, farthest_center)), "LINESTRING")) %>%
    ungroup() %>%
    st_as_sf(sf_column_name="distance_line")
```

```{r lg-map-setup}
county_geom <- counties(state="25", cb=TRUE, class="sf")

popup_vars = c("winner",
               "Driscoll", "Driscoll_pct",
               "Lesser", "Lesser_pct",
               "Gouveia", "Gouveia_pct")

percent_zero <- function(n) {
    percent(n, 0)
}

popup_format <- list(
    Driscoll_pct=list(fun=percent_zero),
    Lesser_pct=list(fun=percent_zero),
    Gouveia_pct=list(fun=percent_zero)
)

m_blue <- "#0bb4ff"
m_yellow <- "#e6d800"
m_green <- "#00bfa0"
m_light_gray <- "#cbd6e4"

cat_palette <- c(
    Driscoll = m_blue,
    Gouveia = m_yellow,
    Lesser = m_green,
    Unreported = m_light_gray
)
```

```{r lg-winner-map, layout="l-page", fig.height=5,fig.cap="Winner of MA Lt. Governor Democratic Primary"}
lg_winner_map <- (
    tm_shape(city_town) +
    tm_fill(col="winner",
            style="cat",
            palette=cat_palette,
            alpha=0.7,
            popup.vars=popup_vars,
            popup.format=popup_format) +
    tm_borders(lwd=1) +
    tm_shape(county_geom) +
    tm_borders(lwd=2, col="black") +
    tm_shape(lt_candidates) +
    tm_symbols(col="black", size=0.3) +
    tm_basemap("OpenStreetMap") +
    tm_view(set.view=8)
)
tmap_leaflet(lg_winner_map)
```

```{r lg-results}
lt_candidates %>%
    as_tibble() %>%
    mutate(votes_pct = percent(votes_pct, 0),
           winner = if_else(votes == max(votes), "X", "")) %>%
    select(name, city_town, position, votes, votes_pct, winner) %>%
    kbl(caption="Results of 2022 Democratic Primary for Lieutenant Governor (unofficial, missing 2 towns)",
        format.args=list(big.mark=",")) %>%
    kable_styling()
```

The Massachusetts State Primary election on September 7 featured
several hotly contested races for statewide consitutional offices,
including a three-way race for the Democratic nomination for
Lieutenant Governor. The candidates on the ballot were Kim Driscoll,
mayor of Salem, Tami Gouveia of Acton, State Representative for the
14th Middlesex District, and Eric Lessor of Longmeadow, State Senator
for the 1st Hampden and Hampshire District. Another Western
Massachusetts State Senator, Adam Hinds (D-Pittsfield), was in
the early running, but failed to get the requisite 15% support
at the Democratic convention in Worcester in early June.

Mayor Driscoll [earned the endorsement of the state
party](https://www.bostonglobe.com/2022/06/04/metro/mass-democratic-convention-activists-gather-healeys-star-set-rise/)
at the convention, and held a [name recognition and preference
lead](https://www.massincpolling.com/the-topline/new-poll-priorities-for-progress-democratic-primary-poll)
over Lesser and Gouveia throughout much of the primary. Driscoll was
also able to secure some Western Massachusetts support after Adam
Hinds left the race, [securing
nominations](https://www.politico.com/newsletters/massachusetts-playbook/2022/06/08/raising-massachusetts-00037998)
from the mayors of Pittsfield and Easthampton.

Senator Lesser came into the race with high hopes and a great deal of
national Democratic support, having served in the Obama administration
as Special Assistant to Senior Advisor David Axelrod. While Lesser
called on his famous White House colleagues for help with fundraising
and volunteer recruiting, he wasn't able to make major inroads in the
non-western parts of the Commonwealth.

Representative Gouveia was a favorite of [progressive
activists](https://www.jpprogressives.com/2022/08/our-final-list-of-2022-primary-endorsements/)
but had even more trouble with name recognition and geographic
reach beyond the Route 2 corridor of Concord, Acton, Littleton,
and Harvard.

## Measuring Geographic Reach

While the winning candidate map gives us an intuitive sense of each
candidate's geographic reach, I found it useful to quantify the reach
by finding the distances between each candidate's home community and
all of the cities and towns where that candidate won the plurality of
the vote. Taking the community with the maximum distance gives an idea
of each candidate's geographic reach. I am ignoring several outlier
towns for Driscoll and Lesser that aren't contiguously attached to the
other supported areas.

```{r lg-geographic-reach}
lt_lines %>%
    as_tibble() %>%
    mutate(votes_pct = percent(votes_pct, 0),
           win_distance = round(win_distance),
           winner = if_else(votes == max(votes), "X", "")) %>%
    select(name, city_town, farthest_win, win_distance) %>%
    kbl(caption="Candidate geographic reach",
        digits=3) %>%
    kable_styling()
```

While Driscoll's maximum reach community was Nantucket, you can also
see her range by her wins in Warwick, MA (74 miles) in Franklin County
and Southbridge, MA (66 miles) near the western edge of Worcester
County with only Sturbridge as a buffer between Lesser's home county
of Hampden.

```{r lg-winner-distance-map, layout="l-page", fig.height=5,fig.cap="Winner of MA Lt. Governor Democratic Primary w/distances"}
lg_winner_dist_map <- (
    tm_shape(city_town) +
    tm_fill(col="winner",
            style="cat",
            palette=cat_palette,
            alpha=0.7,
            popup.vars=popup_vars,
            popup.format=popup_format) +
    tm_borders(lwd=1) +
    tm_shape(county_geom) +
    tm_borders(lwd=2, col="black") +
    tm_shape(lt_candidates) +
    tm_symbols(col="black", size=0.3) +
    tm_shape(lt_lines) + tm_lines(lwd=2) +
    tm_basemap("OpenStreetMap") +
    tm_view(set.view=8)
)
tmap_leaflet(lg_winner_dist_map)
```

## Kim Driscoll Vote Share

The Driscoll vote share map shows incredible strength on the North Shore
in Essex County, as you would expect. While the percentages are not
quite as large as Lesser's in his home district towns, the larger
community sizes lead to much higher total margins. The only weak
areas for Driscoll in Middlesex County are the Route 2 corridor
towns of Gouveia's home district@mdash;while Driscoll maintains
a second place standing in those towns.

Driscoll took 50% of Boston with Lesser and Gouveia splitting
the remaining 50% and remained strong in Worcester County,
the South Shore, and on the Cape and the Islands.

```{r lg-driscoll-map, layout="l-page", fig.height=5,fig.cap="Kim Driscoll vote share MA Lt. Governor Democractic Primary"}
lg_driscoll_map <- (
    tm_shape(city_town) +
    tm_fill(col="Driscoll_pct",
            alpha=0.7,
            popup.vars=popup_vars,
            popup.format=popup_format) +
    tm_borders(lwd=1) +
    tm_shape(county_geom) +
    tm_borders(lwd=2, col="black") +
    tm_shape(lt_candidates) +
    tm_symbols(col="black", size=0.3) +
    tm_basemap("OpenStreetMap") +
    tm_view(set.view=8)
)
tmap_leaflet(lg_driscoll_map)
```

## Eric Lesser Vote Share

Senator Lesser's voter stregth was limited almost exlusively to the
Western Massachusetts counties of Hampden, Hampshire, and Franklin,
and Berkshire County to a smaller extent. Lesser's frontier extends
into a few towns at the edge of Worcester County, but not beyond.

```{r lg-lesser-map, layout="l-page", fig.height=5,fig.cap="Eric Lesser vote share MA Lt. Governor Democractic Primary"}
lg_lesser_map <- (
    tm_shape(city_town) +
    tm_fill(col="Lesser_pct",
            alpha=0.7,
            popup.vars=popup_vars,
            popup.format=popup_format) +
    tm_borders(lwd=1) +
    tm_shape(county_geom) +
    tm_borders(lwd=2, col="black") +
    tm_shape(lt_candidates) +
    tm_symbols(col="black", size=0.3) +
    tm_basemap("OpenStreetMap") +
    tm_view(set.view=8)
)
tmap_leaflet(lg_lesser_map)
```

## Tami Gouveia Vote Share

Two-term State Representative Gouveia was able to get the endorsements of
progressive organizations like the Massachusetts Teachers Assocation and
Progressive Mass, but was not able to parlay that into wider name
recognition around the Commonwealth.

```{r lg-gouveia-map, layout="l-page", fig.height=5,fig.cap="Tami Gouveia vote share MA Lt. Governor Democractic Primary"}
lg_gouveia_map <- (
    tm_shape(city_town) +
    tm_fill(col="Gouveia_pct",
            alpha=0.7,
            popup.vars=popup_vars,
            popup.format=popup_format) +
    tm_borders(lwd=1) +
    tm_shape(county_geom) +
    tm_borders(lwd=2, col="black") +
    tm_shape(lt_candidates) +
    tm_symbols(col="black", size=0.3) +
    tm_basemap("OpenStreetMap") +
    tm_view(set.view=8)
)
tmap_leaflet(lg_gouveia_map)
```

