---
title: "A closer look at the Norfolk, Worcester and Middlesex State Senate fundamentals"
description: |
  A multi-year shift in the region's partisanship explains Democratic strength
author: Brent Benson
date: 2022-11-03
citation: false
preview: https://www.massnumbers.us/images/state_senate_shift_12_20.png
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)

library(metathis)
library(tidyverse)
library(readxl)
library(rstanarm)
library(humaniformat)
library(gt)
library(gtExtras)
library(sf)
library(tmap)
library(formattable)
```

```{r post-meta}
meta() %>%
    meta_description("A multi-year shift in the region's partisanship explains Democratic strength") %>%
    meta_viewport() %>%
    meta_social(
        title = "A closer look at the Norfolk, Worcester and Middlesex State Senate fundamentals",
        url = "https://www.massnumbers.us/posts/2022-11-03-norfolk-worcester-middlesex/",
        image = "https://www.massnumbers.us/images/state_senate_shift_12_20.png",
        image_alt = "A map of state senate district partisan shifts from 2012 to 2020",
        og_type = "article",
        og_author = "Brent Benson",
        twitter_card_type = "summary_large_image",
        twitter_creator = "@bwbensonjr"
    )
```

```{r map-setup}
sen_geom <- read_sf("../../_gis/senate2012/SENATE2012_POLY.shp") %>%
    st_make_valid() %>%
    st_simplify(dTolerance=100, preserveTopology=TRUE) %>%
    select(dist_number = SENDISTNUM,
           district = SEN_DIST)

NUMS_TO_WORDS = c(
    `&`='and',
    `1st`='First',
    `2nd`='Second',
    `3rd`='Third',
    `4th`='Fourth',
    `5th`='Fifth',
    `6th`='Sixth',
    `7th`='Seventh',
    `8th`='Eighth',
    `9th`='Ninth',
    `10th`='Tenth',
    `11th`='Eleventh',
    `12th`='Twelfth',
    `13th`='Thirteenth',
    `14th`='Fourteenth',
    `15th`='Fifteenth',
    `16th`='Sixteenth',
    `17th`='Seventeenth',
    `18th`='Eighteenth',
    `19th`='Nineteenth',
    `20th`='Twentieth',
    `21st`='Twenty-First',
    `22nd`='Twenty-Second',
    `23rd`='Twenty-Third',
    `24th`='Twenty-Fourth',
    `25th`='Twenty-Fifth',
    `26th`='Twenty-Sixth',
    `27th`='Twenty-Seventh',
    `28th`='Twenty-Eighth',
    `29th`='Twenty-Ninth',
    `30th`='Thirtieth',
    `31st`='Thirty-First',
    `32nd`='Thirty-Second',
    `33rd`='Thirty-Third',
    `34th`='Thirty-Fourth',
    `35th`='Thirty-Fifth',
    `36th`='Thirty-Sixth',
    `37th`='Thirty-Seventh'
)

ma_sen_12_16 <- read_csv("../../_data/2020/ma_pcts_districts_pvi_2016.csv") %>%
    rename(district = State_Sen) %>%
    mutate(district = str_replace_all(district, NUMS_TO_WORDS)) %>%
    group_by(district) %>%
    summarize(Obama_12 = sum(Obama_12),
              Romney_12 = sum(Romney_12),
              Clinton_16 = sum(Clinton_16),
              Trump_16 = sum(Trump_16)) %>%
    mutate(dem_pct_12 = Obama_12/(Obama_12 + Romney_12),
           dem_pct_16 = Clinton_16/(Clinton_16 + Trump_16))
    
ma_sen_20 <- read_csv("../../_data/2020/ma_precincts_districts_16_20_pres.csv") %>%
    rename(district = `State Senate`) %>%
    mutate(district = str_replace_all(district, NUMS_TO_WORDS)) %>%
    group_by(district) %>%
    summarize(Biden_20 = sum(Biden_20),
              Trump_20 = sum(Trump_20)) %>%
    mutate(dem_pct_20 = Biden_20/(Biden_20 + Trump_20))

ma_sen_12_20 <- ma_sen_12_16 %>%
    left_join(ma_sen_20, by="district") %>%
    mutate(shift_12_20 = dem_pct_20 - dem_pct_12)

ma_sen_12_20_geom <- sen_geom %>%
    left_join(ma_sen_12_20, by="district") %>%
    select(district,
           dem_pct_12,
           dem_pct_16,
           dem_pct_20,
           shift_12_20)

trend_popup_vars <- c(
    "dem_pct_12",
    "dem_pct_16",
    "dem_pct_20",
    "shift_12_20"
)    

percent_format <- function(pct) {
    percent(pct, digits=0)
}

point_format <- function(pnt) {
    digits((pnt*100), digits=0)
}

shift_legend_format <- list(
    shift_12_20=list(fun=point_format)
)

trend_popup_format <- list(
    dem_pct_12=list(fun=percent_format),
    dem_pct_16=list(fun=percent_format),
    dem_pct_20=list(fun=percent_format),
    shift_12_20=list(fun=point_format)
)

tmap_mode(mode="view")
```

There has been a great deal of interest in the MA State Senate race
for the Norfolk, Worcester and Middlesex District with incumbent
Senator Becca Rausch (D-Needham) and challenger State Representative
Shawn Dooley (R-Norfolk). The previous version of this State Senate
district&mdash;the Norfolk, Bristol and Middlesex&mdash;had been
considered solidly Republican and was represented by Republican
Richard Ross (R-Wrentham) who was ousted by Rausch in 2018, and
represented by Scott Brown (R-Wrentham) before Brown was elected to
the United States Senate in 2010.

Our simple, fundamentals-based model gives Rausch a very high
probability of winning with a range of likely outcomes from a single
digit Democratic margin to a major blowout. The major reason for this
change in circumstances is a continual multi-year shift from
Republican to Democratic strength in the affluent and well-educated
Boston suburbs.

## Regional Partisan Shifts

The rise of Donald Trump as the face of the Rebpulican party in the
United States has led to a profound shift in levels of Democratic and
Republican support in many areas of Massachusetts. This was evident
after the 2016 presidential election as described in a [thoughtful
investigation by WGBH's Adam Reilly](https://www.wgbh.org/news/politics/2018/10/30/in-blue-massachusetts-some-striking-redward-shifts)
supported by [a Mass. Numbers analysis](https://massnumbers.blogspot.com/2018/11/which-massachusetts-cities-and-towns.html).

The process has continued with the 2020 election as well-educated and
affluent suburbs have continued their move towards the Democratic
Party and away from a Trump-led GOP. There has also been a shift in
the rightward direction with much higher levels of support for
Republicans in rural areas with more non-college voters and also in
the gateway cities with large population of Hispanic (Lawrence) and
Portugese (Fall River, New Bedford) voter ancestry.

```{r partisan-shifts-map, layout="l-body-outset", fig.height=5, fig.cap="State Senate partisan shifts 2012-2020"}
senate_shift_map <- (
 tm_shape(ma_sen_12_20_geom) +
 tm_fill(col="shift_12_20",
         title="2012 to 2020 shift",
         palette=c("#FF4949", "white", "#0F4392"),
         midpoint=0,
         alpha=0.8,
         legend.format=shift_legend_format,
         popup.vars=trend_popup_vars,
         popup.format=trend_popup_format) +
 tm_borders(lwd=1, col="black") +
 tm_basemap("OpenStreetMap") +
 tm_view(set.view=8)
) 

tmap_leaflet(senate_shift_map)
```

This map of partisan shift from 2012 to 2020 presidential elections by
state senate district gives a strong picture of which areas are moving
in the Democratic direction (affluent suburbs of Boston) and in the
Republican direction (rural areas and gateway cities). Note that many
of the gateway cities still vote in large majorities for Democrats,
but the margins have come down significantly.

## Impact of shifts on State Senate race

The pre-2021 Norfolk, Bristol and Middlesex State Senate District
represented by Becca Rausch (D-Needham) is in second place behind the
Third Middlesex District of Michael Barrett (D-Lexington), with
respective shifts of 14 and 13 points in the Democratic direction
between 2012 and 2020. However, the Third Middlesex District has
gone from an already-very-blue 60% Democratic vote share to a
much higher 74%, while the Rausch district was within Republican
striking distance at 54% in 2012, but has moved to a 67% Democratic
share in 2020.

These shifts often have important ramifications on the results of
down-ballot state legislative races, which are reflected in our
fundamentals-based model using the Partisan Voter Index (PVI) which
is an average of the last two presidential races, adjusted by
the nationwide result.

An examination of the model's characterization of the races in the
Norfolk, Bristol and Middlesex District for 2016, 2018, and 2020, and
the similar Norfolk, Worcester and Middlesex District for 2022,
show how the playing field has changed. The model's adjustment
for non-presidential years would have significantly reduced
Senator Rausch's chances in 2022 compared to 2020 if it had
not been for the continuing shift of the region towards
Democratic candidates in the 2020 election which raised
the PVI from D+6 to D+12, giving Rausch about the same range
of outcomes in 2022 as in 2020.

```{r leg-model}
ma_leg <- read_csv("../../_data/2020/ma_leg_two_party_2008_2021_pvi.csv") %>%
    mutate(dem_win = if_else(dem_percent > 0.5, 1, 0)) %>%
    filter(!is_special)

bm2 <- stan_glm(dem_win ~ PVI_N + incumbent_status + pres_year, data=ma_leg, family=binomial(link="logit"), refresh=0)
blm2 <- stan_glm(dem_percent ~ PVI_N + incumbent_status + pres_year, data=ma_leg, family=gaussian(link="identity"), refresh=0)
```

```{r nbm-nwm-setup}
nbm_pred_results <- tibble(
    year = c(2016, 2018, 2020, 2022),
    PVI_N = c(2.731936, 6.149234, 6.149234, 11.559097),
    PVI = c("D+3", "D+6", "D+6", "D+12"),
    incumbent_status = c("GOP_Incumbent", "GOP_Incumbent", "Dem_Incumbent", "Dem_Incumbent"),
    pres_year = c(TRUE, FALSE, TRUE, FALSE),
    label=c("Ross (R) over Aleksov (D)",
            "Rausch (D) over Ross (R)",
            "Rausch (D) over Kelly (R)",
            NA)
) %>%
    mutate(dem_win_prob = predict(bm2, newdata=., type="response"),
           pred_dem_pct = predict(blm2, newdata=., type="response"),
           pred_dem_pct_low = predictive_interval(blm2, newdata=.)[,1],
           pred_dem_pct_high = predictive_interval(blm2, newdata=.)[,2],
           pred_dem_margin = pred_dem_pct - (1 - pred_dem_pct),
           pred_dem_margin_low = pred_dem_pct_low - (1 - pred_dem_pct_low),
           pred_dem_margin_high = pred_dem_pct_high - (1 - pred_dem_pct_high)) %>%
   mutate(dem_margin = c(-0.201, 0.0268, 0.195, NA),
          dem_percent = c(0.399, 0.513, 0.597, NA))
```

```{r nbm-nwm-plot, layout="l-body-outset", fig.cap="Norfolk, {Bristol, Worcester}, Middlesex model vs. outcome"}
nbm_nwm_plot <- nbm_pred_results %>%
    ggplot(aes(x=year,
               y=pred_dem_margin,
               ymin=pred_dem_margin_low,
               ymax=pred_dem_margin_high)) +
    geom_pointrange(size=0.2, shape="plus") +
    geom_point(aes(y=dem_margin), color="blue", size=2) +
    geom_text(aes(y=dem_margin, label=label), hjust="left", nudge_x=0.06, size=3) +
    geom_hline(yintercept=0, linetype="dashed") +
    scale_y_continuous(labels = scales::percent_format(accuracy=2)) +
    labs(y="Democratic vote margin") +
    theme_minimal()

nbm_nwm_plot
```

```{r nbm-nwm-table, layout="l-body-outset"}
nbm_table_vals <- nbm_pred_results %>%
    mutate(incumbent = case_when(
                          incumbent_status == "Dem_Incumbent" ~ "Dem",
                          incumbent_status == "GOP_Incumbent" ~ "GOP",
                          TRUE ~ ""),
           candidates = if_else(is.na(label),
                                "Rausch (D) vs. Dooley (R)",
                                label)) %>%
    select(year,
           incumbent,
           pres_year,
           candidates,
           result=dem_margin,
           pred_dem_margin,
           margin_low=pred_dem_margin_low,
           margin_high=pred_dem_margin_high)

nbm_table <- nbm_table_vals %>%
    gt() %>%
    gt_theme_538() %>%
    sub_missing(missing_text = "") %>%
    fmt_percent(columns=c("result",
                          "pred_dem_margin",
                          "margin_low",
                          "margin_high"),
                decimals=0) %>%
    tab_header("Norfolk, {Bristol, Worcester}, Middlsex model vs. outcome")

nbm_table
```

While a win by State Representative Shawn Dooley (R-Norfolk) is not an
impossible outcome, the profound shift of the Boston suburbs away
from the Republican Party make a win by incumbent Becca Rausch
much more likely.
