---
title: "MA 2022 State Senate General Election Ratings"
description: |
  A model-based evaluation of the contested Massachusetts State Senate races
author: Brent Benson
date: 2022-10-31
citation: false
preview: https://www.massnumbers.us/images/ma_state_senate_ratings_2022.png
output:
  distill::distill_article:
    self_contained: false
draft: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)

library(metathis)
library(tidyverse)
library(readxl)
library(rstanarm)
library(humaniformat)
library(gt)
library(gtExtras)
library(sf)
library(tmap)
library(formattable)
```

```{r post-meta}
meta() %>%
    meta_description("A model-based evaluation of the contested State Senate races") %>%
    meta_viewport() %>%
    meta_social(
        title = "MA 2022 State Senate General Election Ratings",
        url = "https://www.massnumbers.us/posts/2022-10-31-state-senate-ratings/",
        image = "https://www.massnumbers.us/images/ma_state_senate_ratings_2022.png",
        image_alt = "MA 2020 State Senate general election ratings",
        og_type = "article",
        og_author = "Brent Benson",
        twitter_card_type = "summary_large_image",
        twitter_creator = "@bwbensonjr"
    )
```

```{r read-data}
party_abbrev <- tibble(
    party = c("Democratic",
              "Independent",
              "Republican",
              "Unenrolled",
              "Workers Party",
              "Green-Rainbow Party",
              "We The People",
              "Non-Party Candidate"),
    pty = c("D",
            "I",
            "R",
            "U",
            "W",
            "GR",
            "WTP",
            "NP")
)

sc <- read_csv("../../_data/2020/ma_state_senate_cands_2022.csv") %>%
    left_join(party_abbrev, by="party") %>%
    mutate(cand_name_parsed = parse_names(candidate),
           first_name = cand_name_parsed$first_name,
           last_name = cand_name_parsed$last_name,
           cand_label = str_c(last_name, " (", pty, "-", city_town, ")")) %>%
    select(-office, -cand_name_parsed) %>%
    nest(candidate = c(candidate, address, city_town, party, pty, first_name, last_name, cand_label)) %>%
    mutate(num_candidates = map_dbl(candidate, nrow))

sr <- read_excel("../../_data/2020/MA_State_Senate_Race_2022.xlsx") %>%
    select(district = District,
           dist_num = Number,
           PVI_N,
           PVI,
           inc_status,
           Incumbent,
           dem_win_prob) %>%
    left_join(sc, by="district") %>%
    unnest(candidate) %>%
    group_by(district) %>%
    mutate(is_incumbent = case_when(
              ((inc_status == "Dem_Incumbent") & (pty == "D")) ~ TRUE,
              ((inc_status == "GOP_Incumbent") & (pty == "R")) ~ TRUE,
              TRUE ~ FALSE)) %>%
    ungroup() %>%
    nest(candidate = c(candidate,
                       address,
                       city_town,
                       party,
                       pty,
                       cand_label,
                       first_name,
                       last_name,
                       is_incumbent))

```

```{r leg-model}
ma_leg <- read_csv("../../_data/2020/ma_leg_two_party_2008_2021_pvi.csv") %>%
    mutate(dem_win = if_else(dem_percent > 0.5, 1, 0)) %>%
    filter(!is_special)

bm2 <- stan_glm(dem_win ~ PVI_N + incumbent_status + pres_year, data=ma_leg, family=binomial(link="logit"), refresh=0)
blm2 <- stan_glm(dem_percent ~ PVI_N + incumbent_status + pres_year, data=ma_leg, family=gaussian(link="identity"), refresh=0)
```

```{r infer-races}
sr$bayes_win_prob <- predict(bm2, newdata=(sr %>% rename(incumbent_status=inc_status) %>% mutate(pres_year = FALSE)), type="response")
sr$bayes_dem_pct <- predict(blm2, newdata=(sr %>% rename(incumbent_status=inc_status) %>% mutate(pres_year = FALSE)), type="response")
sen_intervals <- predictive_interval(blm2, newdata=(sr %>% rename(incumbent_status=inc_status) %>% mutate(pres_year = FALSE)))
sr$bayes_dem_low <- sen_intervals[,1]
sr$bayes_dem_high <- sen_intervals[,2]
```

```{r table-transforms}
flatten_candidates <- function(district, cands) {
    cands %>%
        select(pty, cand_label) %>%
        mutate(pty = case_when(
                   pty == "D" ~ "D",
                   pty == "R" ~ "R",
                   TRUE ~ "other")) %>%
        pivot_wider(names_from = "pty",
                    values_from = "cand_label")
}

table_transforms <- function(df) {
    df %>%
        mutate(margin = (bayes_dem_pct - 0.5)*100,
               margin_low = (bayes_dem_low - 0.5)*100,
               margin_high = (bayes_dem_high - 0.5)*100,
               Incumbent = case_when(
                   inc_status == "Dem_Incumbent" ~ "Dem",
                   inc_status == "GOP_Incumbent" ~ "GOP",
                   TRUE ~ ""),
               candidates = map2(district, candidate, flatten_candidates)) %>%
        unnest(candidates) %>%
        select(district, PVI, PVI_N, Incumbent, num_candidates,
               D, R, other, dem_win_prob=bayes_win_prob,
               margin, margin_low, margin_high)
}

sr_flat <- sr %>%
    table_transforms()
```

```{r map-setupe}
tmap_mode(mode="view")

sen_geom <- read_sf("../../_gis/SENATE2021/SENATE2021_POLY.shp") %>%
    st_make_valid() %>%
    st_simplify(dTolerance=100, preserveTopology=TRUE) %>%
    select(district = SEN_DIST) %>%
    right_join(sr_flat, by="district") %>%
    mutate(contested = (num_candidates > 1),
           margin = if_else(contested, margin, NA_real_),
           margin_low = if_else(contested, margin_low, NA_real_),
           margin_high = if_else(contested, margin_high, NA_real_),
           dem_win_prob = if_else(contested, dem_win_prob, NA_real_))

race_popup_vars <- c(
    "PVI",
    "Incumbent",
    "D", "R", "other",
    "dem_win_prob",
    "margin",
    "margin_low",
    "margin_high"
)

percent_zero <- function(n) {
    percent(n, 0)
}

digits_zero <- function(n) {
    digits(n, digits=0)
}

race_popup_format <- list(
    dem_win_prob=list(fun=percent_zero),
    margin=list(fun=digits_zero),
    margin_low=list(fun=digits_zero),
    margin_high=list(fun=digits_zero)
)
```

```{r state-senate-map, layout="l-body-outset", fig.height=5, fig.cap="Contested State Senate Races"}
state_senate_map <- (
    tm_shape(sen_geom) +
    tm_fill(col="margin",
            palette=c("#FF4949", "white", "#0F4392"),
            midpoint=0,
            alpha=0.7,
            textNA="Single candidate",
            popup.format=race_popup_format,
            popup.vars=race_popup_vars) +
    tm_borders(lwd=1) +
    tm_basemap("OpenStreetMap") +
    tm_view(set.view=8)
)
state_senate_map
```

Just over half of the Massachusetts State Senate seats&mdash;22 out of
40&mdash;have a contested general election race, with a likely safe
GOP hold in the First Plymouth and Norfolk District represented by
Senator Patrick O'Connor (R-Weymouth), and Senator Jamie Eldridge
(D-Acton) in a redistricted D+20 Middlesex and Worcester District that
is unlikely to end with a Republican pickup.

We will use a model-based approach to understand the competitive
landscape of all of the contested State Senate races with respect to
those extremes using a simple regression model trained with historical
data to provide an estimated range of likely outcomes.

## The Model

The Bayesian regression model used for this analysis is described
in detail in the [MA 2022 State Representative Ratings](https://www.massnumbers.us/posts/2022-10-31-state-senate-ratings/).
Briefly, the model has three variables: a measure of partisanship
(the Partisan Voter Index), an incumbency factor (Democratic incumbent, no
incumbent, or GOP incumbent), and an adjustment for presidential election years.
The model is trained with 424 state legislative elections from 2008 through 2021
and explains a little over 70 percent of variance in the Democratic percentage
of the vote. The other 30 percent of the variance is due to important factors
like candidate and campaign quality, which will determine the results in
competitive races. The general, fundamentals-based nature of the model is also
why these ratings should only be used to understand the competitive landscape
rather than as a prediction of race results.

As an example, let's take the Hampden, Hampshire and Worcester race
between sitting State Representative Jake Oliveira (D-Ludlow) and
business owner William Johnson (R-Granby). The Partisan Voter Index
(PVI) of the redistricted Hampden, Hampshire and Worcester district is
D+4 (3.50 with more precision) meaning that over the last two
presidential elections the district is 3.5 points more Democratic than
the country as a whole. The incumbency variable is set to no incumbent
as State Senator Eric Lesser (D-Longmeadow) ran for Lt. Governor,
rather than for re-election. The presidential election variable is
false, as it is for all the 2022, off-year races.

Using the trained model to probabilistically simulate the election
several thousand times (*sampling from the posterior distribution* in
technical parlance) we see the Oliveira winning 59% of the time, and
Johnson winning 41% of the time. If we look at the simulated
Democratic percent of the vote, rather than the win/loss percentage,
we see an average Democratic percent of 52% of the vote&mdash;making
this the most competitive State Senate seat with respect to this
model. We can also generate a 90% confidence interval for the margin
with the Democratic percentage varying from 41% to 63%&mdash;a wide
enough interval to allow for a comfortable GOP win, or a large
Democratic victory. The most likely outcome will be a reasonably
narrow victory for Oliveira or Johnson.

## Race Ratings

For each race we show the candidates, district Partisvan Voter Index (PVI),
and the output of the model: a Democratic win probability, and a projected
Democratic one-way margin (points above 50/50 tie) with a 90% confidence
interval. It is important to remember that this fundamentals-based model
does not take into account particulars of each race including important
factors like candidate quality and campaign efforts and outreach, which will
actually determine the results in competitive races.

In races with a Republican, a third-party, unenrolled, or independent
candidate, and no Democratic candidate, the Democratic win percentage
and margin should be attributed to the non-Republican candidate.

The highlighted races in the table have both candidates with at least
a 10% chance of winning and non-trivial margins on both sides of the
break-even margin line.

```{r state-senate-table, layout="l-screen"}
state_sen_table <- sr_flat %>%
    filter(num_candidates > 1) %>%
    select(-num_candidates) %>%
    arrange(margin) %>%
    gt() %>%
    gt_theme_538() %>%
    sub_missing(missing_text = "") %>%
    fmt_percent(columns="dem_win_prob", decimals=0) %>%
    gt_plt_conf_int(margin,
                    c(margin_low, margin_high),
                    palette=c("black", "black", "white", "black"),
                    ref_line=0.5,
                    text_size=2.5,
                    width=55) %>%
    cols_hide(columns=c(PVI_N, margin_low, margin_high)) %>%
    cols_align(columns=c("margin"), align="center") %>%
    gt_highlight_rows(font_weight="normal",
                      rows=((dem_win_prob > .095) & (dem_win_prob < .9))) %>%
    tab_header("Contested State Senate race ratings")

state_sen_table
```

