---
title: "First post-presidential special election shows Trump era Democratic strength"
description: |
    Hannah Bowen won the Massachusetts 6th Essex State Rep district by a much higher margin than Kamala Harris six months ago
author: Brent Benson
date: 2025-05-15
citation: false
preview: https://www.massnumbers.us/images/sixth-essex-special-diff.png
output:
  distill::distill_article:
    self_contained: false
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(gt)
library(gtExtras)
library(sf)
library(tmap)
library(formattable)
library(metathis)

tmap_mode("view")
```

```{r post-meta}
meta() %>%
    meta_description("Hannah Bowen won the Massachusetts 6th Essex State Rep district by a much higher margin than Kamala Harris six months ago") |>
    meta_viewport() |>
    meta_social(
        title = "First post-presidential special election shows Trump era Democratic strength",
        url = "https://www.massnumbers.us/posts/2025-05-15-sixth-essex-special/",
        image = "https://www.massnumbers.us/images/sixth-essex-special-diff.png",
        image_alt = "Sixth Essex Democratic vote share difference - 2025 special vs. 2024 presidential",
        og_type = "article",
        og_author = "Brent Benson",
        twitter_card_type = "summary_large_image",
        twitter_creator = "@bwbensonjr"
    )
```

```{r}
precinct_display <- function(city_town, ward, precinct) {
     if_else(ward == "-",
             str_c(city_town, " ", precinct),
             str_c(city_town, " ", ward, "-", precinct))
}

sixth_essex <-
    read_csv("sixth_essex_2025_05_13.csv", col_types=list(precinct=col_character())) |>
    mutate(Harris_pct_24 = Harris_24/(Harris_24 + Trump_24),
           Trump_pct_24 = Trump_24/(Harris_24 + Trump_24),
           point_diff_24 = (Harris_pct_24 - Trump_pct_24) * 100,
           Bowen_pct_25 = Bowen_25/(Bowen_25 + Long_25),
           Long_pct_25 = Long_25/(Bowen_25 + Long_25),
           point_diff_25 = (Bowen_pct_25 - Long_pct_25) * 100,
           point_diff = (Bowen_pct_25 - Harris_pct_24) * 100,
           precinct_name = precinct_display(city_town, ward, precinct))
```

On Tuesday, May 13 non-profit executive Hannah Bowen (D-Beverly)
defeated digital marketing professional Medley Long III (R-Beverly) in
a special election for the Sixth Essex State Representative
district. Bowen will replace Jared Parisella (D-Beverly) who was
[appointed to a District Court
judgeship](https://www.bostonglobe.com/2024/11/26/metro/healey-nominates-state-democratic-lawmaker-beverly-judgeship/)
by Governor Maura Healey.

Bowen's victory was not a surprise given the D+18 Partisan Voter Index
(PVI) rating of the Beverly-based Sixth Essex district, but her margin
of 48 points was striking because it is 12 percentage points higher than
Kamala Harris's 35 point victory over Trump in the district.


```{r layout="l-body-outset", fig.height=5, fig.cap="Sixth Essex Democratic vote share difference - 2025 special vs. 2024 presidential"}
sixth_essex_geom <-
    read_sf("http://bwbensonjr.github.io/mapoli/gis/geojson/wardsprecincts2022.geojson") |>
    select(-precinct_name) |>
    right_join(sixth_essex,
               by=c("city_town", "ward", "precinct"))

percent_zero <- function(n) {
    percent(n, 0)
}

district_map <-
   (tm_shape(sixth_essex_geom) +
   tm_polygons(
      col="point_diff",
      title="Margin Difference",
      popup.vars=c(
          "Margin Difference" = "point_diff",
          "Bowen (D)" = "Bowen_pct_25",
          "Long (R)" = "Long_pct_25",
          "Bowen Margin" = "point_diff_25",
          "Harris (D) - 2024" = "Harris_pct_24",
          "Trump (R) - 2024" = "Trump_pct_24",
          "Harris Margin" = "point_diff_24"
      ),
      popup.format=list(
          point_diff=list(digits=0),
          point_diff_25=list(digits=0),
          point_diff_24=list(digits=0),
          Bowen_pct_25=list(fun=percent_zero),
          Long_pct_25=list(fun=percent_zero),
          Harris_pct_24=list(fun=percent_zero),
          Trump_pct_24=list(fun=percent_zero)
      ),
      alpha=0.6,
      palette=c("#FF4949", "white", "#0F4392"),
      midpoint=0,
      id="precinct_name"
   ) +
   tm_text("precinct_name") +
   tm_basemap("OpenStreetMap")
)

tmap_leaflet(district_map)
```

We can see by looking at the individual precinct-level results that
Rep-Elect Bowen outperformed the Democratic presidential candidate in
every precinct except Beverly Ward 6, Precinct 1, which happens to be
the home precinct of Republican candidate Medley Long III. Bowen still
won this precinct by a 20 point margin and achieved a remarkable 85%
of the vote in two precincts, about 10 points higher than Harris in
each.


```{r layout="l-body-outset", fig.cap="Sixth Essex 2025 Special Election vs. 2024 Presidential"}

district_table <-
    sixth_essex |>
    select(
        precinct_name,
        Bowen_pct_25,
        Long_pct_25,
        point_diff_25,
        Harris_pct_24,
        Trump_pct_24,
        point_diff_24,
        point_diff
    ) |>
    gt() |>
    # tab_header(
    #     title="Sixth Essex 2025 Special Election vs. 2024 Presidential"
    # ) |>
    tab_spanner(label="2024", columns=ends_with("_24")) |>
    tab_spanner(label="2025", columns=ends_with("_25")) |>
    cols_label(
        precinct_name="Precinct",
        Bowen_pct_25="Bowen (D)",
        Long_pct_25="Long (R)",
        point_diff_25="Margin",
        Harris_pct_24="Harris (D)",
        Trump_pct_24="Trump (R)",
        point_diff_24="Margin",
        point_diff=""
    ) |>
    fmt_percent(
        columns=contains("_pct"),
        decimals=0
    ) |>
    fmt_number(
        columns=starts_with("point_diff"),
        decimals=0
    ) |>
    gt_plt_bar(
        column=point_diff,
        keep_column=TRUE,
    ) |>
    cols_label(DUPE_COLUMN_PLT = "") |>
    tab_spanner(
        label="Margin Difference",
        columns=c(point_diff, DUPE_COLUMN_PLT)
    ) |>
    tab_footnote(
        footnote="Home precinct of GOP candidate Medley Long III",
        locations=cells_body(
            columns=precinct_name,
            rows=(precinct_name == "Beverly 6-1")
        )
    )

district_table
```

We can't extrapolate too much from a single special election, but the
results seem to show Democrats in a strong position as Donald Trump
continues to take actions that anger a wide swath of Massachusetts
voters, causing likely headwinds for Massachusetts GOP candidates.

There will be another special election test on June 10th in the Third
Bristol district previously help by Representative Carol Doherty
(D-Taunton) who died in February from pancreatic cancer. Lisa Field
(D-Taunton) of the Massachusetts Nurses Association will face funeral
home owner Lawrence Quintal (R-Taunton). The Third Bristol district is
a swing district with a PVI of D+5. It was represented by Republican
Shaunna O'Connell (R-Taunton) for ten years before she was elected
Mayor of Taunton in 2020.





