---
title: "An overview of the 2022 State Senate Primaries"
description: |
    A look at the ten contested MA State Senate primaries
author:
  - name: Brent Benson
    url: https://twitter.com/bwbensonjr
date: 2022-08-25
citation: false
preview: https://www.massnumbers.us/images/berk_hamp_frank_hamp.png
output:
  distill::distill_article:
    self_contained: false
---

```{r libraries, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
options(digits=1, knitr.kable.NA="")

library(metathis)
library(kableExtra)
library(formattable)
library(tidyverse)
library(forcats)
library(sf)
library(tmap)
library(readxl)
library(viridis)
library(tidycensus)
library(tigris)
library(units)
library(nominatimlite)
```

```{r post-meta}
meta() %>%
    meta_description("A look at the ten contested MA State Senate primaries") %>%
    meta_viewport() %>%
    meta_social(
        title = "An overview of the 2022 State Senate Primaries",
        url = "https://www.massnumbers.us/posts/2022-07-05-state-senate-primary-preview/",
        image = "https://www.massnumbers.us/images/berk_hamp_frank_hamp.png",
        image_alt = "The precincts of the Berkshire, Hampshire, Frankling and Hampden district",
        og_type = "article",
        og_author = "Brent Benson",
        twitter_card_type = "summary_large_image",
        twitter_creator = "@bwbensonjr"
    )
```

```{r senate-districts}
sen_dist <- read_excel("../../_data/2020/MA_State_Senate_Race_2022.xlsx",
                       sheet="Districts")

leg_ocpf <- read_csv("../../_data/2020/ma_leg_ocpf_summary_july_2022.csv")

sen_cand <- read_csv("../../_data/2020/ma_state_senate_cands_2022_contested_addr.csv") %>%
    rename(Inc = Incumbent) %>%
    mutate(address = str_c(street_address, city_town, "MA", sep=", ")) %>%
    select(-street_address) %>%
    left_join(leg_ocpf, by="cpf_id")

sen_geom <- read_sf("../../_gis/SENATE2021/SENATE2021_POLY.shp", as_tibble=TRUE) %>%
    st_make_valid() %>%
    select(Number=SENDISTNUM) %>%
    st_transform(st_crs(6491)) %>%
    left_join(sen_dist, by="Number") %>%
    relocate(District)

sen_geom_12 <- read_sf("../../_gis/senate2012/SENATE2012_POLY.shp", as_tibble=TRUE) %>%
    st_make_valid() %>%
    st_transform(st_crs(6491))

tmap_mode(mode="view")
one_m_sq <- set_units(1, m^2)
two_k_m_sq <- set_units(2000, m^2)
five_k_m_sq <- set_units(5000, m^2)
```
```{r district-setup}
vtd_to_city_town <- read_csv("../../_data/2020/ma_vtds_city_town_fips.csv") %>%
    select(vtd_fips, city_town, Ward, Pct, city_town_fips, type)

ma_vtds <- voting_districts(state=25) %>%
    rename(state_code = STATEFP20,
           county_code = COUNTYFP20,
           vtd_code = VTDST20,
           vtd_fips = GEOID20) %>%
    filter(vtd_code != "ZZZZZZ") %>%
    left_join(vtd_to_city_town, by="vtd_fips") %>%
    st_transform(st_crs(6491))

ma_towns <- read_sf("../../_gis/townssurvey_shp/TOWNSSURVEY_POLYM.shp",
                    as_tibble=TRUE) %>%
    st_make_valid() %>%
    mutate(city_town = str_to_title(TOWN)) %>%
    select(city_town) %>%
    st_transform(st_crs(6491))

ma_towns_dems <- read_csv("../../_data/2020/ma_city_town_demographics_pvi_2020.csv") %>%
    mutate(city_town = str_replace(city_town, "Manchester-by-the-Sea", "Manchester")) %>%
    rename(white_pct = race_white_pct,
           black_pct = race_black_pct,
           hispanic_pct = race_hispanic_pct,
           asian_pct = race_asian_pct) %>%
    select(city_town_fips, city_town, PVI, PVI_N, total_population,
           white_pct, black_pct, hispanic_pct, asian_pct)

ma_towns_geom <- merge(ma_towns, ma_towns_dems, by="city_town") %>%
    mutate(center = st_centroid(geometry)) %>%
    relocate(city_town)

ma_vtds_pvi <- read_csv("../../_data/2020/ma_precinct_pvi_2020.csv")
    
ma_vtds_dems <- read_csv("../../_data/2020/ma_census_2020_daves.csv") %>%
    rename(total_population = Tot_2020_tot) %>%
    mutate(white_pct = Wh_2020_tot/total_population,
           black_pct = BlC_2020_tot/total_population,
           hispanic_pct = His_2020_tot/total_population,
           asian_pct = AsnC_2020_tot/total_population) %>%
    select(vtd_fips, total_population, white_pct, black_pct,
           hispanic_pct, asian_pct)

ma_vtds_geom <- ma_vtds %>%
    mutate(center = st_centroid(geometry)) %>%
    left_join(ma_vtds_pvi, by=c("city_town", "Ward", "Pct")) %>%
    left_join(ma_vtds_dems, by="vtd_fips")

ma_sen <- read_sf("../../_gis/SENATE2021/SENATE2021_POLY.shp", as_tibble=TRUE) %>%
    st_make_valid() %>%
    select(SENDISTNUM, SEN_DIST) %>%
    st_transform(st_crs(6491))

ma_house_12 <- read_sf("../../_gis/house2012/HOUSE2012_POLY.shp", as_tibble=TRUE) %>%
    st_make_valid() %>%
    st_transform(st_crs(6491))

ma_city_town_num_pcts <- ma_vtds_geom %>%
    st_drop_geometry() %>%
    group_by(city_town) %>%
    summarize(num_precincts = n())

ward_precinct_name <- function(city_town, ward, precinct) {
    str_c(city_town,
          if_else(ward == "-", "", str_c("Ward ", ward)),
          str_c("Precinct ", precinct),
          sep=" ")
}

## Determine which towns and precincts are in the district.
## If a whole town is contained in the district, include the
## whole town, otherwise, include by precinct.
##
town_precinct_mix <- function(sen_dist_geom, towns_geom, vtds_geom, city_town_num_pcts) {
    towns <- st_intersection(sen_dist_geom, towns_geom) %>%
        filter(st_area(geometry) > five_k_m_sq)
    vtds <- st_intersection(sen_dist_geom, vtds_geom) %>%
        filter(st_area(geometry) > one_m_sq)
    num_pcts <- vtds %>%
        st_drop_geometry() %>%
        group_by(city_town) %>%
        summarize(num_precincts = n())
    completes <- num_pcts %>%
        left_join((city_town_num_pcts %>% rename(total_precincts = num_precincts)), by="city_town") %>%
        mutate(complete = (num_precincts == total_precincts))
    full_towns <- right_join(
            towns,
            (completes %>%
             filter(complete) %>%
             select(city_town)),
            by="city_town") %>% 
        select(name=city_town, PVI, PVI_N, total_population,
               white_pct, black_pct, hispanic_pct, asian_pct)
    precincts <- right_join(
            vtds,
            (completes %>%
             filter(!complete) %>%
             select(city_town)),
            by="city_town") %>%
        mutate(name=ward_precinct_name(city_town, Ward, Pct)) %>%
        select(name, PVI, PVI_N, total_population, white_pct,
               black_pct, hispanic_pct, asian_pct)
    towns_precincts <- rbind(full_towns, precincts)
    towns_precincts
}

## Return the candidates for a particular district and compute
## candidate locations for placing on map.
## 
candidate_locations <- function(cand_list, sen_dist_num) {
    cand_list %>%
        filter(SENDISTNUM == sen_dist_num) %>%
        relocate(Candidate) %>%
        mutate(geometry = (geo_lite_sf(address,
                                       limit=1,
                                       return_addresses=FALSE) %>%
                               pull(geometry))) %>%
        st_as_sf()
}

redistrict_map <- function(sen_dist, towns_pcts, prev_dist) {
    (tm_shape(towns_pcts, bbox=st_bbox(st_union(sen_dist, prev_dist))) +
     tm_borders(lwd=0.5) +
     tm_fill(col="minority_pct",
             style="fixed",
             breaks=c(0, 0.2, 0.5, 0.7, 1),
             palette="PRGn",
             midpoint=0.5,
             alpha=0.7,
             popup.vars=c("PVI", "total_population", "minority_pct", "white_pct", 
                         "black_pct", "hispanic_pct", "asian_pct")) +
    tm_shape(sen_dist) + tm_borders(lwd=3) +
    tm_shape(prev_dist) + tm_borders(lwd=2.5, lty="dotted") +
    tm_basemap("OpenStreetMap"))
}

district_map <- function(sen_dist, towns_pcts, candidates, house_dist_1, house_dist_2) {
    if (missing(house_dist_1)) {
        bbox <- NULL
    } else if (missing(house_dist_2)) {
        bbox <- st_bbox(st_union(sen_dist, house_dist_1))
    } else {
        bbox <- st_bbox(st_union(sen_dist, house_dist_1, house_dist_2))
    }
    base_map <- (tm_shape(towns_pcts, bbox=bbox) +
                 tm_borders(lwd=1) +
                 tm_fill(col="PVI_N",
                         palette=c("#FF4949", "white", "#0F4392"),
                         midpoint=0,
                         alpha=0.5,
                         popup.vars=c("PVI", "total_population", "white_pct",
                                      "black_pct", "hispanic_pct", "asian_pct")) +
                 tm_shape(sen_dist) + tm_borders(lwd=3))
    if (!missing(house_dist_1)) {
        base_map <- base_map + (tm_shape(house_dist_1) + tm_borders(lwd=2.5, lty="dashed"))
    }
    if (!missing(house_dist_2)) {
        base_map <- base_map + (tm_shape(house_dist_2) + tm_borders(lwd=2.5, lty="dotted"))
    }
    (base_map + 
      tm_shape(candidates) +
      tm_symbols(ymod=-0.6, size=0.2, col="black", popup.vars=c("Party", "city_town")) +
      tm_basemap("OpenStreetMap"))
}

district_map_plot <- function(sen_dist, towns_pcts, candidates, house_dist_1, house_dist_2) {
    if (missing(house_dist_1)) {
        bbox <- NULL
    } else if (missing(house_dist_2)) {
        bbox <- st_bbox(st_union(sen_dist, house_dist_1))
    } else {
        bbox <- st_bbox(st_union(sen_dist, house_dist_1, house_dist_2))
    }
    base_map <- (tm_shape(towns_pcts, bbox=bbox) +
                 tm_borders(lwd=0.5) +
                 tm_fill(col="PVI_N",
                         palette=c("#FF4949", "white", "#0F4392"),
                         midpoint=0,
                         alpha=0.5,
                         popup.vars=c("PVI", "total_population", "white_pct",
                                      "black_pct", "hispanic_pct", "asian_pct")) +
                 tm_shape(sen_dist) + tm_borders(lwd=3))
    if (!missing(house_dist_1)) {
        base_map <- base_map + (tm_shape(house_dist_1) + tm_borders(lwd=2.5, lty="dashed"))
    }
    if (!missing(house_dist_2)) {
        base_map <- base_map + (tm_shape(house_dist_2) + tm_borders(lwd=2.5, lty="dotted"))
    }
    (base_map + 
      tm_shape(candidates) +
      tm_symbols(ymod=-0.6, size=0.2, col="black", popup.vars=c("Party", "city_town")))
}

## Produce a readable version of candidates for a particular Senate district
candidate_table <- function(candidates, dist_num) {
    dist_candidates <- candidates %>%
        filter(SENDISTNUM == dist_num)

    caption <- str_c((dist_candidates %>%
                          slice(n=1) %>%
                          pull(District)),
                     " candidates (year-to-date $)")

    dist_candidates %>%
        mutate(Candidate = str_c("[", Candidate, "](", url, '){target="_blank"}', sep="")) %>%
        select(Inc,
               Candidate,
               Party,
               city_town,
               receipts,
               spent = expenditures,
               cash = end) %>%
        mutate(receipts = currency(receipts, digits=0),
               spent = currency(spent, digits=0),
               cash = currency(cash, digits=0)) %>%
        kbl(caption=caption) %>%
        kable_styling()
}
```

```{r new-district-map, layout="l-body-outset", fig.cap="*MA State Senate Districts (2022)*"}
new_sen_map <- (tm_shape(sen_geom) +
                    tm_borders(lwd=2) +
                    tm_fill(col="PVI_N",
                            palette=c("#FF4949", "white", "#0F4392"),
                            midpoint=0,
                            alpha=0.5,
                            popup.vars=c("Incumbent",
                                         "PVI",
                                         "NH_White_Pct",
                                         "Old District",
                                         "primaries")) +
                    ## tm_shape(sen_geom_12) +
                    ## tm_borders(lwd=2, lty="dashed") +
                    tm_basemap("OpenStreetMap"))
tmap_leaflet(new_sen_map)
```

## State Senate Primary Overview

There are ten State Senate districts with contested primaries. There
are three Democratic incumbents with primary challengers—Joan Lovely
(D-Salem), Walter Timility (D-Milton), Michael Brady (D-Brockton)—and
one Republican with a primary challenge, Patrick O'Connor
(R-Weymouth).

There are five open seats. Four of the open seats have a Democratic
primary only, and one seat—Hampden, Hampshire and Worcester—has
both a Democratic and Republican primary.

```{r state-senate-districts}
sen_dist %>%
    mutate(inc_status = fct_recode(inc_status,
                                   Dem="Dem_Incumbent",
                                   GOP="GOP_Incumbent",
                                   None="No_Incumbent")) %>%
    select(c(District,
             PVI,
             `Inc. Pary`=inc_status,
              Incumbent,
              primaries)) %>%
    kbl(caption="2022 State Senate districts (last incumbent)*",) %>%
    kable_styling()
```

## District/Primary Breakdowns

In the following sections focusing on each district/primary race, the
map shows a measure of partisanship (PVI) of each town (or precinct in
districts where cities or towns are split), overlaps of any incumbent
districts, and a small dot indicating the home region of each
candidate.

## Berkshire, Hampden, Franklin and Hampshire

```{r bhfh-setup}
ma_sen_1 <- ma_sen %>%
    filter(SENDISTNUM == 1)

ma_towns_1 <-st_intersection(ma_sen_1, ma_towns_geom) %>%
    filter(st_area(geometry) > five_k_m_sq) %>%
    relocate(city_town)

ma_house_berk_2 <- ma_house_12 %>%
    filter(DIST_CODE == "BERK02")

ma_sen_cands_1 <- candidate_locations(sen_cand, 1)

```

```{r bhfh-map, fig.cap="*Berkshire, Hampden, Franklin and Hampshire*   "}
ma_sen_1_map <- district_map(ma_sen_1, ma_towns_1, ma_sen_cands_1, ma_house_berk_2)
tmap_leaflet(ma_sen_1_map)
```

```{r bhfh-cand}
candidate_table(sen_cand, 1)
```

The newly created and enormous Berkshire, Hampden, Franklin and Hampshire
district contains 57 cities and towns in Western Massachusetts. The incumbent of
the existing and similarly named Berkshire, Hampshire, Franklin and Hampden
district, Adam Hinds (D-Pittsfield) ran an unsuccessful campaign for Lieutenant
Governor, leaving this district without an incumbent senator.

The Democratic primary will feature sitting State Representative Paul
Mark (D-Becket) whose current Second Berkshire State Representative
district is outlined in the map, and Williamstown former small
business owner and local municipal board member Huff Templeton
III. Representative Mark will have an advantage as about 20% of the
Senate district's Democratic primary voters will have seen him on the
ballot (using an average of vote totals from previous elections), and
likely will have voted for him in the past.

The D+18 Partisan Voter Index (PVI) will make the winner of the Democratic
primary the prohibitive favorite in November's general election against
unenrolled candidate Brendan Phair of Pittsfield.

## Hampden, Hampshire and Worcester

```{r hhw-setup}
ma_sen_4 <- ma_sen %>%
    filter(SENDISTNUM == 4)

ma_towns_pcts_4 <- town_precinct_mix(
                        ma_sen_4,
                        ma_towns_geom,
                        ma_vtds_geom,
                        ma_city_town_num_pcts)

ma_house_hampd_7 <- ma_house_12 %>%
    filter(DIST_CODE == "HAMPD07")

ma_sen_cands_4 <- candidate_locations(sen_cand, 4)
```

```{r hhw-map, fig.cap="*Hampden, Hampshire and Worcester*m"}
ma_sen_4_map <- district_map(ma_sen_4, ma_towns_pcts_4, ma_sen_cands_4, ma_house_hampd_7)

tmap_leaflet(ma_sen_4_map)
```

```{r hhw-cand}
candidate_table(sen_cand, 4)
```

The new Hampden, Hampshire and Worcester district is closely related to the
previous First Hampden and Hampshire district currently held by Eric
Lesser (D-Longmeadow) who is running for Lieutenant Governor. It is the only
state senate seat to feature both a Democratic and Republican primary in 2022.

On the Democratic side, first term State Representative Jacob Oliveira
(D-Ludlow) faces campaign strategist Sydney Levin-Epstein of
Longmeadow. The dashed line on map shows the overlap of the
Oliveira-held Seventh Hampden State Rep district.

Oliveira comes off a hotly contested State Rep win where he defeated
Republican James "Chip" Harrington by 134 votes after a
recount. Perennial candidate Harrington is running for state rep again
against Democrat Aaron Saunders of Belchertown.

Levin-Epstein served as Deputy Finance Director for Ed Markey's 2020
re-election campaign and is bringing her political organizing
experience to bear on her own campaign this year.

The two GOP candidates are John Harding of East Longmeadow, a former
military office, and William Johnson, a business owner from Granby.

## First Worcester

```{r first-worc-setup}
ma_sen_9 <- ma_sen %>%
    filter(SENDISTNUM == 9)

ma_towns_pcts_9 <- town_precinct_mix(
                        ma_sen_9,
                        ma_towns_geom,
                        ma_vtds_geom,
                        ma_city_town_num_pcts)

ma_sen_cands_9 <- candidate_locations(sen_cand, 9)
```

```{r first-worc-map, fig.cap="*Hampden, Hampshire and Worcester*m"}
ma_sen_9_map <- district_map(ma_sen_9, ma_towns_pcts_9, ma_sen_cands_9)

tmap_leaflet(ma_sen_9_map)
```

```{r first-worc-cand}
candidate_table(sen_cand, 9)
```

Senator Harriet Chandler (D-Worcester) is retiring after a 31-year
career in public service. There are two Democrats vying for the
Democratic nomination in the First Worcester district, comprised of
much of Worcester, Boylston, West Boylston, Northborough, Berlin,
and Bolton.

Robyn Kennedy of Worcester is the Executive Director or the YWCA of
Central Massachusetts and has the backing of former Governor Deval
Patrick and his Worcester-based Lt. Governor Tim Murray.

Joseph Petty has been serving as the mayor of Worcester since 2012 and
has gained the endorsement of Senator Ed Markey.

Despite the Markey endorsement for Petty, Kennedy has been getting
support from many of the more progressive groups. Petty has been
running on his experience, presenting an establishment candidate
message.

Petty started with $48,859 in his mayoral campaign account when he
changed its purpose to the State Senate seat in February, 2022, giving
him a big head start over Kennedy who was starting from
scratch. Kennedy has since raised approximately two thirds as much as
Petty, spent about half, and had about 40% of Petty's cash on hand as
of the end of July.

The winner of the Democratic primary will face undeclared candidate
Lisa Mair of Berlin who is running on a platform of informed
medical consent and lower taxes. The Democrat will have a strong
advantage in this D+17 district.

## First Essex

```{r first-essex-setup}
ma_sen_19 <- ma_sen %>%
    filter(SENDISTNUM == 19)

ma_towns_pcts_19 <- town_precinct_mix(
                        ma_sen_19,
                        ma_towns_geom,
                        ma_vtds_geom,
                        ma_city_town_num_pcts)

ma_sen_cands_19 <- candidate_locations(sen_cand, 19)
```

```{r first-essex-map, fig.cap="*First Essex Senate District*m"}
ma_sen_19_map <- district_map(ma_sen_19, ma_towns_pcts_19, ma_sen_cands_19)
tmap_leaflet(ma_sen_19_map)
```

```{r first-essex-cands}
candidate_table(sen_cand, 19)
```

As mentioned in our[redistricting
post](https://www.massnumbers.us/posts/2022-07-18-senate-majority-minority/),
the First Essex district is a new majority-Hispanic district
containing all of Lawrence, all of Methuen, and some inner precincts
of Haverhill.  There is no incumbent senator running as Diana DiZoglio
(D-Methuen) is running for auditor.

Pavel Payano is an at-large City Councilor in Lawrence; Doris
Rodriguiez is a tax layer from Lawrence with public and private sector
experience; and Eunice Zeigler is a City Councilor from Methuen. Payano
and Zeigler have been able to raise and spend similar amounts
of campaign funds.

A look at the total number of votes from each of the cities in
the precincts contained in the First Essex District show that
over 50% of the votes come from Lawrence, while 26% come from
Methuen, and 23% from Haverhill, giving the Lawrence candidates
somewhat of an advantage.

```{r first-essex-votes-per-city}
votes_per_city <- st_intersection(ma_sen_19, ma_vtds_geom) %>%
    filter(st_area(geometry) > one_m_sq) %>%
    st_drop_geometry() %>%
    left_join(read_csv("../../_data/2020/ma_precincts_districts_16_20_pres.csv"),
              by=c("city_town", "Ward", "Pct")) %>%
    mutate(pres_votes = sum(Biden_20, Trump_20, Clinton_16, Trump_16)) %>%
    group_by(city_town) %>%
    summarize(votes = sum(pres_votes)) %>%
    mutate(percent = percent(votes/sum(votes), 0))

votes_per_city %>%
    kbl(caption="First Essex vote distribution 2016/2020 presidential",
        format.args=list(scientific=FALSE, big.mark=",")) %>%
    kable_styling()
```

## Second Essex

```{r second-essex-setup}
ma_sen_22 <- ma_sen %>%
    filter(SENDISTNUM == 22)

ma_towns_pcts_22 <- town_precinct_mix(
                        ma_sen_22,
                        ma_towns_geom,
                        ma_vtds_geom,
                        ma_city_town_num_pcts)

ma_sen_cands_22 <- candidate_locations(sen_cand, 22)
```

```{r second-essex-map, fig.cap="*Second Essex State Senate District*m"}
ma_sen_22_map <- district_map(ma_sen_22, ma_towns_pcts_22, ma_sen_cands_22)

tmap_leaflet(ma_sen_22_map)
```

```{r second-essex-cands}
candidate_table(sen_cand, 22)
```

Assistant Majority Leader Joan Lovely (D-Salem) faces a primary
challenge from Kyle Davis of Danvers. Davis has pursued a career as an
artist and performer, supporting himself with day jobs as a waiter and
food delivery driver, and is running in a attempt to diversify the
legislature with people from atypical political backgrounds.

Senator Lovely's fundraising isn't particularly strong for an
incumbent State Senator, but challenger Davis's numbers are weaker
than they appear based on totals. The $27-28,000 raised and spent
numbers seem reasonable until you look at the particulars. There was a
mistaken transfer into the account of $15,000 (since transfered out)
and a mistaken personal expense written as a check from the account
for $2,483 (since returned), combined with candidate loans to the
campaign of $2,400 and $1,500 which brings the actual amount raised
from non-candidate sources to $6,185.

The winner for the Democratic primary will face Republican Damian
Anketell who seems to have previously run for Essex County Sheriff as
a Democrat. Anketell has been featured at Mass. GOP events promoted by
Howie Carr to get a referendum question on the ballot to deny
drivers licenses for undocumented immigrants. Anketell has made
the extraordinary move of filling out Office of Campaign and
Political Finance (OCPF) from CPF D104 which states that he will
not raise or expend campaign funds in this race.

The D+12 district will likely go to the Democrat, regardless of
primary outcome, but the incumbency advantage of Lovely would make it
highly unlikely for a GOP pickup, especially with the promise of
no money raised or spent by the GOP candidate.

## Second Suffolk

```{r second-suffolk-setup}
ma_sen_29 <- ma_sen %>%
    filter(SENDISTNUM == 29)

ma_towns_pcts_29 <- town_precinct_mix(
                        ma_sen_29,
                        ma_towns_geom,
                        ma_vtds_geom,
                        ma_city_town_num_pcts)

ma_house_5th_suffolk <- ma_house_12 %>%
    filter(DIST_CODE == "SUFF05")

ma_house_15th_suffolk <- ma_house_12 %>%
    filter(DIST_CODE == "SUFF15")

ma_sen_cands_29 <- candidate_locations(sen_cand, 29)
```

```{r second-suffolk-map, fig.cap="*Second Suffolk State Senate District*m"}
ma_sen_29_map <- district_map(ma_sen_29, ma_towns_pcts_29, ma_sen_cands_29, ma_house_5th_suffolk, ma_house_15th_suffolk)

tmap_leaflet(ma_sen_29_map)
```

```{r second-suffolk-cands}
candidate_table(sen_cand, 29)
```

Senator Sonia Chang-Diaz (D-Boston) decided to run for Massachusetts
Governor, rather than for re-election to the State Senate, leaving the
true blue Second Suffolk District as an open seat. There are five
candidates vying for the Democratic nomination, although James Grant
does not seem to pursuing a strong campaign, raising money, or
participating in debates. The primary winner will take office in
January, given that no non-Democratic candidates qualified for the
race, not to mention its status as the most Democratic Massachusetts
state senate district with a whopping PVI of D+42.

There are two sitting State Representatives in the race for the Second
Suffolk.  Nika Elguardo (D-Jamaica Plain) represents the 15th Suffolk
State Representative District (dotted outline in the map), and Liz
Miranda (D-Boston) represents the 5th Suffolk State Representative
District (dashed outline in the map). Neither of these state
representative districts overlaps with Second Suffolk State Senate
district to a large extent.

Miniard Culpepper is a Baptist minister, lawyer, community activist,
and former regional counsel for the Department of Housing and Urban
Development (HUD).

Diane Wilkerson is a former State Senator who plead guitly to
extortion charges and spent two and a half years in prison. Wilkerson
is attempting to make a political comeback, counting on voters to
forgive the transgressions for which she has paid a steep price.

Culpepper, Elugardo, and Miranda have all raised serious amounts of
campaign money. Miranda wins the fundraising and spending titles so
far, having raised $172,074 and spent $188,753. Elugardo has raised
$102,324 and spent $111,936 this year. Culpepper raised more than
Elugardo in the same period, $157,273, but has only spent $45,935. It
is not quite to the point as of July 31 filing that having Culpepper's
$111,338 left in the bank is a liability, but the closer it gets to
primary day, the less useful unspent dollars become.

## First Plymouth and Norfolk

```{r fpn-setup}
ma_sen_31 <- ma_sen %>%
    filter(SENDISTNUM == 31)

ma_towns_pcts_31 <- town_precinct_mix(
                        ma_sen_31,
                        ma_towns_geom,
                        ma_vtds_geom,
                        ma_city_town_num_pcts)

ma_sen_cands_31 <- candidate_locations(sen_cand, 31)
```

```{r fpn-map, fig.cap="*First Plymouth and Norfolk District*"}
ma_sen_31_map <- district_map(ma_sen_31, ma_towns_pcts_31, ma_sen_cands_31)

tmap_leaflet(ma_sen_31_map)
```

```{r fpn-cands}
candidate_table(sen_cand, 31)
```

Senator Patrick O'Connor (D-Weymouth) is the only incumbent GOP
senator to face electoral opposition in 2022. O'Connor is widely
considered a moderate Republican in the Massachusetts vein. He did
[vote against the final version of the MA ROE
Act](https://www.patriotledger.com/story/news/2022/05/03/quincy-residents-protest-draft-opinion-would-overturn-roe-v-wade/9627424002/),
citing issues of expanding third trimester abortions and removal of
most parental consent, but is a proponent of clean energy and
universal full-day kindergarten.

Senator O'Connor faces Marshfield Republican Ronald Patuto in the GOP
primary. There isn't a great deal of campaign information about Patuto
on the internet, but [a radio forum on
WATD](https://959watd.com/blog/2022/08/audio-from-watd-political-forum-for-state-senate-candidates-in-the-plymouth-and-norfolk-district/)
has Patuto arguing for less regulation of the lobstering business
where he gets his livelihood, and against public school mask mandates.

While I am not publishing model-based predictions for the primaries,
this race is one where the general election model shows a big
difference between the probability of a GOP general election win if
O'Connor, the incumbent wins the primary (93% likelihood of GOP win),
and if Patuto, the challenger wins the primary (21% likelihood of GOP
win). This contrast demonstrates the power of incumbency and the fact
that this moderate-for-MA state senate district (it is the 13th most
Republican-leaning State Senate District) has as a Partisan Voter
Index (PVI) of D+9, making it an easy-enough Democratic target without
a moderate incumbent Republican on the ballot.

However, the Democratic candidate for the First Plymouth and Norfolk
District, Robert Stephens Jr. Hanson, is nowhere to be found on the
internet and has not raised any campaign funds as of July 31.

## Norfolk, Plymouth and Bristol

```{r npb-setup}
ma_sen_33 <- ma_sen %>%
    filter(SENDISTNUM == 33)

ma_towns_pcts_33 <- town_precinct_mix(
                        ma_sen_33,
                        ma_towns_geom,
                        ma_vtds_geom,
                        ma_city_town_num_pcts)

ma_sen_cands_33 <- candidate_locations(sen_cand, 33)
```

```{r npb-map, fig.cap="*Norfolk, Plymouth and Bristol District*"}
ma_sen_33_map <- district_map(ma_sen_33, ma_towns_pcts_33, ma_sen_cands_33)

tmap_leaflet(ma_sen_33_map)
```

```{r npb-cands}
candidate_table(sen_cand, 33)
```

State Senator Walter Timilty (D-Milton) is considered on the
conservative side of the Massachusetts Democratic Party. One recent
example would be his "No" vote on the MA ROE Act to expand and codify
the right to an abortion in Massachusetts.

Timilty's challenger for the Democratic nomination is Kathleen
Crogan-Camara who is running on a platform promoting disability
rights, but is also challenging Timilty with a much more progressive
view on issues like women's reproductive rights, LGBTQ+ rights, and
immigrant rights. Crogan-Camara has gotten the endorsement of the
Planned Parenthood Action Fund and has raised enough money
to run a creditable campaign.

The winner of the Democratic primary will face another perennial
candidate, Republican Brian Muello of Braintree who has run
unsuccessfully for Mayor of Braintree, Braintree selectman, Braintree
Town Clerk, and U.S. Senate.

## Second Plymouth and Norfolk

```{r spn-setup}
ma_sen_34 <- ma_sen %>%
    filter(SENDISTNUM == 34)

ma_towns_pcts_34 <- town_precinct_mix(
                        ma_sen_34,
                        ma_towns_geom,
                        ma_vtds_geom,
                        ma_city_town_num_pcts)

ma_sen_cands_34 <- candidate_locations(sen_cand, 34)
```

```{r spn-map, fig.cap="*Second Plymouth and Norfolk*"}
ma_sen_34_map <- district_map(ma_sen_34, ma_towns_pcts_34, ma_sen_cands_34)

tmap_leaflet(ma_sen_34_map)
```

```{r spn-cands}
candidate_table(sen_cand, 34)
```

Another incumbent State Senator facing both a Democratic primary and a
Republican general election challenger is Michael Brady
(D-Brockton). As mentioned in [our redistricting
post](https://www.massnumbers.us/posts/2022-07-18-senate-majority-minority/),
the Second Plymouth and Norfolk district is also newly minted majority
minority district with 49% White, 35% Black, 8% Hispanic, and 3% Asian
in terms of voting age population. However, the black citizen voting
age population is only 28% of the electorate, making this a difficult
district for a black voting block to elect its candidate of choice.

Senator Brady has weathered some challenges over the past several
years, having been arrested for drunk driving in 2018 and being
stripped of his committee chairmanship. Brady has since been
re-instated to his leadership positions, presumably after serving out
his probation with a dismissal of charges as per agreement.

Challenger Katrina Huff-Larmond, a Randolph Town Councilor, is
campaigning on change, leadership, a deeper connection with the
people. In addition, Huff-Larmond participated in the redistricting
process and created a more diverse senate district in the process.
Huff-Larmond has raised $17,314 for the race so far, not
insignificant, but still at a disadvantage to Brady's larger and
longer-term campaign funds.

The winner of the Democratic primary will face Republican Jim Gordon
of Hanson who is running on a platform for parental review of
educational material, restoring qualified immunity for law
enforcement, and cutting taxes while increasing infrastructure
spending. The simple model I use for estimating general
election results based on PVI, incumbency, and whether it is a
presidential voting year gives Gordon a 2 out of 100 chance of winning
against Brady and a 1 out of 10 chance at beating Huff-Larmond.

## Cape and Islands

```{r cape-islands-setup}
ma_sen_40 <- ma_sen %>%
    filter(SENDISTNUM == 40)

ma_towns_pcts_40 <- town_precinct_mix(
                        ma_sen_40,
                        ma_towns_geom,
                        ma_vtds_geom,
                        ma_city_town_num_pcts)

ma_sen_cands_40 <- candidate_locations(sen_cand, 40)
```

```{r cape-islands-map, fig.cap="*Second Plymouth and Norfolk*"}
ma_sen_40_map <- district_map(ma_sen_40, ma_towns_pcts_40, ma_sen_cands_40)

tmap_leaflet(ma_sen_40_map)
```

```{r cape-islands-cands}
candidate_table(sen_cand, 40)
```

State Senator Julian Cyr is free from Democratic primary challengers
in 2022, but will face the winner of a two-way Republican primary in
November. It is quite unlikely that the Republican nominee will unseat
incumbent Cyr in the D+12 Cape and Islands District.

Daralyn Heywood is a private investigator and former State Police
lieutenant running on a anti-illegal immigrant and pro-America platform.

Christopher Lauzon is a former cruise deck-hand who now works in
a family-owned auto repair business in Hyanis. Lauzon is running
on a platform of lowering taxes, reducing regulation, and
supporting and defending law enforcement.

Both GOP candidates have raised at the lower end of the spectrum
for credible state senate challenger campaign funds, and are
nowhere close to Senator Cyr's ability to raise, spend, and
save campaign cash.

