---
title: "An overview of the 2022 State Representative Primaries"
description: |
   A look at the MA State Representative contested primaries
author: Brent Benson
date: 2022-08-31
citation: false
preview: https://www.massnumbers.us/images/first_middlesex_candidates.png
output:
  distill::distill_article:
    self_contained: false
draft: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)

library(metathis)
library(kableExtra)
library(tidyverse)
library(tigris)
library(readxl)
library(forcats)
library(tmap)
library(formattable)
library(sf)
library(kableExtra)
library(nominatimlite)
```

```{r post-meta}
meta() %>%
    meta_description("A look at the MA State Representative contested primaries") %>%
    meta_viewport() %>%
    meta_social(
        title = "An overview of the 2022 State Representative Primaries",
        url = "https://www.massnumbers.us/posts/2022-08-31-state-rep-primaries/",
        image = "https://www.massnumbers.us/images/first_middlesex_candidates.png",
        image_alt = "The precincts and candidates for the 1st Middlesex State Rep race",
        og_type = "article",
        og_author = "Brent Benson",
        twitter_card_type = "summary_large_image",
        twitter_creator = "@bwbensonjr"
    )
```

```{r support-code}
ocpf_location <- read_csv("../../_data/2020/ma_house_ocpf_addr_aug_2022.csv",
                          col_types=cols(cpf_id = "c")) %>%
    select(cpf_id, lat, lon)

house_cands <- read_excel("../../_data/2020/MA_State_Rep_Races_2022.xlsx",
                          sheet="Contested") %>%
    rename(district_id = Code) %>%
    left_join(ocpf_location, by="cpf_id") %>%
    relocate(Candidate)

house_cands_geom <- house_cands %>%
    st_as_sf(coords = c("lon", "lat"), crs="+proj=longlat +datum=WGS84") %>%
    st_transform(6491)

## PVI

Biden_20 <- 81281502
Trump_20 <- 74222593
Clinton_16 <- 65853514
Trump_16 <- 62984828

dem_percent <- function(dem_20, gop_20, dem_16, gop_16) {
    dem_votes <- dem_20 + dem_16
    gop_votes <- gop_20 + gop_16
    dem_pct <- dem_votes/(dem_votes + gop_votes)
    dem_pct
}

pvi_string <- function(pvi_n) {
    case_when(
        pvi_n <= 0.5 ~ str_c("R+", round(abs(pvi_n))),
        pvi_n >= 0.5 ~ str_c("D+", round(pvi_n)),
        TRUE ~ "EVEN"
        )
}

US_PVI_2020 <- dem_percent(Biden_20, Trump_20, Clinton_16, Trump_16)

add_pvi <- function(df) {
    df %>%
        mutate(PVI_N = ((dem_percent(pres_2020_dem,
                                     pres_2020_gop,
                                     pres_2016_dem,
                                     pres_2016_gop) - US_PVI_2020) * 100),
               PVI = pvi_string(PVI_N))
    
}

add_voting_percents <- function(df) {
    df %>%
        mutate(cvap_white_pct = cvap_2019_white/cvap_2019_total,
               cvap_hispanic_pct = cvap_2019_hispanic/cvap_2019_total,
               cvap_black_pct = cvap_2019_black/cvap_2019_total,
               cvap_asian_pct = cvap_2019_asian/cvap_2019_total,
               cvap_pred_group = case_when(((cvap_black_pct >= 0.4) &
                                       (cvap_hispanic_pct >= 0.4)) ~ "Black_and_Hispanic",
                                  (cvap_black_pct >= 0.4) ~ "Black",
                                  (cvap_hispanic_pct >= 0.4) ~ "Hispanic",
                                  (cvap_asian_pct >= 0.4) ~ "Asian",
                                  TRUE ~ "White"),
               vap_white_pct = vap_white_2020/vap_total_2020,
               vap_hispanic_pct = vap_hispanic_2020/vap_total_2020,
               vap_black_pct = vap_black_2020/vap_total_2020,
               vap_asian_pct = vap_asian_2020/vap_total_2020,
               vap_pred_group = case_when(((vap_black_pct >= 0.4) &
                                       (vap_hispanic_pct >= 0.4)) ~ "Black_and_Hispanic",
                                  (vap_black_pct >= 0.4) ~ "Black",
                                  (vap_hispanic_pct >= 0.4) ~ "Hispanic",
                                  (vap_asian_pct >= 0.4) ~ "Asian",
                                  TRUE ~ "White"))
}

ward_precinct_name <- function(city_town, ward, precinct, partial_pct, partial_pct_num, single_pct) {
    if_else(single_pct,
            city_town,
            str_c(city_town,
                  if_else(ward == "-", "", str_c(" Ward ", ward)),
                  str_c(" Precinct ", precinct),
                  if_else(partial_pct, str_c(" (part ", partial_pct_num, ")"), "")))
}

house_dist_blks <- read_csv("../../_data/2020/ma_blocks_to_house_dist_2021.csv",
                            col_types=cols("c", "c"))

house_vtd_demos <- read_csv("../../_data/2020/ma_house_census_2020_daves.csv",
                            col_types=cols(vtd_fips="c"))

vtd_demos <- house_vtd_demos %>%
    select(-c(district_id, partial_pct)) %>%
    group_by(vtd_fips) %>%
    summarize_all(sum)

vtd_names <- read_csv("../../_data/2020/ma_vtds_city_town_fips.csv",
                      col_types=cols(.default="c")) %>%
    select(vtd_fips, city_town, Ward, Pct)

vtd_geom <- voting_districts(state=25, cb=TRUE) %>%
    select(vtd_fips = GEOID20) %>%
    st_transform(6491) %>%
    left_join(vtd_names, by="vtd_fips") %>%
    left_join(vtd_demos, by="vtd_fips")

block_geom <- blocks(state=25, year=2020) %>%
    select(block_fips = GEOID20) %>%
    st_transform(6491)

vtd_blks <- read_csv("../../_data/2020/ma_blocks_to_vtds.csv",
                     col_types=cols(.default="c")) %>%
    left_join(vtd_names, by="vtd_fips") %>%
    inner_join(block_geom, by="block_fips")

blks_house_dist_vtd <- vtd_blks %>%
    left_join(house_dist_blks, by="block_fips")

house_vtds <- blks_house_dist_vtd %>%
    nest(blocks = c(block_fips, geometry)) %>%
    group_by(vtd_fips) %>%
    mutate(partial_pct = (n() > 1),
           partial_pct_num = row_number()) %>%
    ungroup() %>%
    group_by(city_town) %>%
    mutate(single_pct = (n() == 1)) %>%
    ungroup() %>%
    mutate(geometry = map(blocks, ~st_union(.x$geometry))) %>%
    unnest(geometry) %>%
    select(-blocks)

house_vtd_comp <- house_vtds %>%
    left_join(house_vtd_demos,
              by=c("vtd_fips", "district_id", "partial_pct")) %>%
    mutate(name = ward_precinct_name(city_town,
                                     Ward,
                                     Pct,
                                     partial_pct,
                                     partial_pct_num,
                                     single_pct)) %>%
    add_pvi() %>%
    add_voting_percents() %>%
    relocate(name) %>%
    st_as_sf()

## New 2021 House Districts

house_dist_demos <- house_vtd_demos %>%
    select(-c(vtd_fips, partial_pct)) %>%
    group_by(district_id) %>%
    summarize_all(sum) %>%
    add_pvi() %>%
    add_voting_percents() %>%
    select(district_id, PVI, PVI_N,
           cvap_white_pct, cvap_hispanic_pct, cvap_black_pct, cvap_asian_pct, cvap_pred_group,          
           vap_white_pct, vap_hispanic_pct, vap_black_pct, vap_asian_pct, vap_pred_group)

house_dists <- read_excel("../../_data/2020/MA_State_Rep_Races_2022.xlsx",
                          sheet="Races") %>%
    select(district_id = Code,
           District,
           inc_status,
           Incumbent,
           Primaries) %>%
    mutate(inc_status = fct_recode(inc_status,
                                   Dem="Dem_Incumbent",
                                   GOP="GOP_Incumbent",
                                   None="No_Incumbent"),
           Primaries = replace_na(Primaries, "")) %>%
    left_join(house_dist_demos, by="district_id")

house_geom <- read_sf("../../_gis/HOUSE2021/HOUSE2021_POLY.shp") %>%
    st_make_valid() %>%
    st_transform(6491) %>%
    select(district_id = DIST_CODE) %>%
    left_join(house_dists, by="district_id") %>%
    arrange(district_id) %>%
    relocate(District)

## Maps

percent_zero <- function(n) {
    percent(n, 0)
}

district_map <- function(dist_id, has_water=FALSE) {
    dist_cands <- house_cands_geom %>%
        filter(district_id == dist_id)
    
    dist_outline <- house_geom %>%
        filter(district_id == dist_id)

    dist_vtds <- house_vtd_comp %>%
        filter(district_id == dist_id)
    if (has_water) {
        dist_vtds <- dist_vtds %>%
            erase_water(area_threshold=0.6)
    }
    base_map <- (tm_shape(dist_outline) +
                 tm_borders(lwd=3) +
                 tm_shape(dist_vtds) +
                 tm_borders(lwd=1) +
                 tm_fill(col="PVI_N",
                         palette=c("#FF4949", "white", "#0F4392"),
                         midpoint=0,
                         alpha=0.7,
                         popup.vars=c("PVI")))
    if (nrow(dist_cands) > 0) {
        base_map <- (base_map + 
                     tm_shape(dist_cands) +
                     tm_symbols(col="Party",
                                legend.col.show=FALSE,
                                size=0.3,
                                palette=c(D="blue", R="red", U="gray"),
                                popup.vars=c("Party", "city_town", "Incumbent")))
    }
    (base_map + tm_basemap("OpenStreetMap"))
}

district_table <- function(dist_id) {
    cands <- house_cands %>%
        filter(district_id == dist_id)
    
    dist_name <- (cands %>% slice(1) %>% pull(District))
    caption <- str_c(dist_name, " District Candidates")

    cands %>%
        select(Inc = Incumbent,
               Candidate,
               Party,
               city_town,
               receipts,
               spent = expenditures,
               cash = end) %>%
        mutate(receipts = currency(receipts, digits=0),
               spent = currency(spent, digits=0),
               cash = currency(cash, digits=0)) %>%
        kbl(caption=caption) %>%
        kable_styling()
}

tmap_mode(mode="view")
```

```{r new-district-map, layout="l-page", fig.cap="*MA State Representative Districts (2022)*"}
dist_popup_vars <- c(
    "Incumbent",
    "inc_status",
    "PVI",
    "Primaries",
    "Hispanic CVAP"="cvap_hispanic_pct",
    "Black CVAP"="cvap_black_pct",
    "Asian CVAP"="cvap_asian_pct",
    "White CVAP"="cvap_white_pct"
)

dist_popup_format <- list(
    cvap_hispanic_pct=list(fun=percent_zero),
    cvap_black_pct=list(fun=percent_zero),
    cvap_asian_pct=list(fun=percent_zero),
    cvap_white_pct=list(fun=percent_zero)
    )

new_house_map <- (tm_shape(house_geom) +
                    tm_borders(lwd=2) +
                    tm_fill(col="PVI_N",
                            palette=c("#FF4949", "white", "#0F4392"),
                            midpoint=0,
                            alpha=0.6,
                            popup.format=dist_popup_format,
                            popup.vars=dist_popup_vars) +
                  tm_shape(house_cands_geom) +
                  tm_symbols(col="Party",
                             legend.col.show=FALSE,
                             size=0.01,
                             palette=c(D="blue", R="red", U="gray"),
                             popup.vars=c("Party", "city_town")) +
                  tm_basemap("OpenStreetMap"))

tmap_leaflet(new_house_map)
```

## State Represenative Primary Overview

There are 34 State Representative districts with contested primaries
with the State Primary occurring on Sept 6, 2022. There are 19 open
seats with 13 having a Democratic primary, 1 have a GOP primary, 4
with only a general election matchup, and 1 seat&mdash;the 15th
Essex&mdash;with a single candidate.

There are There are 17 Democratic incumbents with primary challenges,
while no Republican House incumbent faces a primary challenge. One
incumbent Democrat, William Straus (D-Mattapoisett) of the 10th Bristol
District, faces a primary challenge and two GOP challengers, only one
of which will survive a Republican primary.

```{r state-rep-districts}
house_dists %>%
    select(c(District,
             PVI,
             `Inc. Party`=inc_status,
              Incumbent,
              Primaries)) %>%
    kbl(caption="2022 State Rep districts (last incumbent)*") %>%
    kable_styling()
```

## Open Seats

There are nineteen open State Representative seats of which
thirteen have Democratic primaries and one has a Republican
primary.

## 4th Essex

```{r essex-04-map, fig.cap="4th Essex District Map"}
essex_04_map <- district_map("ESSEX04")
tmap_leaflet(essex_04_map)
```

```{r essex-04-table}
district_table("ESSEX04")
```

The entirely reconfigured 4th Essex district has a Hispanic citizen
voting age population (CVAP) of 57%. The current 4th Essex
representative, Jamie Belsito (D-Topsfield), as elected in a November
2021 special election to replace Bradford Hill (R-Ipswich) was
redistricted into the 13th Middlesex District represented by
Sally Kerans (D-Danvers). Belsito does not appear to be challenging
Kerans.

Perennial candidate, former Mayor, and former State Representative
William Lantigua is in the running, but has not raised any money and
did not participate in [candidate
forums](http://archive.methuentv.org/Video/7612) and debates.

James McCarty is a Methuen City Councilor who is finishing up his time
on the council due to term limits and is a current law student who plans
to take the bar in July 2023.

Estela Reyes is five term City Councilor from Lawrence and a long-term
presence on the Lawrence political scene.

The 4th Essex district is almost evenly split by population between
Methuen and Lawrence, but 58% of the citizen age voting population
comes from Methuen vs. 42% for Lawrence. There may end up being
some level of splitting caused by the Lantigua candidacy, reducing
the power of the Lawrence and hispanic voting populations.

## 7th Essex

```{r essex-07-map, fig.cap="7th Essex District Map"}
essex_07_map <- district_map("ESSEX07", has_water=TRUE)
tmap_leaflet(essex_07_map)
```

```{r essex-07-table}
district_table("ESSEX07")
```

The 7th Essex District is one of the four State Representative
districts that is comprised of a single, complete
municipality&mdash;in this case Salem. State Representative
Paul Tucker (D-Salem) chose to run for Essex County District
Attorney, leaving the seat open.

Manny Cruz is a former Tucker aide and current Vice Chair of the Salem
Public School District School Committee. Cruz has received some high
profile endorsements, including that of Salme Mayor Kim Driscoll, who
is also a candidate for Lieutenant Governor.

Domingo Dominguez is an at-large Salem City Councilor and was a
[backer of Mayor Driscoll's
challenger](https://www.salemnews.com/news/cruz-formally-launches-rep-campaign-dominguez-joins-race/article_c5871008-7861-11ec-8249-dffa552817fa.html)
in the previous mayoral election.

Candidate Gene Collins has not raised any money for the campaign and
there is [little
information](https://www.youtube.com/watch?v=T5VF73RE3io) about
Collins or his platform in the news.

## 8th Essex

```{r essex-08-map, fig.cap="8th Essex District Map"}
essex_08_map <- district_map("ESSEX08", has_water=TRUE)
tmap_leaflet(essex_08_map)
```

```{r essex-08-table}
district_table("ESSEX08")
```

There is a [crowded
field](https://www.salemnews.com/news/5-enter-race-for-8th-essex-seat/article_3e245226-a62c-11ec-b19e-0fd3506550bc.html)
vying to replace State Representative Lori Ehrlich (D-Marblehead) of
who is leaving her 8th Essex seat to become Region 1 Administrator for
the Federal Emergency Management Agency (FEMA).

There are six Democratic candidates, three from Marblehead and three
from Swampscott. The district also contains two precincts from Lynn.

The [major story of the
campaign](https://www.salemnews.com/election/battle-lines-drawn-in-8th-essex-rep-race-after-endorsements/article_c8957178-27b0-11ed-b8fe-6bf9a08733e4.html)
has become the endorsement of political neophyte Tristan Smith by
Massachusetts Senator Ed Markey, who has a long-standing relationship
with Smith's father James Smith.

There has also been a debate about candidates like Douglas Thompson
who have significantly self-funded their campaigns, rather than
being forced to compete with their ideas and credentials for
more and smaller individual donations.

## 8th Hampden

```{r hampde-08-map, fig.cap="8th Hampden District Map"}
hampden_08_map <- district_map("HAMPDEN08")
tmap_leaflet(hampden_08_map)
```

```{r hampden-08-table}
district_table("HAMPDEN08")
```

State Representative Joe Wagner (D-Chicopee) is retiring after having
represented the 8th Hampden District for 30 years. The 8th consists of
a little over three quarters of Chicopee, by population.

The candidates for the Democratic nomination are Ward 1 City Councilor
Joel McAuliffe and At Large City Councilor Shirley Arriaga.

Arriaga is a public school teacher at Chicopee High School and is an
Air Force veteran, and has received the endorsement of Attorney General
and Governor candidate Maura Healey.

McAuliffe has served as an aide to Mayor Richard Kos, and then State
Senator Eric Lesser and has received the endorsement of Lesser.

## 1st Middlesex

```{r middle-01-map, fig.cap="1st Middlesex District Map"}
middle_01_map <- district_map("MIDDLE01")
tmap_leaflet(middle_01_map)
```

```{r middle-01-table}
district_table("MIDDLE01")
```

The 1st Middlesex has been represented by State Representative Sheila
Harrington (R-Groton) who has been appointed as Clerk Magistrate of
the Gardner District Court. The district has been reconfigured to take
three precincts of Lunenburg from the 37th Middlesex District and
should be one of the most [competitive general election
races](https://www.bostonglobe.com/2022/07/22/metro/little-red-pocket-new-hampshire-border-national-fissures-splitting-competitive-mass-state-rep-race/?event=event12)
in November with a Partisan Voter Index (PVI) of D+2.

The only primary is for the Republican nomination which features local
farmer Andrew Sullivan of Townsend and Lynn Archambault of
Pepperell. Sullivan has attempted to portray himself as a moderate
conservative, while Archambault hews more to Trump-style politics.

The winner of the GOP primary will face Democrat Margarate Scarsdale
of Pepperell and undeclared candidate Catherine Lundeen.

## 6th Middlesex

```{r middle-06-map, fig.cap="6th Middlesex District Map"}
middle_06_map <- district_map("MIDDLE06")
tmap_leaflet(middle_06_map)
```

```{r middle-06-table}
district_table("MIDDLE06")
```

The 6th Middlesex District was redistricted without attention
to incumbent Maria Robinson (D-Framingham) who has been in 
the process of taking a job with the U.S. Department of Energy.

There are three candidates for the Democratic nomination: former
Framingham City Councilor Margareth Shepard, Framingham School
Committee member Priscila Sousa, and local activist and Former
Framingham mayoral candidate Dhruba Sen.

## 14th Middlesex

```{r middle-14-map, fig.cap="14th Middlesex District Map"}
middle_14_map <- district_map("MIDDLE14")
tmap_leaflet(middle_14_map)
```

```{r middle-14-table}
district_table("MIDDLE14")
```

The current 14th Middlesex State Representative Tami Gouveia (D-Acton)
is running for Lieutenant Governor leaving the seat open. There
Democrats [vying for the Democratic
nomination](https://www.lowellsun.com/2022/09/03/democratic-candidates-vie-for-14th-middlesex-seat/).

Vivian Birchall is a TV producer from Acton and has secured the
endorsements of State Senator Jamie Eldridge (D-Acton) and State
Representative Daniel Sena (D-Acton). Prosecutor Simon Cataldo of
Concord is running largely on a reproductive rights platform and has
the endorsement of the Planned Parenthood Advocy Fund of
Massachusetts. Chelmsford Select Board Member Pat Wojtas has been
running on her municipal experience and is advocating for a change
to the states funding and local aide models.

## Incumbent Challenges

There are 17 incumbent House Democrats facing an intra-party primary
challenge. There are no Republican incumbents facing a GOP primary
challenge.

## 4th Barnstable

```{r barn-04-map, fig.cap="4th Barnstable District Map"}
barn_04_map <- district_map("BARN04")
tmap_leaflet(barn_04_map)
```

```{r barn-04-table}
district_table("BARN04")
```

Eight term incumbent State Representative Sarah Peake (D-Provincetown)
[faces
challenger](https://www.wickedlocal.com/story/cape-codder/2022/08/17/jack-stanton-and-sarah-peake-face-off-democratic-primary/10305447002/)
Jack Stanton, also of Provincetown. Stanton previously ran against
Republican State Representative Randy Hunt in the 5th Barnstable
district when he lived in Sandwhich. Peake, the Second Assistant
Majority Leader in the House is stressing the need for more affordable
housing, while Stanton is campaigning for increased transparency in
the legislature with term limits for Speaker of the House.

## 1st Berkshire

```{r berk-01-map, fig.cap="1st Berkshire District Map"}
berk_01_map <- district_map("BERK01")
tmap_leaflet(berk_01_map)
```

```{r berk-01-table}
district_table("BERK01")
```

State Representative John Barrett (D-North Adams) [faces a primary
challenge](https://www.wamc.org/news/2022-08-25/1st-berkshire-district-challenger-kingsbury-evans-says-state-rep-barrett-mispresented-beacon-hill-transparency-at-north-adams-forum)
from Paula Kingsbury-Evans who is a recent graduate of the
Massachusetts College of Liberal Arts, also from North Adams. Barrett
is the former Mayor of North Adams and was first elected in a 2017
special election after the death of Representative Gailanne Cariddi.
Kingsbury-Evans is also running on the issue of transparency in the
legislature.

## 9th Bristol

```{r bristol-09-map, fig.cap="9th Bristol District Map"}
bristol_09_map <- district_map("BRISTOL09", has_water=TRUE)
tmap_leaflet(bristol_09_map)
```

```{r bristol-09-table}
district_table("BRISTOL09")
```

State Representative Christopher Markey (D-Dartmouth) [faces a
challenge](https://dartmouth.theweektoday.com/article/state-rep-markey-face-first-primary-challenger-2014/58020)
from New Bedford resident Cameron Costa, a 20 year old recent
graduate of University of Massachusetts Dartmouth. Markey is
stressing his experience, while Costa says that change is needed
and is arguing for increased education funding.

## 5th Essex

```{r essex-05-map, fig.cap="5th Essex District Map"}
essex_05_map <- district_map("ESSEX05")
tmap_leaflet(essex_05_map)
```

```{r essex-05-table}
district_table("ESSEX05")
```

State Representative Ann-Margaret Ferrante (D-Gloucester) [faces a
primary
challenge](https://www.gloucestertimes.com/election/rockporter-to-challenge-ferrante-in-primary/article_03639896-d623-11ec-bea9-ef1b0c302033.html)
from D. Nathaniel Mulcahy of Rockport. Ferrante is Vice Chair of House
Committee on Ways and Means, the important committee that determines
spending. Mulcahy is a former teacher and is the Chair of the Rockport
Democratic Town Committee. Mulcahy is campaigning for increased
transparency in the legislature, a theme among Democratic challengers.

## 16th Essex

```{r essex-16-map, fig.cap="16th Essex District Map"}
essex_16_map <- district_map("ESSEX16")
tmap_leaflet(essex_16_map)
```

```{r essex-16-table}
district_table("ESSEX16")
```

State Representative Marcos Devers (D-Lawrence) faces a [primary
challenge](https://www.eagletribune.com/news/new_hampshire/16th-essex-state-representative-candidates-voice-views/article_40dcc364-0927-11ed-a77e-9f01431bcb43.html)
from Francisco Paulino of Methuen. Devers has been an educator in the
region for thirty years. Paulino is a 22 year old lawyer and tax
associate who advises the City of Lawrence on economic development
and has served on the Lawrence School Committee.

## 11th Hampden

```{r hampden-11-map, fig.cap="11th Hampden District Map"}
hampden_11_map <- district_map("HAMPDEN11")
tmap_leaflet(hampden_11_map)
```

```{r hampden-11-table}
district_table("HAMPDEN11")
```

State Representative Bud Williams (D-Springfield) faces a challenge
from Jynai McDonald, also of Springfield. Williams has served as
State Representative of the 11th Hampden District since 2016 and
before that was a fixture on the Springfield City Council. McDonald 
is a local activist who has not held political office and wants
to expand investments in the community.

## 9th Middlesex

```{r middle-09-map, fig.cap="9th Middlesex District Map"}
middle_09_map <- district_map("MIDDLE09")
tmap_leaflet(middle_09_map)
```

```{r middle-09-table}
district_table("MIDDLE09")
```

State Representative Tom Stanley (D-Waltham) faces a [primary
challenge](https://patch.com/massachusetts/waltham/heather-may-run-waltham-state-representative)
from Heather May of Waltham. May is Chair of Waltham Demoractic Town
Committee, a volunteer, and educator at Emerson College. May is
running as an outsider and is calling for transparency and
accountability. Stanley is running on his office's constituent
services and successful passing of clean energy and economic
development legislation.

## 18th Middlesex

```{r middle-18-map, fig.cap="18th Middlesex District Map"}
middle_18_map <- district_map("MIDDLE18")
tmap_leaflet(middle_18_map)
```

```{r middle-18-table}
district_table("MIDDLE18")
```

State Representative Rady Mom (D-Lowell) faces two challengers for the
Democratic nomination, recent UMass Lowell graduate Tara Hong and Lowell
School Committee member Dominik Lay. When Mom was first elected to
the legislature in 2014, he was the first Cambodian-American lawmaker in
the country. His two challengers are also Cambodian born. Mom is running
on his strong legislative record. Hong portrays himself as a progressive
reformer, and Lay stresses independenc.

## 21st Middlesex

```{r middle-21-map, fig.cap="21st Middlesex District Map"}
middle_21_map <- district_map("MIDDLE21")
tmap_leaflet(middle_21_map)
```

```{r middle-21-table}
district_table("MIDDLE21")
```

State Representative Kenneth Gordon (D-Bedford) faces a [primary
challenge](https://www.thebedfordcitizen.org/2022/06/burlington-resident-24-challenges-rep-ken-gordon-in-primary/)
from 24 year old Timothy Sullivan of Burlington. Sullivan is
running on a platform of increasing affordability of public
college education. Gordon has received endorsements from
Congresswoman Catherine Clark and Senator Ed Markey.

## 27th Middlesex

```{r middle-27-map, fig.cap="27th Middlesex District Map"}
middle_27_map <- district_map("MIDDLE27")
tmap_leaflet(middle_27_map)
```

```{r middle-27-table}
district_table("MIDDLE27")
```

First term State Representative Erika Uyterhoeven (D-Somerville) faces
a [primary
challenge](http://www.thesomervilletimes.com/archives/117222) from
entrepreneur Jason Mackey, also of Somerville in the incredibly
progressive D+39 27th Middlesex District. Uyterhoeven ran on an
unapologetically Democratic Socialist platform in 2020 and the
differences with Mackey are not large. Mackey stresses his personal
story and work for social justice.

## 28th Middlesex

```{r middle-28-map, fig.cap="28th Middlesex District Map"}
middle_28_map <- district_map("MIDDLE28")
tmap_leaflet(middle_28_map)
```

```{r middle-28-table}
district_table("MIDDLE28")
```

State Representative Joseph McGonagle (D-Everett) faces a [primary
challenge](https://everettleader.com/2022/07/01/guerline-alcy-is-running-for-state-representative-against-rep-joe-mcgonagle/)
from Everett city employee, financial professional, Haitian activist,
and community organizer Guerline Alcy.

## 35th Middlesex

```{r middle-35-map, fig.cap="35th Middlesex District Map"}
middle_35_map <- district_map("MIDDLE35")
tmap_leaflet(middle_35_map)
```

```{r middle-35-table}
district_table("MIDDLE35")
```

State Representative Paul Denato (D-Medford) faces a [primary
challenge](https://patch.com/massachusetts/malden/nichole-mossalam-enters-state-rep-race-malden-medford)
from Nichole Mossalam, founding director of The Islamic Cultural Center of Medford
and former chair of Malden's Disability Commission.

## 1st Norfolk

```{r norfolk-01-map, fig.cap="1st Norfolk District Map"}
norfolk_01_map <- district_map("NORFOLK01", has_water=TRUE)
tmap_leaflet(norfolk_01_map)
```

```{r norfolk-01-table}
district_table("NORFOLK01")
```

State Representative Bruce Ayers (D-North Quincy) faces a [primary
challenge](https://www.politico.com/newsletters/massachusetts-playbook/2022/04/27/dont-call-it-a-comeback-or-do-00028116)
from small business owner Casey Dooley.

## 6th Norfolk

```{r norfolk-06-map, fig.cap="6th Norfolk District Map"}
norfolk_06_map <- district_map("NORFOLK06")
tmap_leaflet(norfolk_06_map)
```

```{r norfolk-06-table}
district_table("NORFOLK06")
```

State Representative William Galvin (D-Canton) faces a [primary
challenge](https://www.tamishacivil.com/about) from Tamisha Civil of
Stoughton who is a probation officer and previously worked in the
Suffolk County District Attorney's office.

## 15th Norfolk

```{r norfolk-15-map, fig.cap="15th Norfolk District Map"}
norfolk_15_map <- district_map("NORFOLK15")
tmap_leaflet(norfolk_15_map)
```

```{r norfolk-15-table}
district_table("NORFOLK15")
```

The challange to State Representative Tommy Vitolo (D-Brookline) from
Brookline Select Board Member Raul Fernandez has generated a [great
deal of
attention](https://commonwealthmagazine.org/politics/a-liberal-showdown-in-brookline-rep-race/)
due to the number of high-profile endorsements on both
sides. Governor Michael Dukakis is backing Vitolo, while
Sonia Chang-Diaz and Joe Kennedy III are backing Fernandez.

## 6th Suffolk

```{r suffolk-06-map, fig.cap="6th Suffolk District Map"}
suffolk_06_map <- district_map("SUFFOLK06")
tmap_leaflet(suffolk_06_map)
```

```{r suffolk-06-table}
district_table("SUFFOLK06")
```

State Representative Russell Holmes (D-Mattapan) faces a [primary
challenge](https://www.dotnews.com/2022/sixth-suffolk-house-match-it-s-hardaway-v-holmes)
from former elder services worker and consultant Haris Hardaway of
Dorchester.

## 17th Worcester

```{r wor-17-map, fig.cap="17th Worcester District Map"}
wor_17_map <- district_map("WOR17")
tmap_leaflet(wor_17_map)
```

```{r wor-17-table}
district_table("WOR17")
```

State Representative David LeBoeuf (D-Worcester) faces a [primary
challenge](https://www.masslive.com/worcester/2022/05/worcester-rep-david-leboeuf-could-face-primary-challenger-following-drunken-driving-arrest-allegedly-blowing-4x-legal-limit.html?fr=operanews)
from former Worcester Library Trustee Michelle Keane. LeBoeuf has been
dealing with the after effects of a drunk driving charge and treatment
for alcoholism. Keane has somewhat downplayed the drunk driving charge
while making the case that she would be better able to affect change
for the district.

