---
title: "Has your city or town gotten more Republican or Democratic over the last 50 years?"
description: |
  A municipality-based look at Massachusetts presidential voting from 1972 through 2020.
author: Brent Benson
date: 2023-01-23
citation: false
preview: https://www.massnumbers.us/images/change_1972_2020_pres_margin_vs_state.png
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)

library(metathis)
library(tidyverse)
library(gt)
library(gtExtras)
library(sf)
library(tmap)
library(formattable)
library(xaringanExtra)
set.seed(999)
tmap_mode(mode="view")
```

```{r post-meta}
meta() %>%
    meta_description("A municipality-based look at Massachusetts presidential voting from 1972 through 2020") %>%
    meta_viewport() %>%
    meta_social(
        title = "Many Massachusetts towns have changed, some have remained the same",
        url = "https://www.massnumbers.us/posts/2023-01-02-town-presidential-voting-history/",
        image = "https://www.massnumbers.us/images/change_1972_2020_pres_margin_vs_state.png",
        image_alt = "Change in 1972 to 2020 Margin vs. State",
        og_type = "article",
        og_author = "Brent Benson",
        twitter_card_type = "summary_large_image",
        twitter_creator = "@bwbensonjr"
    )
```

```{r read-data}
compass_changes <- c(
    "N. " = "North ",
    "E. " = "East ",
    "S. " = "South ",
    "W. " = "West "
)

election_votes <- function(file_name, dem_column, gop_column) {
    read_csv(str_c("../../_data/2023/", file_name)) %>%
        rename_with(~c("city_town", "dem_votes", "gop_votes", "total_votes"),
                    all_of(c("City/Town", dem_column, gop_column, "Total Votes Cast"))) %>%
        select(city_town, dem_votes, gop_votes, total_votes) %>%
        mutate(city_town = str_replace_all(city_town, compass_changes),
               dem_percent = dem_votes/(dem_votes + gop_votes),
               gop_percent = gop_votes/(dem_votes + gop_votes),
               dem_margin = (dem_percent - gop_percent)*100)
}

election_summary <- function(year,
                             dem_candidate,
                             dem_percent,
                             gop_candidate,
                             gop_percent) {
    dem_pct_str <- percent(dem_percent, 0)
    gop_pct_str <- percent(gop_percent, 0)
    if_else((dem_percent > gop_percent),
            str_glue("{year}: {dem_candidate} (D) {dem_pct_str} over ",
                     "{gop_candidate} (R) {gop_pct_str}"),
            str_glue("{year}: {gop_candidate} (R) {gop_pct_str} over ",
                     "{dem_candidate} (D) {dem_pct_str}"))
}

## Year-oriented dataset
pres_elecs <- read_csv("../../_data/2023/ma_pres_elecs_1972_2020.csv") %>%
    mutate(city_town = pmap(list(file_name, dem_column, gop_column),
                            election_votes),
           dem_percent = dem_votes/(dem_votes + gop_votes),
           gop_percent = gop_votes/(dem_votes + gop_votes),
           dem_margin = (dem_percent - gop_percent)*100,
           title = election_summary(year, dem_candidate, dem_percent,
                                    gop_candidate, gop_percent))

## City/Town-oriented dataset
city_town_elecs <- pres_elecs %>%
    select(year, state_margin=dem_margin, city_town) %>%
    unnest(city_town) %>%
    mutate(dem_margin_vs_state = dem_margin - state_margin)

## Margin Table
ct_margin_table <- city_town_elecs %>%
    select(city_town, year, dem_margin) %>%
    pivot_wider(id_cols=city_town,
                names_from=year,
                values_from=dem_margin)
```

>  It's not like Massachusetts, where they're baptized Democrats.

Massachusetts is often seen as a liberal bastion and uniformly
supportive of the Democratic Party as in this quote by pollster Joel
Benenson (he was contrasting Massachusetts with a Democratic-trending
New Jersey) but its history and present behavior is much more
complicated.

In the last 50 years Massachusetts presidential election maps have
appeared as a sea of red cities and towns puncuated with several blue
islands one year (1984), and then blue as far as the eye can see with
only a handful of red towns 12 years later (1996). There have been
several noticeable geographic trends in Massachusetts partisanship
over the last 50 years including increased Democratic dominance in far
Western Massachusetts, a move from Republican to Democratic support in
the rich suburbs of Boston, and stubborn Republican consistency in the
rural parts of Hampden County and much of Worcester, Bristol, and
Plymouth counties.

In 1984 82% of Massachusetts' cities and towns voted for Republican
Ronald Reagan over Walter Mondale and Reagan carried a small majority
of the Commonweath's voters. In 1996, Bill Clinton won a majority of
votes in all but nine Massachusetts municipalities while besting Bob
Dole by over 25 points.

This post looks at the two-party (Dem/GOP) vote margin for every
Massachusetts presidential election from 1972 through 2020 and
identifies municipalities that have been conistently Democratic,
consistently Republican, and towns that have seen political
transformation over the last 50 years.

```{r yearly-map-generation}
city_town_geom <- read_sf("../../_gis/towns2019/TOWNSSURVEY_POLYM_GENCOAST.shp") %>%
    st_make_valid() %>%
    st_simplify(dTolerance=100, preserveTopology=TRUE) %>%
    mutate(city_town = str_replace(str_to_title(TOWN), "Manchester", "Manchester-by-the-Sea")) %>%
    select(city_town)

counties <- read_sf("../../_gis/counties/COUNTIESSURVEY_POLYM_GENCOAST.shp") %>%
    mutate(county_name = str_to_title(COUNTY)) %>%
    select(county_name) %>%
    st_simplify(dTolerance=100, preserveTopology=TRUE) %>%
    st_cast()

year_results <- function(pres_elecs, elec_year) {
    pres_elecs %>%
        filter(year == elec_year) %>%
        rename(state_dem_margin = dem_margin) %>%
        select(-c(election_id,
                  dem_column, gop_column,
                  dem_votes, gop_votes,
                  file_name,
                  dem_percent, gop_percent)) %>%
        unnest(city_town) %>%
        mutate(dem_margin_vs_state = (dem_margin -
                                      state_dem_margin)) %>%
        relocate(city_town, dem_margin)
}

margin_map_for_year <- function(pres_elecs, ct_geom, county_geom, elec_year) {
    ct_elecs <- ct_geom %>%
        left_join(year_results(pres_elecs, elec_year),
                  by="city_town")
    map_title <- ct_elecs$title[1]
    (tm_shape(ct_elecs) +
     tm_fill(col="dem_margin",
             title="Democratic Margin",
             popup.vars=c(
                 "dem_candidate",
                 "gop_candidate",
                 "dem_margin",
                 "state_dem_margin",
                 "dem_margin_vs_state",
                 "dem_votes",
                 "gop_votes"),
             popup.format=list(digits=0),
             alpha=0.7,
             palette=c("#FF4949", "white", "#0F4392"),
             breaks=c(-70,-50,-20,-5,0,10,25,50,90),
             midpoint=0) +
     tm_borders(lwd=1) +
     tm_shape(county_geom) +
     tm_borders(lwd=2, col="black") +
     tm_basemap("OpenStreetMap") +
     # tm_layout(title=map_title) +
     tm_view(set.view=8))
}
```

```{r panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

::: l-page

::::: {.panelset}

::: {.panel}

### 1972 {.panel-name}

```{r map-1972, fig.height=5, layout="l-page"}
margin_map_for_year(pres_elecs, city_town_geom, counties, 1972)
```

George McGovern (D) 55% over Richard Nixon (R) 45%

:::

::: {.panel}

### 1976 {.panel-name}

```{r map-1976, fig.height=5, layout="l-page"}
margin_map_for_year(pres_elecs, city_town_geom, counties, 1976)
```

Jimmy Carter (D) 58% over Gerald Ford (R) 42%

:::

::: {.panel}

### 1980 {.panel-name}

```{r map-1980, fig.height=5, layout="l-page"}
margin_map_for_year(pres_elecs, city_town_geom, counties, 1980)
```

Ronald Reagan (R) 50% over Jimmy Carter (D) 50%

:::

::: {.panel}

### 1984 {.panel-name}

```{r map-1984, fig.height=5, layout="l-page"}
margin_map_for_year(pres_elecs, city_town_geom, counties, 1984)
```

Ronald Reagan (R) 51% over Walter Mondale (D) 49%

:::

::: {.panel}

### 1988 {.panel-name}

```{r map-1988, fig.height=5, layout="l-page"}
margin_map_for_year(pres_elecs, city_town_geom, counties, 1988)
```

Michael Dukakis (D) 54% over George HW Bush (R) 46%

:::

::: {.panel}

### 1992 {.panel-name}

```{r map-1992, fig.height=5, layout="l-page"}
margin_map_for_year(pres_elecs, city_town_geom, counties, 1992)
```

Bill Clinton (D) 62% over George HW Bush (R) 38%

:::

::: {.panel}

### 1996 {.panel-name}

```{r map-1996, fig.height=5, layout="l-page"}
margin_map_for_year(pres_elecs, city_town_geom, counties, 1996)
```

Bill Clinton (D) 69% over Bob Dole (R) 31%

:::

::: {.panel}

### 2000 {.panel-name}

```{r map-2000, fig.height=5, layout="l-page"}
margin_map_for_year(pres_elecs, city_town_geom, counties, 2000)
```

Al Gore (D) 65% over George W Bush (R) 35%

:::

::: {.panel}

### 2004 {.panel-name}

```{r map-2004, fig.height=5, layout="l-page"}
margin_map_for_year(pres_elecs, city_town_geom, counties, 2004)
```

John Kerry (D) 63% over George W Bush (R) 37%

:::

::: {.panel}

### 2008 {.panel-name}

```{r map-2008, fig.height=5, layout="l-page"}
margin_map_for_year(pres_elecs, city_town_geom, counties, 2008)
```

Barack Obama (D) 63% over John McCain (R) 37%

:::

::: {.panel}

### 2012 {.panel-name}

```{r map-2012, fig.height=5, layout="l-page"}
margin_map_for_year(pres_elecs, city_town_geom, counties, 2012)
```

Barack Obama (D) 62% over Mitt Romney (R) 38%

:::

::: {.panel}

### 2016 {.panel-name}

```{r map-2016, fig.height=5, layout="l-page"}
margin_map_for_year(pres_elecs, city_town_geom, counties, 2016)
```

Hillary Clinton (D) 65% over Donald Trump (R) 35%

:::

::: {.panel}

### 2020 {.panel-name}

```{r map-2020, fig.height=5, layout="l-page"}
margin_map_for_year(pres_elecs, city_town_geom, counties, 2020)
```

Joe Biden (D) 67% over Donald Trump (R) 33%

:::

:::::

:::

## Voting History for Every City/Town

Sorted by 2020 Democratic vote margin

```{r margin-table, layout="l-page"}
ct_margin_table %>%
    arrange(`2020`) %>%
    gt() %>%
    tab_header(title="MA Dem-GOP Presidential margin 1972-2020") %>%
    fmt_number(-city_town, decimals=0) %>%
    gt_color_rows(-city_town,
                  domain=c(-100, 100),
                  palette=c("#FF4949", "white", "#0F4392")) %>%
    tab_style(style=list(cell_text(weight="bold")),
              locations=cells_column_labels(columns=`2020`))
```

## Consistently Republican towns

```{r cluster}
num_clusters <- 35

clustering <- kmeans((ct_margin_table %>%
                      column_to_rownames(var="city_town")),
                     num_clusters)

ct_margin_table$cluster <- clustering$cluster
gop_cluster_num <- ct_margin_table %>%
    filter(city_town == "Granville") %>%
    pull(cluster)
gop_cluster <- ct_margin_table %>%
    filter(cluster == gop_cluster_num) %>%
    select(-cluster)
dem_cluster_num <- ct_margin_table %>%
    filter(city_town == "Cambridge") %>%
    pull(cluster)
dem_cluster <- ct_margin_table %>%
    filter(cluster == dem_cluster_num) %>%
    select(-cluster)
```

The only Massachusetts municipality that has voted Republican in every
presidential election from 1972 through 2020 is Granville, a small
town in Hampden County on the Connecticut border. It's western
neighbor, Tolland, came close to sharing this distinction, but was
prevented by a 75 to 75 vote tie between President Bill Clinton and
challenger Senator Bob Dole in 1996.

```{r gop-towns}
gop_cluster %>%
    gt() %>%
    tab_header(title="Historically Republican towns (Dem margin 1972-2020)") %>%
    fmt_number(-city_town, decimals=0) %>%
    gt_color_rows(-city_town,
                  domain=c(-100, 100),
                  palette=c("#FF4949", "white", "#0F4392")) %>%
    tab_footnote(footnote="Granville is the only city/town to vote Republican in every election",
                 locations=cells_body(
                     columns=city_town,
                     rows=(city_town == "Granville"))) %>%
    tab_footnote(footnote="Tolland's only non-GOP victory was a 75-75 vote Clinton-Dole tie in 1996",
                 locations=cells_body(
                     columns=`1996`,
                     rows=(city_town == "Tolland")))
```

All four of these solidly Republican towns are in the western
Massachusetts county of Hampden. The Democratic leaning Hampden cities
like Springfield, Chicopee, and Holyoke keeps Hampden County from
being the most Republican in the state.

## The bluest of the blue

The most Democratic municipalities in Massachusetts include the
university-dominated cities of Amherst and Cambridge, and two towns on
the ocean, Provincetown on the tip of Cape Cod and Aquinnah on the
western end of the island of Martha's Vineyard.

```{r dem-towns}
dem_cluster %>%
    gt() %>%
    tab_header(title="Historically Democratic cities/towns (Dem margin 1972-2020)") %>%
    fmt_number(-city_town, decimals=0) %>%
    gt_color_rows(-city_town,
                  domain=c(-100, 100),
                  palette=c("#FF4949", "white", "#0F4392")) %>%
    tab_footnote(footnote="Aquinnah had the largest average Democratic margin of 73 points",
                 locations=cells_body(
                     columns=city_town,
                     rows=(city_town == "Aquinnah"))) %>%
    tab_footnote(footnote="Provincetown had the largest median Democratic margin of 74.5 points",
                 locations=cells_body(
                     columns=city_town,
                     rows=(city_town == "Provincetown")))
```

Communities with colleges and universities are often more liberal than
surrounding communities (but is is [not
clear](https://academic.oup.com/ijpor/article-abstract/14/2/141/719728)
whether this is a causal relationship).

The resort town of Provincetown has had a long association and support
for the LGBTQ+ community and progressive politics in general. The
island town of Aquinnah is an import ancestral area for the indigenous
Wampanoag people and continues to have a significant indigigenous
population, which [may explain some of its Democratic
strength](https://www.brookings.edu/blog/how-we-rise/2022/11/15/native-americans-support-democrats-over-republicans-across-house-and-senate-races/)).

## Which towns have changed the most?

```{r margin-change-map, fig.height=5, layout="l-page"}
ct_margin_vs_state_change <- city_town_elecs %>%
    select(city_town, year, dem_margin_vs_state) %>%
    pivot_wider(id_cols=city_town,
                names_from=year,
                values_from=dem_margin_vs_state) %>%
    select(city_town, `1972`, `2020`) %>%
    mutate(margin_change = `2020` - `1972`) %>%
    arrange(margin_change)

ct_margin_vs_state_change_geom <- city_town_geom %>%
    left_join(ct_margin_vs_state_change, by="city_town")

(tm_shape(ct_margin_vs_state_change_geom) +
 tm_fill(col="margin_change",
         title="Democratic Margin Change",
         popup.vars=c(
             "margin_change",
             "1972",
             "2020"),
         popup.format=list(digits=0),
         alpha=0.7,
         palette=c("#FF4949", "white", "#0F4392"),
         breaks=c(-70, -40, -20, -10, 0, 10, 20, 40, 70, 101),
         midpoint=0) +
 tm_borders(lwd=1) +
 tm_shape(counties) +
 tm_borders(lwd=2, col="black") +
 tm_basemap("OpenStreetMap") +
 tm_view(set.view=8))
```

The areas of Massachusetts that have become more liberal over the last
50 years are the western Massachusets counties of Berkshire,
Hampshire, and Franklin counties, and also the ring of affluent Boston
suburban communities that may still vote for moderate Republicans like
Charlie Baker, but voted in large numbers for Hillary Clinton in 2016
and by even larger margins for Joe Biden in 2020.

The towns that have moved the most towards the GOP since 1972 are
Achushent (-60), Millville (-56), Ware (-52), and (Dudley -49).

The towns that have moved the most in the Democratic direction are
Mount Washington (+101), Alford (+85), Gosnold (+83), and Shelburne
(+80).

Regardless of national trends, there is little sign that Massachusetts
will become universally blue and that some towns will remain
sympathetic to Republican candidates and conservative attitudes.
