---
title: "MA 2022 State Representative General Election Ratings"
description: |
   A model-based evaluation of the contested Massachusetts State Representative races
author: Brent Benson
date: 2022-10-24
citation: false
preview: https://www.massnumbers.us/images/ma_state_rep_ratings_2022.png
output:
  distill::distill_article:
    self_contained: false
draft: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)

library(metathis)
library(tidyverse)
library(readxl)
library(rstanarm)
library(humaniformat)
library(gt)
library(gtExtras)
library(sf)
library(tmap)
library(formattable)
```

```{r post-meta}
meta() %>%
    meta_description("A model-based evaluation of the contested State Representative seats") %>%
    meta_viewport() %>%
    meta_social(
        title = "MA 2022 State Representative General Election Ratings",
        url = "https://www.massnumbers.us/posts/2022-10-24-state-rep-ratings/",
        image = "https://www.massnumbers.us/images/ma_state_rep_ratings_2022.png",
        image_alt = "MA 2020 State Representative general election ratings",
        og_type = "article",
        og_author = "Brent Benson",
        twitter_card_type = "summary_large_image",
        twitter_creator = "@bwbensonjr"
    )
```

```{r read-data}
party_abbrev <- tibble(
    party = c("Democratic",
              "Independent",
              "Republican",
              "Unenrolled",
              "Workers Party",
              "Green-Rainbow Party",
              "We The People",
              "Non-Party Candidate"),
    pty = c("D",
            "I",
            "R",
            "U",
            "W",
            "GR",
            "WTP",
            "NP")
)

rc <- read_csv("../../_data/2020/ma_state_rep_cands_2022.csv") %>%
    left_join(party_abbrev, by="party") %>%
    mutate(cand_name_parsed = parse_names(candidate),
           first_name = cand_name_parsed$first_name,
           last_name = cand_name_parsed$last_name,
           cand_label = str_c(last_name, " (", pty, "-", city_town, ")")) %>%
    select(-office, -cand_name_parsed) %>%
    nest(candidate = c(candidate, address, city_town, party, pty, first_name, last_name, cand_label)) %>%
    mutate(num_candidates = map_dbl(candidate, nrow))

rr <- read_excel("../../_data/2020/MA_State_Rep_Races_2022.xlsx") %>%
    select(district = District,
           dist_num = Number,
           dist_code = Code,
           PVI_N,
           PVI,
           inc_status,
           Incumbent,
           dem_win_prob) %>%
    mutate(district = str_replace(district, " & ", " and ")) %>%
    left_join(rc, by="district") %>%
    unnest(candidate) %>%
    group_by(district) %>%
    mutate(is_incumbent = case_when(
              ((inc_status == "Dem_Incumbent") & (pty == "D")) ~ TRUE,
              ((inc_status == "GOP_Incumbent") & (pty == "R") & (district != "2nd Franklin")) ~ TRUE,
              ((district == "2nd Franklin") & (candidate == "Susannah M. Whipps")) ~ TRUE,
              TRUE ~ FALSE)) %>%
    ungroup() %>%
    nest(candidate = c(candidate,
                       address,
                       city_town,
                       party,
                       pty,
                       cand_label,
                       first_name,
                       last_name,
                       is_incumbent))
```

```{r leg-model}
ma_leg <- read_csv("../../_data/2020/ma_leg_two_party_2008_2021_pvi.csv") %>%
    mutate(dem_win = if_else(dem_percent > 0.5, 1, 0)) %>%
    filter(!is_special)

bm2 <- stan_glm(dem_win ~ PVI_N + incumbent_status + pres_year, data=ma_leg, family=binomial(link="logit"), refresh=0)
blm2 <- stan_glm(dem_percent ~ PVI_N + incumbent_status + pres_year, data=ma_leg, family=gaussian(link="identity"), refresh=0)
```

```{r infer-races}
rr$bayes_win_prob <- predict(bm2, newdata=(rr %>% rename(incumbent_status=inc_status) %>% mutate(pres_year = FALSE)), type="response")
rr$bayes_dem_pct <- predict(blm2, newdata=(rr %>% rename(incumbent_status=inc_status) %>% mutate(pres_year = FALSE)), type="response")
rep_intervals <- predictive_interval(blm2, newdata=(rr %>% rename(incumbent_status=inc_status) %>% mutate(pres_year = FALSE)))
rr$bayes_dem_low <- rep_intervals[,1]
rr$bayes_dem_high <- rep_intervals[,2]
```

```{r table-transforms}
flatten_candidates <- function(district, cands) {
    cands %>%
        select(pty, cand_label) %>%
        mutate(pty = case_when(
                   pty == "D" ~ "D",
                   pty == "R" ~ "R",
                   ((pty == "I") & (district == "2nd Franklin")) ~ "D",
                   TRUE ~ "other")) %>%
        pivot_wider(names_from = "pty",
                    values_from = "cand_label")
}

table_transforms <- function(df) {
    df %>%
        mutate(margin = (bayes_dem_pct - 0.5)*100,
               margin_low = (bayes_dem_low - 0.5)*100,
               margin_high = (bayes_dem_high - 0.5)*100,
               Incumbent = case_when(
                   inc_status == "Dem_Incumbent" ~ "Dem",
                   inc_status == "GOP_Incumbent" ~ "GOP",
                   TRUE ~ ""),
               candidates = map2(district, candidate, flatten_candidates)) %>%
        select(district, PVI, PVI_N, Incumbent, num_candidates,
               candidates, dem_win_prob=bayes_win_prob,
               margin, margin_low, margin_high) %>%
        unnest(candidates)
}

rr_flat <- rr %>%
    table_transforms()
```

```{r state-rep-map, layout="l-body-outset", fig.height=5, fig.cap="Contested State Representative Races"}
tmap_mode(mode="view")

rep_geom <- read_sf("../../_gis/HOUSE2021/HOUSE2021_POLY.shp") %>%
    st_make_valid() %>%
    st_simplify(dTolerance=100, preserveTopology=TRUE) %>%
    select(district = REP_DIST) %>%
    right_join(rr_flat, by="district") %>%
    mutate(contested = (num_candidates > 1),
           margin = if_else(contested, margin, NA_real_),
           margin_low = if_else(contested, margin_low, NA_real_),
           margin_high = if_else(contested, margin_high, NA_real_),
           dem_win_prob = if_else(contested, dem_win_prob, NA_real_))

race_popup_vars <- c(
    "PVI",
    "Incumbent",
    "D", "R", "other",
    "dem_win_prob",
    "margin",
    "margin_low",
    "margin_high"
)

percent_zero <- function(n) {
    percent(n, 0)
}

digits_zero <- function(n) {
    digits(n, digits=0)
}

race_popup_format <- list(
    dem_win_prob=list(fun=percent_zero),
    margin=list(fun=digits_zero),
    margin_low=list(fun=digits_zero),
    margin_high=list(fun=digits_zero)
)

(tm_shape(rep_geom) +
    tm_fill(col="margin",
            palette=c("#FF4949", "white", "#0F4392"),
            midpoint=0,
            alpha=0.7,
            textNA="Single candidate",
            popup.format=race_popup_format,
            popup.vars=race_popup_vars) +
    tm_borders(lwd=1) +
    tm_basemap("OpenStreetMap") +
    tm_view(set.view=8))
```

The competitive landscape for the contested State Representative races
to be decided on November 8th range from a very likely GOP hold by incumbent
Nicholas Boldyga (R-Southwick) in the Western Massachusetts Republican 
stronghold of the 3rd Hampden District, to a sure hold in the D+42
5th Suffolk District where Chris Worrell should replace Rep. Liz Miranda
who is running for State Senator.

In order to understand where the other State Rep elections fall between these
extremes we have used a simple regression model trained with historical data to
provide an estimated range of possible outcomes.

## The Model

The Bayesian regression model has three variables: a measure of partisanship
(the Partisan Voter Index), an incumbency factor (Democratic incumbent, no
incumbent, or GOP incumbent), and an adjustment for presidential election years.
The model is trained with 424 state legislative elections from 2008 through 2021
and explains a little over 70 percent of variance in the Democratic percentage
of the vote. The other 30 percent of the variance is due to important factors
like candidate and campaign quality, which will determine the results in
competitive races. The general, fundamentals-based nature of the model is also
why these ratings should only be used to understand the competitive landscape
rather than as a prediction of race results.

The model does not adjust for third-party, independent, or unenrolled
candidates, as a simple treatment of the third candidate as an additional model
parameter does not provide additional explanatory value. An improved model might
treat different types of third-party, independent, and unenrolled candidates
differently, depending on their role with respect to the other candidate(s)
(challenge from the left for Democrat, libertarian challenge to establishment
Republican, etc.). This would require investigation and classification of each
individual historical race, and each race being put through the model for
inference purposes. We are considering adding additional features to the model
that do not require this type of manual coding, including length of incumbency
tenure, and an adjustment for a candidate *home town effect*.

The Bayesian regression model allows us to interpret the results in a
probabilistic context and also provides easily interpretable confidence
intervals.

As an example, let's take the 1st Middlesex race between Margaret Scarsdale
(D-Pepperell), Andrew Shephard (R-Townsend), and independent candidate Catherine
Lundeen of Pepperell. The Partisan Voter Index (PVI) of the redistricted 1st
Middlesex District is D+2 (or 2.27 with a bit more precision) meaning that over
the last two presidential elections the district is 2.27 points more Democratic
than the country as a whole. The incumbency variable is set to no incumbent as
former 1st Middlesex Rep. Sheila Harrington (R-Groton) left the state house to
become Clerk Magistrate of the Gardner District Court. The presidential election
variable is false, as it is for all the 2022, off-year races.

If we use the trained model to probabilistically simulate the election several
thousand times (*sampling from the posterior distribution* in technical
parlance) we see the Democrat winning 53% of the time, and the Republican 47% of
the time. If we look at the simulated Democratic percent of the vote, rather
than the win/loss percentage, we see an average Democratic percent of 51% of the
vote&mdash;making this the most competitive State Rep seat with respect to this
model. We can also generate a 90% onfidence interval for the voting result with
the Democratic percentage varying from 40% to 62%&mdash;a wide enough interval
to allow for a comfortable GOP win, or a large Democratic victory. The most
likely outcome will be a reasonably narrow victory for Scarsdale or Shephard.

## Race Ratings

For each race we show the candidates, district Partisvan Voter Index (PVI),
and the output of the model: a Democratic win probability, and a projected
Democratic one-way margin (points above 50/50 tie) with a 90% confidence
interval. It is important to remember that this fundamentals-based model
does not take into account particulars of each race including important
factors like candidate quality and campaign efforts and outreach, which will
actually determine the results in competitive races.

The one race that needs to be interpreted differently is the 2nd Franklin
race between Susannah Whipps (I-Athol) and Jeffrey Raymond (R-Athol). We
are treating Whipps as an incumbent Democrat, with respect to the model,
given that she has a Republican opponent. It would have made sense
to treat her as a Republican incumbent if she had a Democratic opponent.

The highlighted races in the table have both candidates with at least a 10%
chance of winning and non-trivial margins on both sides of the break-even margin
line.

```{r state-rep-table, layout="l-screen"}
state_rep_table <- rr_flat %>%
    filter(num_candidates > 1) %>%
    select(-num_candidates) %>%
    arrange(margin) %>%
    gt() %>%
    gt_theme_538() %>%
    sub_missing(missing_text = "") %>%
    fmt_percent(columns="dem_win_prob", decimals=0) %>%
    gt_plt_conf_int(margin,
                    c(margin_low, margin_high),
                    palette=c("black", "black", "white", "black"),
                    ref_line=0.5,
                    text_size=2.5,
                    width=55) %>%
    cols_hide(columns=c(PVI_N, margin_low, margin_high)) %>%
    cols_align(columns=c("margin"), align="center") %>%
    gt_highlight_rows(font_weight="normal",
                      rows=((dem_win_prob > .1) & (dem_win_prob < .9))) %>%
    tab_footnote(footnote="Whipps is independent, but is treated as an incumbent Democrat for modeling this race.",
                 locations=cells_body(
                     columns=Incumbent,
                     rows=(district == "2nd Franklin"))) %>%
    tab_header("Contested State Rep race ratings")

state_rep_table
```

