---
title: "State Representative General Election Ratings"
description: |
   A model-based evaluation of the contested State Representative seats
author: Brent Benson
date: 2022-10-24
output:
  distill::distill_article:
    self_contained: false
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)

library(metathis)
library(tidyverse)
library(readxl)
library(rstanarm)
library(humaniformat)
library(gt)
library(gtExtras)
```

```{r read-data}
party_abbrev <- tibble(
    party = c("Democratic",
              "Independent",
              "Republican",
              "Unenrolled",
              "Workers Party",
              "Green-Rainbow Party",
              "We The People",
              "Non-Party Candidate"),
    pty = c("D",
            "I",
            "R",
            "U",
            "W",
            "GR",
            "WTP",
            "NP")
)

sc <- read_csv("../../_data/2020/ma_state_senate_cands_2022.csv") %>%
    left_join(party_abbrev, by="party") %>%
    mutate(cand_name_parsed = parse_names(candidate),
           first_name = cand_name_parsed$first_name,
           last_name = cand_name_parsed$last_name,
           cand_label = str_c(last_name, " (", pty, "-", city_town, ")")) %>%
    select(-office, -cand_name_parsed) %>%
    nest(candidate = c(candidate, address, city_town, party, pty, first_name, last_name, cand_label)) %>%
    mutate(num_candidates = map_dbl(candidate, nrow))

sr <- read_excel("../../_data/2020/MA_State_Senate_Race_2022.xlsx") %>%
    select(district = District,
           dist_num = Number,
           PVI_N,
           PVI,
           inc_status,
           Incumbent,
           dem_win_prob) %>%
    left_join(sc, by="district") %>%
    unnest(candidate) %>%
    group_by(district) %>%
    mutate(is_incumbent = case_when(
              ((inc_status == "Dem_Incumbent") & (pty == "D")) ~ TRUE,
              ((inc_status == "GOP_Incumbent") & (pty == "R")) ~ TRUE,
              TRUE ~ FALSE)) %>%
    ungroup() %>%
    nest(candidate = c(candidate,
                       address,
                       city_town,
                       party,
                       pty,
                       first_name,
                       last_name,
                       cand_label,
                       is_incumbent))

rc <- read_csv("../../_data/2020/ma_state_rep_cands_2022.csv") %>%
    left_join(party_abbrev, by="party") %>%
    mutate(cand_name_parsed = parse_names(candidate),
           first_name = cand_name_parsed$first_name,
           last_name = cand_name_parsed$last_name,
           cand_label = str_c(last_name, " (", pty, "-", city_town, ")")) %>%
    select(-office, -cand_name_parsed) %>%
    nest(candidate = c(candidate, address, city_town, party, pty, first_name, last_name, cand_label)) %>%
    mutate(num_candidates = map_dbl(candidate, nrow))

rr <- read_excel("../../_data/2020/MA_State_Rep_Races_2022.xlsx") %>%
    select(district = District,
           dist_num = Number,
           dist_code = Code,
           PVI_N,
           PVI,
           inc_status,
           Incumbent,
           dem_win_prob) %>%
    mutate(district = str_replace(district, " & ", " and ")) %>%
    left_join(rc, by="district") %>%
    unnest(candidate) %>%
    group_by(district) %>%
    mutate(is_incumbent = case_when(
              ((inc_status == "Dem_Incumbent") & (pty == "D")) ~ TRUE,
              ((inc_status == "GOP_Incumbent") & (pty == "R") & (district != "2nd Franklin")) ~ TRUE,
              ((district == "2nd Franklin") & (candidate == "Susannah M. Whipps")) ~ TRUE,
              TRUE ~ FALSE)) %>%
    ungroup() %>%
    nest(candidate = c(candidate,
                       address,
                       city_town,
                       party,
                       pty,
                       cand_label,
                       first_name,
                       last_name,
                       is_incumbent))
```

```{r leg-model}
ma_leg <- read_csv("../../_data/2020/ma_leg_two_party_2008_2021_pvi.csv") %>%
    mutate(dem_win = if_else(dem_percent > 0.5, 1, 0)) %>%
    filter(!is_special)

bm2 <- stan_glm(dem_win ~ PVI_N + incumbent_status + pres_year, data=ma_leg, family=binomial(link="logit"), refresh=0)
blm2 <- stan_glm(dem_percent ~ PVI_N + incumbent_status + pres_year, data=ma_leg, family=gaussian(link="identity"), refresh=0)
```

```{r predict-races}
rr$bayes_win_prob <- predict(bm2, newdata=(rr %>% rename(incumbent_status=inc_status) %>% mutate(pres_year = FALSE)), type="response")
rr$bayes_dem_pct <- predict(blm2, newdata=(rr %>% rename(incumbent_status=inc_status) %>% mutate(pres_year = FALSE)), type="response")
rep_intervals <- predictive_interval(blm2, newdata=(rr %>% rename(incumbent_status=inc_status) %>% mutate(pres_year = FALSE)))
rr$bayes_dem_low <- rep_intervals[,1]
rr$bayes_dem_high <- rep_intervals[,2]

sr$bayes_win_prob <- predict(bm2, newdata=(sr %>% rename(incumbent_status=inc_status) %>% mutate(pres_year = FALSE)), type="response")
sr$bayes_dem_pct <- predict(blm2, newdata=(sr %>% rename(incumbent_status=inc_status) %>% mutate(pres_year = FALSE)), type="response")
sen_intervals <- predictive_interval(blm2, newdata=(sr %>% rename(incumbent_status=inc_status) %>% mutate(pres_year = FALSE)))
sr$bayes_dem_low <- sen_intervals[,1]
sr$bayes_dem_high <- sen_intervals[,2]
```


We will use a simple regression model trained with historical data to give
a feel for how competitive the races are in the contested 2022 Massachusetts
State Representative races.

## The Model

The Bayesian regression model has three variables: a measure of partisanship
(the Partisan Voter Index), an incumbency factor (Democratic incumbent, no
incumbent, or GOP incumbent), and an adjustment for presidential election years.
The model is trained with 424 state legislative elections from 2008 through 2021
and explains a little over 70 percent of variance in the Democratic percentage
of the vote. The other 30 percent of the variance would be explained by
important factors like candidate and campaign quality, which will determine the
results in competitive races. That is also why these ratings should only be used
to get a feel for the competitive landscape rather than as a prediction of each
race's result.

The model does not specifically take into account third-party, independent, or
unenrolled candidates, as a simple treatment of the third candidate as an
additional model parameter does not provide more explanatory value. An improved
model might treat different types of third-party, independent, and unenrolled
candidates differently, depending on their role with respect to the other
candidate(s) (challenge from the left for Democrat, libertarian challenge to
establishment Republican, etc.). This would require investigation and
classification of each individual historical race, and each race being put
through the model for inference purposes. We are considering adding additional
features to the model that do not require this type of manual coding, including
amount of incumbency tenure, and an adjustment for a candidate *home town
effect*.

The Bayesian regression model allows us to interpret the results in a
probabilistic context and also provides easily interpretable confidence
intervals.

As an example, let's take the 1st Middlesex race between Margaret Scarsdale
(D-Pepperell) and Andrew Shephard (R-Townsend) (and independent candidate
Catherine Lundeen of Pepperell). The Partisan Voter Index (PVI) of the
redistricted 1st Middlesex District is D+2 (or 2.27 with a bit more precision)
meaning that over the last two presidential elections the district is 2.27
points more Democratic than the country as a whole. The incumbency variable
indicates no incumbent as Rep. Sheila Harrington (R-Groton) left the state house
to become Clerk Magistrate of the Gardner District Court. The presidential
election variable is false, as it is for all the 2022, off-year races.

If we use the trained model to probabilistically simulate the election several
thousand times (*sampling from the posterior distribution* in fancy
Bayesian-speak) we see the Democrat winning 53% of the time, and the Republican
47% of the time. If we look at the simulated Democratic percent of the vote,
rather than the win/loss percentage, we see an average Democratic percent of 51%
of the vote&mdash;making this the most competitive State Rep seat with respect
to this model. We can also generate a 90% onfidence interval for the voting
result with the Democratic percentage varying from 40% to 62%&mdash;a wide
enough interval to allow for a comfortable GOP win, or a large Democratic
victory. However, the most likely outcome will be a slim victory for Scarsdale
or Shephard.

## Race Ratings

For each race we show the candidates, district Partisvan Voter Index (PVI),
and the output of the model: a Democratic win probability, and a projected
Democratic one-way margin (points above 50/50 tie) with a 90% confidence
interval. It is important to remember that this fundamentals-based model
does not take into account particulars of each race including important
factors like candidate quality and campaign efforts and outreach, which will
actually determine the results in competitive races.

The one race that needs to be interpreted differently is the 2nd Franklin
race between Susannah Whipps (I-Athol) and Jeffrey Raymond (R-Athol). We
are treating Whipps as an incumbent Democrat, with respect to the model,
given that she has a Republican opponent. We probably would have chosen
to treat her as a Republican incumbent if she had a Democratic opponent.

The highlighted races in the table have both candidates with at least a 10%
chance of winning and non-trivial margins on both sides of the break-even line.

```{r table-transforms}
flatten_candidates <- function(district, cands) {
    cands %>%
        select(pty, cand_label) %>%
        mutate(pty = case_when(
                   pty == "D" ~ "D",
                   pty == "R" ~ "R",
                   ((pty == "I") & (district == "2nd Franklin")) ~ "D",
                   TRUE ~ "other")) %>%
        pivot_wider(names_from = "pty",
                    values_from = "cand_label")
}

table_transforms <- function(df) {
    df %>%
        mutate(margin = (bayes_dem_pct - 0.5)*100,
               margin_low = (bayes_dem_low - 0.5)*100,
               margin_high = (bayes_dem_high - 0.5)*100,
               Incumbent = case_when(
                   inc_status == "Dem_Incumbent" ~ "Dem",
                   inc_status == "GOP_Incumbent" ~ "GOP",
                   TRUE ~ ""),
               candidates = map2(district, candidate, flatten_candidates)) %>%
        select(district, PVI, Incumbent, num_candidates,
               candidates, dem_win_prob=bayes_win_prob,
               margin, margin_low, margin_high) %>%
        unnest(candidates)
}

sr_flat <- sr %>%
    table_transforms()

rr_flat <- rr %>%
    table_transforms()
```

```{r state-rep-table, layout="l-screen"}
state_rep_table <- rr_flat %>%
    filter(num_candidates > 1) %>%
    select(-num_candidates) %>%
    arrange(margin) %>%
    gt() %>%
    gt_theme_538() %>%
    sub_missing(missing_text = "") %>%
    fmt_percent(columns="dem_win_prob", decimals=0) %>%
    gt_plt_conf_int(margin,
                    c(margin_low, margin_high),
                    palette=c("black", "black", "white", "black"),
                    ref_line=0.5,
                    text_size=2.5,
                    width=55) %>%
    cols_hide(columns=c(margin_low, margin_high)) %>%
    cols_align(columns=c("margin"), align="center") %>%
    gt_highlight_rows(font_weight="normal",
                      rows=((dem_win_prob > .1) & (dem_win_prob < .9))) %>%
    tab_footnote(footnote="Whipps is independent, but is treated as an incumbent Democrat for modeling this race.",
                 locations=cells_body(
                     columns=Incumbent,
                     rows=(district == "2nd Franklin"))) %>%
    tab_header("Contested State Rep ratings")

state_rep_table
```

