---
title: "A look at the state primary results"
description: |
  There were ousted incumbents, some real nail-biters, and a slate of new name winners for open seats
author: Brent Benson
date: 2024-09-14
citation: false
preview: https://www.massnumbers.us/images/state_rep_primaries_2024.png
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)

library(tidyverse)
library(readxl)
library(glue)
library(gt)
library(sf)
library(tmap)
library(metathis)
```

```{r post-meta}
meta() %>%
    meta_description("There were ousted incumbents, some real nail-biters, and a slate of new name winners for open seats") %>%
    meta_viewport() %>%
    meta_social(
        title = "A look at the state primary results",
        url = "https://www.massnumbers.us/posts/2024-09-14-primary-results/",
        image = "https://www.massnumbers.us/images/state_rep_primaries_2024.png",
        image_alt = "MA 2024 contested State Representative primaries map",
        og_type = "article",
        og_author = "Brent Benson",
        twitter_card_type = "summary_large_image",
        twitter_creator = "@bwbensonjr"
    )
```

```{r}

tmap_mode(mode="view")

# Color of group-row
group_color <- "lightgrey"

# Leave out registrars and clerks
offices_to_cover <- c(
    "State Representative",
    "State Senate",
    "Governor's Council",
    "Representative in Congress",
    "Senator in Congress"
)      

candidates <- read_excel("../../_data/2024/2024_candidates_primary_results.xlsx") %>%
    # Column transformations
    mutate(is_incumbent = (`Incumbent Y/N` == "Y"),
           is_winner = (winner == "Y"),
           district = str_replace(District, " & ", " and "),
           candidate = str_to_title(`Candidate`),
                      street_address = str_split_i(Address, ", ", 1),
           city_town = str_split_i(Address, ", ", 2),
           display_name = str_glue("{candidate} ({Party}-{city_town})")) %>%
    # Renaming and selection of columns to keep
    select(office = Office,
           district,
           district_id = `District ID`,
           candidate,
           display_name,
           is_incumbent,
           votes,
           is_winner,
           party = Party,
           street_address,
           city_town,
           statement = `Statement of Public Office/Political Designation`)

calculated_incumbents <- read_csv("../../_data/2024/ma_incumbents_2024_08_03.csv") %>%
    mutate(district = str_replace(district, " & ", " and "),
           incumbent_party = str_sub(party_incumbent, 1, 1),
           incumbent_display = str_glue("{name_incumbent} ({incumbent_party}-{city_town_incumbent})")) %>%
    select(office,
           district,
           incumbent = name_incumbent,
           incumbent_display,
           incumbent_party,
           incumbent_city_town = city_town_incumbent)

running_incumbents <- candidates %>%
    filter(is_incumbent) %>%
    select(office,
           district,
           incumbent=candidate,
           incumbent_display=display_name,
           incumbent_party=party,
           incumbent_city_town=city_town)

# Give the set of candidates for a particular district,
# return a dataframe with a column per party with the
# count of the candidates in that party.
party_counts <- function(candidates) {
    candidates %>%
        group_by(party) %>%
        tally() %>%
        pivot_wider(names_from=party, values_from=n)
}

district_summaries <- read_excel("../../_data/2024/notable_primary_results_2024.xlsx") %>%
    filter(!is.na(summary)) %>%
    select(district_id, summary)

districts <- candidates %>%
    # Transform into row-per seat
    nest(candidate = c(candidate,
                       display_name,
                       is_incumbent,
                       party,
                       is_winner,
                       votes,
                       street_address,
                       city_town,
                       statement)) %>%
    # Join to incumbency data
    left_join(running_incumbents, by=c("office", "district")) %>%
    # District-level data
    mutate(num_candidates = map_int(candidate, nrow),
           is_contested = (num_candidates > 1),
           is_open = is.na(incumbent),
           parties = map(candidate, party_counts)) %>%
    unnest(parties) %>%
    # This only handles D, R, and U for `party`
    replace_na(list(D=0, R=0, U=0)) %>%
    mutate(inc_party_count = case_when(
        incumbent_party == "D" ~ D,
        incumbent_party == "R" ~ R,
        incumbent_party == "U" ~ U,
        TRUE ~ NA_real_)) %>%
    mutate(incumbent_primary = (inc_party_count > 1),
           dem_primary = (D > 1),
           gop_primary = (R > 1),
           party_primary = (dem_primary | gop_primary),
           primary_type = case_when(
               (dem_primary & gop_primary) ~ "Both",
               dem_primary ~ "Democratic",
               gop_primary ~ "Republican",
               TRUE ~ "None"
               ),
           contested_no_primary = (is_contested & !incumbent_primary)) %>%
    left_join(district_summaries, by="district_id") %>%
    rows_patch(calculated_incumbents,
               by=c("office", "district"),
               unmatched="ignore") %>%
    filter(office %in% offices_to_cover)

## notable_districts <- districts %>%
##     filter(office %in% offices_to_cover,
##           (incumbent_primary | is_open | contested_no_primary)) %>%
##     mutate(candidate_names = map_chr(candidate,
##                                      ~paste(.x$display_name, collapse = ", ")))
##
## notable_districts %>% write_csv("notable_districts_2024.csv")

```

While much went as expected in the September 3, 2024 state primary
elections there were some notable losses to incumbents and some close
incumbent victories while the open seat wins pave the way for a new set of
leaders in the state legislature and Governor's Council.

```{r}
primary_districts <- districts %>%
    filter((D > 1) | (R > 1))

primary_view <- primary_districts %>%
    unnest(candidate) %>%
    mutate(incumbent = if_else(is_incumbent, "Incumbent", ""),
           winner = if_else(is_winner, "Winner", "")) %>%
    group_by(office, district) %>%
    filter(((party == "D") & dem_primary) |
           ((party == "R") & gop_primary)) %>%
    arrange(party, desc(votes)) %>%
    ungroup() %>%
    group_by(office, district, party) %>%
    arrange(desc(votes)) %>%
    mutate(vote_margin = if_else(is_winner, votes - lead(votes), NA),
           margin_pct = if_else(is_winner, vote_margin/sum(votes), NA)) %>%
    ungroup() %>%
    mutate(row_group = str_glue("**{office} - {district}**<br/>*Incumbent {incumbent_display} - {summary}*"))

map_districts <- primary_view %>%
    group_by(office, district) %>%
    summarize(primary_type = first(primary_type),
              incumbent = first(incumbent_display),
              candidates = str_c(display_name, collapse="<br/>"),
              .groups="drop")

office_table_display <- function(df) {
    df %>%
        select(row_group, incumbent, winner, display_name, votes, vote_margin, margin_pct, statement) %>%
        gt(groupname_col="row_group", process_md=TRUE) %>%
        tab_options(
            column_labels.hidden=TRUE,
            row_group.background.color = group_color) %>%
        fmt_number(columns="votes", decimals=0, sep_mark=",") %>%
        fmt_number(columns="vote_margin", decimals=0, sep_mark=",", force_sign=TRUE) %>%
        fmt_percent(columns="margin_pct", decimals=0, force_sign=TRUE) %>%
        sub_missing(missing_text="") %>%
        cols_align(columns="display_name", align="right") %>%
        opt_row_striping(row_striping=FALSE)
}

office_primary_map <- function(office_group, geom, map_dists) {
    (tm_shape(geom %>%
              select(district) %>%
              left_join((map_dists %>%
                           filter(office == office_group)),
                         by="district")) +
     tm_fill(col="primary_type",
             alpha=0.7,
             popup.vars=TRUE,
             popup.format=list(html.escape=FALSE),
             palette=c("Democratic"="#3E74C6",
                       "Republican"="#FF4949",
                       "Both"="purple",
                       "None"=NA)) +
     tm_borders(lwd=1, col="black") +
     tm_basemap("OpenStreetMap") +
     tm_view(set.view=8))     
}

```
## Incumbent losses

It is very difficult to challenge and defeat a sitting state
legislator or Governor's Council member from within the same party, but
three incumbents lost in September 3 Democratic primaries.

```{r}
inc_loss_races <- primary_view %>%
    filter(incumbent_primary, is_incumbent, !is_winner) %>%
    select(office, district, party)

inc_losses <- primary_view %>%
    right_join(inc_loss_races,
               by=c("office", "district", "party"))

office_table_display(inc_losses)
```

## Close incumbent wins

There were five races for State Representative where the incumbent
victory was probably closer than they would have hoped, including a 41
vote recount win for Representative Marjorie Decker after challenger
Evan Mackay had declared victory on election night.

```{r}
inc_close_win_races <- primary_view %>%
    filter(incumbent_primary,
           is_incumbent,
           is_winner,
           margin_pct < 0.2) %>%
    select(office, district, party)

inc_close_wins <- primary_view %>%
    right_join(inc_close_win_races,
               by=c("office", "district", "party"))

office_table_display(inc_close_wins %>%
                     arrange(margin_pct))
```

## Open seat winners

There will be new faces in the Governor's Council and state
legislatures from this round of primary winners.

```{r}
office_table_display(primary_view %>%
                     filter(is_open))
```

## Full results

### U.S. Senate

```{r}
office_table_display(primary_view %>%
                     filter(office == "Senator in Congress"))
```

### U.S. Congress
 
```{r}
office_table_display(primary_view %>%
                     filter(office == "Representative in Congress"))
```

### Governor's Council

```{r, layout="l-body-outset"}

office_primary_map(
    "Governor's Council",
    read_sf("https://bwbensonjr.github.io/mapoli/gis/geojson/govcouncil2021.geojson"),
    map_districts
)
```

```{r}
office_table_display(primary_view %>%
                     filter(office == "Governor's Council"))
```

### State Senate

```{r, layout="l-body-outset"}

office_primary_map(
    "State Senate",
    read_sf("https://bwbensonjr.github.io/mapoli/gis/geojson/senate2021.geojson"),
    map_districts
)
```

```{r}
office_table_display(
    (primary_view %>%
     filter(office == "State Senate"))
)
```

### State Representative

```{r, layout="l-body-outset"}

office_primary_map(
    "State Representative",
    read_sf("https://bwbensonjr.github.io/mapoli/gis/geojson/house2021.geojson"),
    map_districts
)
```

```{r}
office_table_display(
    (primary_view %>%
     filter(office == "State Representative"))
)
```



