---
title: "MA 2022 State Legislative Results"
description: |
  No major surprises when comparing race results to model fundamentals
author: Brent Benson
date: 2022-11-14
citation: false
preview: https://www.massnumbers.us/images/state_rep_results_2022_11_08.png
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)

library(metathis)
library(tidyverse)
library(readxl)
library(rstanarm)
library(humaniformat)
library(gt)
library(gtExtras)
library(sf)
library(tmap)
library(formattable)
library(plotly)
```
```{r post-meta}
meta() %>%
    meta_description("No major surprises when comparing race results to model fundamentals") %>%
    meta_viewport() %>%
    meta_social(
        title = "MA 2022 State Legislative Results",
        url = "https://www.massnumbers.us/posts/2022-11-11-state-legislative-results/",
        image = "https://www.massnumbers.us/images/state_rep_results_2022_11_08.png",
        image_alt = "MA 2020 State Representative general election results",
        og_type = "article",
        og_author = "Brent Benson",
        twitter_card_type = "summary_large_image",
        twitter_creator = "@bwbensonjr"
    )
```

```{r read-data}
party_abbrev <- tibble(
    party = c("Democratic",
              "Independent",
              "Republican",
              "Unenrolled",
              "Workers Party",
              "Green-Rainbow Party",
              "We The People",
              "Non-Party Candidate"),
    pty = c("D",
            "I",
            "R",
            "U",
            "W",
            "GR",
            "WTP",
            "NP")
)

# State Rep
rc <- read_csv("../../_data/2020/ma_state_rep_cands_2022.csv") %>%
    left_join(party_abbrev, by="party") %>%
    mutate(cand_name_parsed = parse_names(candidate),
           first_name = cand_name_parsed$first_name,
           last_name = cand_name_parsed$last_name,
           cand_label = str_c(last_name, " (", pty, "-", city_town, ")")) %>%
    select(-office, -cand_name_parsed) %>%
    nest(candidate = c(candidate, address, city_town, party, pty, first_name, last_name, cand_label)) %>%
    mutate(num_candidates = map_dbl(candidate, nrow))

rr <- read_excel("../../_data/2020/MA_State_Rep_Races_2022.xlsx") %>%
    select(district = District,
           dist_num = Number,
           dist_code = Code,
           PVI_N,
           PVI,
           inc_status,
           Incumbent,
           dem_win_prob) %>%
    mutate(district = str_replace(district, " & ", " and ")) %>%
    left_join(rc, by="district") %>%
    unnest(candidate) %>%
    group_by(district) %>%
    mutate(is_incumbent = case_when(
              ((inc_status == "Dem_Incumbent") & (pty == "D")) ~ TRUE,
              ((inc_status == "GOP_Incumbent") & (pty == "R") & (district != "2nd Franklin")) ~ TRUE,
              ((district == "2nd Franklin") & (candidate == "Susannah M. Whipps")) ~ TRUE,
              TRUE ~ FALSE)) %>%
    ungroup() %>%
    nest(candidate = c(candidate,
                       address,
                       city_town,
                       party,
                       pty,
                       cand_label,
                       first_name,
                       last_name,
                       is_incumbent))

#State Senate
sc <- read_csv("../../_data/2020/ma_state_senate_cands_2022.csv") %>%
    left_join(party_abbrev, by="party") %>%
    mutate(cand_name_parsed = parse_names(candidate),
           first_name = cand_name_parsed$first_name,
           last_name = cand_name_parsed$last_name,
           cand_label = str_c(last_name, " (", pty, "-", city_town, ")")) %>%
    select(-office, -cand_name_parsed) %>%
    nest(candidate = c(candidate, address, city_town, party, pty, first_name, last_name, cand_label)) %>%
    mutate(num_candidates = map_dbl(candidate, nrow))

sr <- read_excel("../../_data/2020/MA_State_Senate_Race_2022.xlsx") %>%
    select(district = District,
           dist_num = Number,
           PVI_N,
           PVI,
           inc_status,
           Incumbent,
           dem_win_prob) %>%
    left_join(sc, by="district") %>%
    unnest(candidate) %>%
    group_by(district) %>%
    mutate(is_incumbent = case_when(
              ((inc_status == "Dem_Incumbent") & (pty == "D")) ~ TRUE,
              ((inc_status == "GOP_Incumbent") & (pty == "R")) ~ TRUE,
              TRUE ~ FALSE)) %>%
    ungroup() %>%
    nest(candidate = c(candidate,
                       address,
                       city_town,
                       party,
                       pty,
                       cand_label,
                       first_name,
                       last_name,
                       is_incumbent))
```

```{r leg-model}
ma_leg <- read_csv("../../_data/2020/ma_leg_two_party_2008_2021_pvi.csv") %>%
    mutate(dem_win = if_else(dem_percent > 0.5, 1, 0)) %>%
    filter(!is_special)

bm2 <- stan_glm(dem_win ~ PVI_N + incumbent_status + pres_year, data=ma_leg, family=binomial(link="logit"), refresh=0)
blm2 <- stan_glm(dem_percent ~ PVI_N + incumbent_status + pres_year, data=ma_leg, family=gaussian(link="identity"), refresh=0)
```

```{r infer-races}
rr$bayes_win_prob <- predict(bm2, newdata=(rr %>% rename(incumbent_status=inc_status) %>% mutate(pres_year = FALSE)), type="response")
rr$bayes_dem_pct <- predict(blm2, newdata=(rr %>% rename(incumbent_status=inc_status) %>% mutate(pres_year = FALSE)), type="response")
rep_intervals <- predictive_interval(blm2, newdata=(rr %>% rename(incumbent_status=inc_status) %>% mutate(pres_year = FALSE)))
rr$bayes_dem_low <- rep_intervals[,1]
rr$bayes_dem_high <- rep_intervals[,2]

sr$bayes_win_prob <- predict(bm2, newdata=(sr %>% rename(incumbent_status=inc_status) %>% mutate(pres_year = FALSE)), type="response")
sr$bayes_dem_pct <- predict(blm2, newdata=(sr %>% rename(incumbent_status=inc_status) %>% mutate(pres_year = FALSE)), type="response")
sen_intervals <- predictive_interval(blm2, newdata=(sr %>% rename(incumbent_status=inc_status) %>% mutate(pres_year = FALSE)))
sr$bayes_dem_low <- sen_intervals[,1]
sr$bayes_dem_high <- sen_intervals[,2]
```

```{r district-name-transform}
NUMS_TO_WORDS <- c(
    `&`='and',
    `1st`='First',
    `2nd`='Second',
    `3rd`='Third',
    `4th`='Fourth',
    `5th`='Fifth',
    `6th`='Sixth',
    `7th`='Seventh',
    `8th`='Eighth',
    `9th`='Ninth',
    `10th`='Tenth',
    `11th`='Eleventh',
    `12th`='Twelfth',
    `13th`='Thirteenth',
    `14th`='Fourteenth',
    `15th`='Fifteenth',
    `16th`='Sixteenth',
    `17th`='Seventeenth',
    `18th`='Eighteenth',
    `19th`='Nineteenth',
    `20th`='Twentieth',
    `21st`='Twenty-First',
    `22nd`='Twenty-Second',
    `23rd`='Twenty-Third',
    `24th`='Twenty-Fourth',
    `25th`='Twenty-Fifth',
    `26th`='Twenty-Sixth',
    `27th`='Twenty-Seventh',
    `28th`='Twenty-Eighth',
    `29th`='Twenty-Ninth',
    `30th`='Thirtieth',
    `31st`='Thirty-First',
    `32nd`='Thirty-Second',
    `33rd`='Thirty-Third',
    `34th`='Thirty-Fourth',
    `35th`='Thirty-Fifth',
    `36th`='Thirty-Sixth',
    `37th`='Thirty-Seventh'
)
```

```{r table-transforms}
flatten_candidates <- function(district, cands) {
    cands %>%
        select(pty, cand_label) %>%
        mutate(pty = case_when(
                   pty == "D" ~ "D",
                   pty == "R" ~ "R",
                   ((pty == "I") & (district == "2nd Franklin")) ~ "D",
                   TRUE ~ "other")) %>%
        pivot_wider(names_from = "pty",
                    values_from = "cand_label")
}

table_transforms <- function(df) {
    df %>%
        mutate(margin = (bayes_dem_pct - 0.5)*100,
               margin_low = (bayes_dem_low - 0.5)*100,
               margin_high = (bayes_dem_high - 0.5)*100,
               Incumbent = case_when(
                   inc_status == "Dem_Incumbent" ~ "Dem",
                   inc_status == "GOP_Incumbent" ~ "GOP",
                   TRUE ~ ""),
               candidates = map2(district, candidate, flatten_candidates)) %>%
        unnest(candidates) %>%
        select(district, PVI, PVI_N, Incumbent, num_candidates,
               D, R, other, dem_win_prob=bayes_win_prob,
               margin_low, margin, margin_high)
}

rr_flat <- rr %>%
    table_transforms()

rep_results <- read_csv("../../_data/2020/state_rep_results_2022.csv") %>%
    select(district, winning_party, pcts_reporting_pct, 
           votes_Dem, votes_GOP, votes_Other, dem_percent, dem_margin)

rr_contested <- rr_flat %>%
    filter(num_candidates > 1) %>%
    left_join(rep_results, by="district") %>%
    mutate(dem_margin = dem_margin * 100,
           winning_party = if_else(is.na(winning_party), "Unk", winning_party))

sr_flat <- sr %>%
    table_transforms()

sen_results <- read_csv("../../_data/2020/state_senate_results_2022.csv") %>%
    mutate(district = str_replace_all(district, NUMS_TO_WORDS)) %>%
    select(district, winning_party, pcts_reporting_pct, 
           votes_Dem, votes_GOP, votes_Other, dem_percent, dem_margin)

sr_contested <- sr_flat %>%
    filter(num_candidates > 1) %>%
    left_join(sen_results, by="district") %>%
    mutate(dem_margin = dem_margin * 100,
           winning_party = if_else(is.na(winning_party), "Unk", winning_party))
```

```{r table-plot-helpers}
partisan_palette <- function(x) {
     f_neg <- scales::col_numeric(
         palette = c("#FF4949", "white"),
         domain = c(-50, 0)
     )
     f_pos <- scales::col_numeric(
         palette = c("white", "#0F4392"),
         domain = c(0, 50)
     )
     ifelse(x < 0, f_neg(x), f_pos(x))
}

leg_results_table <- function(results, title) {
    results %>%
    arrange(dem_margin) %>%
    select(-c(PVI_N, num_candidates, dem_win_prob,
              votes_Dem, votes_GOP, votes_Other, dem_percent)) %>%
    gt() %>%
    gt_theme_538() %>%
    cols_hide(columns=c(pcts_reporting_pct)) %>%
    cols_label(margin_low="Low",
               margin_high="High",
               winning_party="Winner",
               dem_margin="Margin") %>%
    cols_move(columns=c(winning_party,
                        dem_margin,
                        margin_low,
                        margin,
                        margin_high),
              after=Incumbent) %>%
    sub_missing(missing_text = "") %>%
    tab_spanner("Model", columns=c(margin_low,
                                   margin,
                                   margin_high)) %>%
    fmt_number(columns=c(margin,
                         margin_low,
                         margin_high,
                         dem_margin), decimals=0) %>%
    data_color(columns=c(margin, margin_low, margin_high, dem_margin),
               colors=partisan_palette) %>%
    tab_style(style=cell_text(weight="bold"),
              locations=cells_body(columns=D,
                                   rows=(Incumbent == "Dem"))) %>%
    tab_style(style=cell_text(weight="bold"),
              locations=cells_body(columns=R,
                                   rows=(Incumbent == "GOP"))) %>%
    tab_style(style=cell_fill(color="#0F4392", alpha=0.4),
              locations=cells_body(columns=D,
                                   rows=(winning_party == "Dem"))) %>%
    tab_style(style=cell_fill(color="gray40", alpha=0.4),
              locations=cells_body(columns=D,
                                   rows=(winning_party == "Ind"))) %>%
    tab_style(style=cell_fill(color="#FF4949", alpha=0.4),
              locations=cells_body(columns=R,
                                   rows=(winning_party == "GOP"))) %>%
    tab_footnote(footnote="Not all results reported",
                 locations=cells_body(columns=dem_margin,
                                      rows=(pcts_reporting_pct != 100))) %>%
    tab_footnote(footnote="Winner has not been called by AP",
                 locations=cells_body(columns=winning_party,
                                      rows=(winning_party == "Unk"))) %>%
    tab_header(title)
}

margin_result_plot <- function(results, title) {
    results %>%
    mutate(district = fct_reorder(district, -margin)) %>%
    ggplot(aes(x=margin,
               xmin=margin_low,
               xmax=margin_high,
               y=district,
               PVI=PVI,
               Incumbent=Incumbent,
               D=D,
               R=R,
               Other=other,
               pred_margin=margin,
               dem_margin=dem_margin)) +
    geom_pointrange(size=0.2, shape="plus") +
    geom_point(aes(x=dem_margin), color="blue", size=2) +
    geom_vline(xintercept=0, linetype="dashed") +
    labs(title=title, y="") + 
    theme_minimal() +
    theme(axis.text=element_text(size=9))
}

tooltip_aes = c(
    "district",
    "PVI",
    "Incumbent",
    "D",
    "R",
    "Other",
    "pred_margin",
    "margin"
)
```

There were no major surprises in the contested Massachusetts state
legislative elections on November 8, 2022. The State Representative
race that looked like it would be the most competitive based on our
fundamentals-based model (1st Middlesex) did turn out incredibly close
and has not yet been called. Another race that looked reasonably safe
(2nd Essex) ended up closer than expected and has not been called by
the AP. A State Senate race that looked close from a fundamentals
point of view (Hampden, Hampton and Worcester) came down safely in
Democratic territory, as did a race where pundits felt GOP strength
might win out (Norfolk, Worcester and Middlesex).

## 1st Middlesex

The [model expected a
nail-biter](https://www.massnumbers.us/posts/2022-10-24-state-rep-ratings/)
in the 1st Middlesex State Representative race between Margaret
Scarsdale (D-Pepperell), Andrew Shepherd (R-Townsend), and independent
candidate Catherine Lundeen (I-Pepperell) and nail-biter is what we
got. As of this writing, Shepherd looks to be ahead by 34 votes out of
19,633 votes cast, a margin of 0.17 percentage points. The race seems
to be headed for a recount.

## 2nd Essex

The biggest surprise of the night was the tight margin in the 2nd
Essex State Representative race between incumbent Rep. Leonard Mirra
(R-Georgetown) and challenger Kristin Kassner (D-Hamilton). The model
expected a high-single digit two-party margin in favor of Mirra, but
the current count shows a tiny 84 vote lead out of 23,114 votes case,
a margin of 0.36 percentage points. This race has also not been
officially called by the Associated Press. The results are still
inside the 90% error bars of the model, but it is an outlier.

## Hampden, Hampton and Worcester

On the State Senate side our model rated the Hampden, Hampton and
Worcester District&mdash;the post-redistricting version of the
district represented by Lt. Governor candidate Eric Lesser
(D-Longmeadow)&mdash;as the most competitive race. A reader who
understood the specifics of this race [gave the
opinion](https://twitter.com/MSzafranski413/status/1587573604530769922?s=20&t=971hGtckUOe-pgrWKLOXbw)
that State Rep. Oliveira (D-Ludlow) held a strong advantage over
business person William Johsnon (R-Granby) based on the structure of
the district and the influence of Oliveira's State Representative
district. Tuesday's result seemed to bear that out with Oliveira
getting 56% of the two-party vote, not close, but still well within
the 90% confidence interval of the model.

## Norfolk, Worcester and Middlesex

I took a [fairly close
look](https://www.massnumbers.us/posts/2022-11-03-norfolk-worcester-middlesex/)
at the Norfolk, Worcester and Middlesex state senate race between
incumbent Senator Becca Rausch (D-Needham) and sitting State
Rep. Shawn Dooley (R-Norfolk). There seemed to be some conventional
wisdom that the race would be close because of the red-to-purple
reputation of the district in its previous incarnation as the Norfolk,
Worcester and Bristol with former GOP State Senators Richard Ross
(R-Wrentham) and Scott Brown (R-Wrentham).

However, the district is full of towns that have [swung in the
Democratic
direction](https://www.wgbh.org/news/politics/2018/10/30/in-blue-massachusetts-some-striking-redward-shifts)
since the Trump era. This swing pushed the model to have high
expectations of a Rausch win. Rausch's victory with 55% of the
two-party vote was fairly comfortable, and could have been higher if
not for the strong name recognition of Rep. Dooley in the area and the
[large amount of
money](https://commonwealthmagazine.org/politics/in-high-stakes-race-rausch-declares-victory-over-dooley/)
spent on boths sides. It is hard to see this district as purple
anymore with its D+12 Partisan Voter Index.

```{r rep-results-table, layout="l-screen"}
rep_results_table <- leg_results_table(rr_contested,
                                       "Contested State Rep results")

rep_results_table
```

```{r rep-compare-results-model, fig.height=10}
rep_results_plot <- margin_result_plot(rr_contested,
                                       "State Rep actual margins vs. model")

ggplotly(rep_results_plot, tooltip=tooltip_aes)
```

```{r senate-results-table, layout="l-screen"}
sen_results_table <- leg_results_table(sr_contested,
                                       "Contested State Senate results")

sen_results_table
```

```{r sen-compare-results-model, fig.height=6}
sen_results_plot <- margin_result_plot(sr_contested,
                                       "State Senate actual margins vs. model")

ggplotly(sen_results_plot, tooltip=tooltip_aes)
```

