---
title: "Swing district special election results in a 17-vote margin for Democrat"
description: |
  Democrat Lisa Field wins Third Bristol special election to replace Carol Doherty by 17 votes
author: Brent Benson
date: 2025-06-12
citation: false  
preview: https://www.massnumbers.us/images/third-bristol-special-results.png
output:
  distill::distill_article:
    self_contained: false
    fig_caption: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(gt)
library(gtExtras)
library(sf)
library(tmap)
library(formattable)
library(metathis)

## third_bristol <-
##     read_csv("http://bwbensonjr.github.io/mapoli/pvi/ma_precincts_districts_pres_2024.csv") |>
##     filter(State_Rep == "Third Bristol") |>
##     mutate(Field_25 = 0, Quintal_25 = 0) # To be filled in from town clerk results

## third_bristol |>
##     write_csv("third_bristol_2025_06_10.csv")

tmap_mode("view")
```

```{r post-meta}
meta() %>%
    meta_description("Democrat Lisa Field wins Third Bristol special election to replace Carol Doherty by 24 votes") |>
    meta_viewport() |>
    meta_social(
        title = "Swing district special election results in a 24-vote margin for Democrat",
        url = "https://www.massnumbers.us/posts/2025-06-12-third-bristol-special/",
        image = "https://www.massnumbers.us/images/third-bristol-special-results.png",
        image_alt = "Third Bristol special election precinct map",
        og_type = "article",
        og_author = "Brent Benson",
        twitter_card_type = "summary_large_image",
        twitter_creator = "@bwbensonjr"
    )
```

```{r}
precinct_display <- function(city_town, ward, precinct) {
     if_else(ward == "-",
             str_c(city_town, " ", precinct),
             str_c(city_town, " ", ward, "-", precinct))
}

third_bristol <-
    read_csv("third_bristol_2025_06_10.csv", col_types=list(precinct=col_character())) |>
    mutate(Harris_pct_24 = Harris_24/(Harris_24 + Trump_24),
           Trump_pct_24 = Trump_24/(Harris_24 + Trump_24),
           point_diff_24 = (Harris_pct_24 - Trump_pct_24) * 100,
           Field_pct_25 = Field_25/(Field_25 + Quintal_25),
           Quintal_pct_25 = Quintal_25/(Field_25 + Quintal_25),
           point_diff_25 = (Field_pct_25 - Quintal_pct_25) * 100,
           point_diff = (Field_pct_25 - Harris_pct_24) * 100,
           precinct_name = precinct_display(city_town, ward, precinct))
```

```{r layout="l-body", fig.height=6, fig.cap="Third Bristol special election results"}
third_bristol_geom <-
    read_sf("http://bwbensonjr.github.io/mapoli/gis/geojson/wards_pcts_subs_2022.geojson") |>
    rename(ward = Ward, precinct = Pct) |>
    right_join(third_bristol,
               by=c("city_town", "ward", "precinct"))

percent_zero <- function(n) {
    percent(n, 0)
}

district_map <-
   (tm_shape(third_bristol_geom) +
   tm_polygons(
      col="point_diff_25",
      title="Special Election Margin",
      popup.vars=c(
          "Margin Difference" = "point_diff",
          "Field (D)" = "Field_pct_25",
          "Quintal (R)" = "Quintal_pct_25",
          "Field Margin" = "point_diff_25",
          "Harris (D) - 2024" = "Harris_pct_24",
          "Trump (R) - 2024" = "Trump_pct_24",
          "Harris Margin" = "point_diff_24"
      ),
      popup.format=list(
          point_diff=list(digits=0),
          point_diff_25=list(digits=0),
          point_diff_24=list(digits=0),
          Field_pct_25=list(fun=percent_zero),
          Quintal_pct_25=list(fun=percent_zero),
          Harris_pct_24=list(fun=percent_zero),
          Trump_pct_24=list(fun=percent_zero)
      ),
      alpha=0.6,
      palette=c("#FF4949", "white", "#0F4392"),
      midpoint=0,
      id="precinct_name"
   ) +
   tm_text("precinct_name") +
   tm_basemap("OpenStreetMap")
)

tmap_leaflet(district_map)
```

*[This article was revised and republished on June 16, 2025 using the
final certified results from Easton and Taunton election officials.]*

Swing districts don't swing every election, but the 17 vote margin in
Tuesday's special election for Third Bristol State Representative
emphasizes how close swing district elections can be, even when the
winning party is the same as last election.

Lisa Field, the Democratic candidate from Taunton emerged with a 17
vote margin based on the results posted by
[Easton](https://eastontownma.documents-on-demand.com/Document/b21c8526-d546-f011-a451-000c29a59557/2025.06.10%20Special%20State%20Election%20-%20Results.pdf)
and
[Taunton](https://www.taunton-ma.gov/DocumentCenter/View/5606/Official-Certified-City-of-Taunton-Special-State-Election-Results-6102025)
election officials. Lawrence Quintal, the Republican candidate from
Taunton, has said that he will request a recount which is not
automatic, but requires a small number of signatures. Once the
signatures have been certified, the state takes on the funding and
organization of the recount. It is rare for a recount to change the
result.

The special election took place because of the passing of
Representative Carol Doherty (D-Taunton) who died in February from
pancreatic cancer. Lisa Field is a lobbyist for the Massachusetts
Nurses Association and Lawrence Quintal is the co-owner of Silva
Funeral Home and a Taunton City Councilor.

```{r}
vote_table <-
    third_bristol |>
    select(city_town, precinct_name, Field_25, Quintal_25) |>
    mutate(vote_margin = Field_25 - Quintal_25) |>
    gt(rowname_col="precinct_name", groupname_col="city_town") |>
    tab_header(
        title = "Third Bristol special election votes by precinct/city/town",
    ) |>
    cols_label(
        Field_25 = "Lisa Field (D)",
        Quintal_25 = "Lawrence Quintal (R)",
        vote_margin = "Margin"
    ) |>
    summary_rows(
        columns = c(Field_25, Quintal_25, vote_margin),
        fns = list(Total=~sum(., na.rm=TRUE)),
        fmt = list(~fmt_integer(., use_seps=TRUE))
    ) |>
    grand_summary_rows(
        columns = c(Field_25, Quintal_25, vote_margin),
        fns = list("Grand Total"=~sum(., na.rm=TRUE)),
        fmt = list(~fmt_integer(., use_seps=TRUE))
    )

vote_table
```

Lisa Field won the Easton precincts by 72 votes, while Taunton's
precincts created a 55 vote margin for Lawrence Quintal. Quintal had
more votes than Field in eight precincts and Field had more votes in five
precincts.

```{r layout="l-body-outset", fig.cap="Third Bristol 2025 Special Election vs. 2024 Presidential"}

partisan_palette <- function(x) {
     f_neg <- scales::col_numeric(
         palette = c("#FF4949", "white"),
         domain = c(-40, 0)
     )
     f_pos <- scales::col_numeric(
         palette = c("white", "#0F4392"),
         domain = c(0, 40)
     )
     ifelse(x < 0, f_neg(x), f_pos(x))
}

district_table <-
    third_bristol |>
    arrange(desc(point_diff_25)) |>
    select(
        precinct_name,
        Field_pct_25,
        Quintal_pct_25,
        point_diff_25,
        Harris_pct_24,
        Trump_pct_24,
        point_diff_24,
        point_diff
    ) |>
    gt() |>
    tab_header(
        title = "Third Bristol special election margin vs. 2024 presidential"
    ) |>
    tab_spanner(label="2024", columns=ends_with("_24")) |>
    tab_spanner(label="2025", columns=ends_with("_25")) |>
    data_color(
        columns = c(point_diff_25, point_diff_24),
        domain=c(-40, 40),
        fn = partisan_palette
    ) |>
    cols_label(
        precinct_name="Precinct",
        Field_pct_25="Field (D)",
        Quintal_pct_25="Quintal (R)",
        point_diff_25="Margin",
        Harris_pct_24="Harris (D)",
        Trump_pct_24="Trump (R)",
        point_diff_24="Margin",
        point_diff=""
    ) |>
    fmt_percent(
        columns=contains("_pct"),
        decimals=0
    ) |>
    fmt_number(
        columns=starts_with("point_diff"),
        decimals=0
    ) |>
    gt_plt_bar(
        column=point_diff,
        keep_column=TRUE,
    ) |>
    cols_label(DUPE_COLUMN_PLT = "") |>
    tab_spanner(
        label="Margin Difference",
        columns=c(point_diff, DUPE_COLUMN_PLT)
    )

district_table
```

In contrast to [last month's special election for the Sixth Essex
district](https://www.massnumbers.us/posts/2025-05-15-sixth-essex-special/)
where Hannah Bowen (D-Beverly) out-performed Kamala Harris'a margin in
the district by 12 points, Lisa Field slightly underperformed compared
to the 2024 presidential. Fields margin of 0 points is about 3 points
less than Harris's margin in November, 2024. 

Based on the candidate fundraising reports to the [Office of Campaign
and Political Finance](https://www.ocpf.us/filers/specialelections),
both candidates raised over $50,000 and were able to put their case in
front of prospective voters, resulting in a close election in a
district that was represented by Republican Mayer Shaunna O'Connell
(R-Taunton) for almost ten years from 2011 until 2020 before
[Rep. Doherty's win in a 2020 special
election](https://bwbensonjr.github.io/mapoli/districts/pages/state-rep-third-bristol.html).

