---
title: "The top eight most competitive state legislative races"
description: |
  A mix of GOPs incumbents in swing districts and competitive open seats look
  to be the tightest races according to our fundamentals-based model
author: Brent Benson
date: 2024-10-31
citation: false
preview: https://www.massnumbers.us/images/third_barnstable_muni_pvi.png
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(readxl)
library(rstanarm)
library(glue)
library(gt)
library(sf)
library(tmap)
library(tmaptools)
library(metathis)
```

```{r post-meta}
meta() %>%
    meta_description("A mix of GOPs incumbents in swing districts and competitive open seats look to be the tightest races according to our fundamentals-based model") %>%
    meta_viewport() %>%
    meta_social(
        title = "The top eight most competitive state legislative races",
        url = "https://www.massnumbers.us/posts/2024-10-31-competitive-seats/",
        image = "https://www.massnumbers.us/images/third_barnstable_muni_pvi.png",
        image_alt = "A map of the Third Barnstable State Representative district PVI",
        og_type = "article",
        og_author = "Brent Benson",
        twitter_card_type = "summary_large_image",
        twitter_creator = "@bwbensonjr"
    )
```

```{r}
candidates <- read_excel("../../_data/2024/2024_candidates_primary_results.xlsx") %>%
    # Column transformations
    mutate(is_incumbent = (`Incumbent Y/N` == "Y"),
           is_winner = (winner == "Y") | is.na(winner),
           district = str_replace(District, " & ", " and "),
           party = case_when(
               Party == "D" ~ "Democratic",
               Party == "R" ~ "Republican",
               Party == "U" ~ `Statement of Public Office/Political Designation`
           ),
           candidate = str_to_title(`Candidate`),
                      street_address = str_split_i(Address, ", ", 1),
           city_town = str_split_i(Address, ", ", 2),
           display_name = str_glue("{candidate} ({Party}-{city_town})")) %>%
    select(office = Office,
           district,
           district_id = `District ID`,
           candidate,
           display_name,
           is_incumbent,
           votes,
           is_winner,
           party,
           party_abbr = Party,
           street_address,
           city_town,
           statement = `Statement of Public Office/Political Designation`) %>%
    filter(office %in% c("State Senate", "State Representative"),
           is_winner) %>%
    select(-is_winner) %>%
    group_by(office, district) %>%
    filter(n() > 1) %>%
    ungroup()

summarize_candidates <- function(cands) {
    incumbent <- filter(cands, is_incumbent)
    dem <- filter(cands, party_abbr == "D")
    gop <- filter(cands, party_abbr == "R")
    third_party <- filter(cands, party_abbr == "U") %>%
        slice(1)
    rbind(incumbent, dem, gop, third_party) %>%
        select(party_abbr, party, display_name) %>%
        pivot_wider(names_from=party_abbr,
                    values_from=c(display_name,
                                  party))
}

calc_incumbent_status <- function(cands) {
    incumbent <- filter(cands, is_incumbent)
    if (nrow(incumbent) == 0) {
        "No_Incumbent"
    } else if (incumbent$party_abbr== "D") {
        "Dem_Incumbent"
    } else if (incumbent$party_abbr == "R") {
        "GOP_Incumbent"
    } else {
        "Other_Incumbent"
    }
}

leg_elecs <- read_csv("../../_data/2024/ma_leg_two_party_2008_2024.csv")

margin_model <- stan_glm(dem_margin ~ PVI_N + incumbent_status + pres_elec,
                         data=leg_elecs,
                         family=gaussian(link="identity"),
                         refresh=0)

NUMS_TO_WORDS = rev(c(
    `&`='and',
    `1st`='First',
    `2nd`='Second',
    `3rd`='Third',
    `4th`='Fourth',
    `5th`='Fifth',
    `6th`='Sixth',
    `7th`='Seventh',
    `8th`='Eighth',
    `9th`='Ninth',
    `10th`='Tenth',
    `11th`='Eleventh',
    `12th`='Twelfth',
    `13th`='Thirteenth',
    `14th`='Fourteenth',
    `15th`='Fifteenth',
    `16th`='Sixteenth',
    `17th`='Seventeenth',
    `18th`='Eighteenth',
    `19th`='Nineteenth',
    `20th`='Twentieth',
    `21st`='Twenty-First',
    `22nd`='Twenty-Second',
    `23rd`='Twenty-Third',
    `24th`='Twenty-Fourth',
    `25th`='Twenty-Fifth',
    `26th`='Twenty-Sixth',
    `27th`='Twenty-Seventh',
    `28th`='Twenty-Eighth',
    `29th`='Twenty-Ninth',
    `30th`='Thirtieth',
    `31st`='Thirty-First',
    `32nd`='Thirty-Second',
    `33rd`='Thirty-Third',
    `34th`='Thirty-Fourth',
    `35th`='Thirty-Fifth',
    `36th`='Thirty-Sixth',
    `37th`='Thirty-Seventh'
))

district_pvi <- read_csv("../../_data/2024/ma_state_leg_pvi_2008_2022.csv") %>%
    filter(pvi_year == 2022) %>%
    select(-pvi_year) %>%
    mutate(district = str_replace_all(district, NUMS_TO_WORDS))

districts <- candidates %>%
    nest(.by=c(office, district, district_id),
         .key="candidate") %>%
    mutate(num_candidates = map_int(candidate, nrow),
           incumbent_status = map_chr(candidate, calc_incumbent_status),
           incumbent_status = str_replace(incumbent_status,
                                          "Other_Incumbent",
                                          "Dem_Incumbent"),
           pres_elec = TRUE) %>%
    left_join(district_pvi, by=c("office", "district")) 

dem_margin_format <- function(dem_margin) {
    case_when(
    (dem_margin >= -0.5) & (dem_margin <= 0.5) ~ "EVEN",
    (dem_margin > 0.5) ~ str_glue("D+{round(dem_margin)}"),
    (dem_margin < -0.5) ~ str_glue("R+{round(abs(dem_margin))}")
    )
}

district_preds <- districts %>%
    mutate(pred_dem_margin = predict(margin_model, newdata=districts, type="response"),
           pred_margin_str = dem_margin_format(pred_dem_margin),
           pred_interval = predictive_interval(margin_model, newdata=districts),
           pred_dem_margin_low = pred_interval[,1],
           pred_dem_margin_high = pred_interval[,2]) %>%
    select(-c(pres_elec, pred_interval)) %>%
    arrange(abs(pred_dem_margin))

comp_districts <- district_preds %>%
    filter(abs(pred_dem_margin) < 10)

comp_table <- function(districts) {
    districts %>%
        unnest(candidate) %>%
        group_by(office, district) %>%
        arrange(desc(is_incumbent), party_abbr) %>%
        ungroup() %>%
        select(office,
               district,
               PVI,
               pred_margin_str,
               display_name,
               is_incumbent) %>%
        mutate(row_group = str_glue("**{office} - {district}**<br/>*PVI*: {PVI}, *Model*: {pred_margin_str}"),
               incumbent = if_else(is_incumbent, "Incumbent", "")) %>%
        select(row_group, display_name, incumbent) %>%
        gt(groupname_col="row_group", process_md=TRUE) %>%
        tab_options(column_labels.hidden=TRUE,
                    row_group.background.color = "lightgrey") %>%
        opt_row_striping(row_striping=FALSE)
}
```

```{r}

US_Biden_20 <- 81281502
US_Trump_20 <- 74222593
US_Clinton_16 <- 65853514
US_Trump_16 <- 62984828

dem_percent <- function(dem_2, gop_2, dem_1, gop_1) {
    dem_votes <- dem_2 + dem_1
    gop_votes <- gop_2 + gop_1
    dem_pct <- dem_votes/(dem_votes + gop_votes)
    dem_pct
}

pvi_string <- function(pvi_n) {
    case_when(
        pvi_n <= -0.5 ~ str_c("R+", round(abs(pvi_n))),
        pvi_n >= 0.5 ~ str_c("D+", round(pvi_n)),
        TRUE ~ "EVEN",
    )
}

US_PVI <- dem_percent(
    US_Biden_20,
    US_Trump_20,
    US_Clinton_16,
    US_Trump_16
)

add_pvi_20 <- function(df) {
    df %>%
        mutate(PVI_N = ((dem_percent(Biden_20,
                                     Trump_20,
                                     Clinton_16,
                                     Trump_16) - US_PVI) * 100),
               PVI = pvi_string(PVI_N))
}

district_pcts <- read_csv("https://bwbensonjr.github.io/mapoli/pvi/ma_precincts_districts_pres_2022.csv") 

pct_geometry <- read_sf("https://bwbensonjr.github.io/mapoli/gis/geojson/wardsprecincts2022.geojson") %>%
    select(city_town, ward, precinct)

district_muni_pvi <- function(office, district, dist_pcts, pct_geom) {
    if (office == "State Senate") {
        precincts <- dist_pcts %>%
            filter(State_Senate == district) %>%
            select(-c(State_Rep, US_House, Gov_Council))
    } else {
        precincts <- dist_pcts %>%
            filter(State_Rep == district) %>%
            select(-c(State_Senate, US_House, Gov_Council))
    }
    pct_geom %>%
        right_join(precincts,
                   by=c("city_town", "ward", "precinct")) %>%
        group_by(city_town) %>%
        summarize(Biden_20 = sum(Biden_20),
                  Trump_20 = sum(Trump_20),
                  Clinton_16 = sum(Clinton_16),
                  Trump_16 = sum(Trump_16)) %>%
        st_simplify(dTolerance = 100) %>%
        add_pvi_20() %>%
        select(-c(Biden_20, Trump_20, Clinton_16, Trump_16)) %>%
        mutate(label = str_glue("{city_town}\n{PVI}"))
}

district_muni_map <- function(muni_pvi, map_title) {
    bbox <- st_bbox(muni_pvi)
    osm_tiles <- read_osm(bbox, zoom=NULL)
    (tm_shape(osm_tiles) +
     tm_rgb() +
     tm_shape(muni_pvi) +
     tm_borders(lwd=1) +
     tm_fill(col="PVI_N",
             popup.vars=TRUE,
             palette=c("#FF4949", "white", "#0F4392"),
             midpoint=0,
             legend.show=FALSE,
             alpha=0.7) + 
     tm_text("label") +
     tm_layout(map_title))
}
```

```{r fig.height=6}
district_muni_map(district_muni_pvi("State Representative",
                                    "Third Barnstable",
                                    district_pcts,
                                    pct_geometry),
                  "State Representative - Third Barnstable (D+6)")
```

Several key factors that can be used to model Massachusetts state
legislative outcomes are overall partisanship, which can be
approximated using the Cook Political Report's Partisan Voter Index
(PVI), whether there is a Democratic incumbent, a Republican
incumbent, or no incumbent, and whether the election coincides with a
presidential election.

I applied this model to the State Senate and State Representative
matchups that will be decided on November 5, 2024 and the following
eight districts have an average simulated Democratic margin that is
+/- 10 points from evenly split, making them reasonably competitive
from the model's point of view.

Five of the competitive state legislative elections feature Republican
incumbents in districts that have PVIs between D+1 and D+6, which puts
them in a more competitive position in presidential election years.

```{r}
comp_table(comp_districts %>% filter(incumbent_status == "GOP_Incumbent"))
```

The Ninth Norfolk district features a rematch between now-incumbent
Marcus Vaughn (D-Wrentham) and Kevin Kalkut (D-Norfolk) who faced each
other in 2022 when Shawn Dooley (R-Norfolk) ran against Becca Rausch
(D-Needham) for the Norfolk, Middlesex and Worcester State Senate
Seat. The model gives Vaughn extra points as an incumbent, but the
presidential year advantage for Democrats keeps the model's average
simulated outcome to an R+2 margin.


```{r fig.height=6}
district_muni_map(district_muni_pvi("State Representative",
                                    "Ninth Norfolk",
                                    district_pcts,
                                    pct_geometry),
                  "State Representative - Ninth Norfolk (D+6)")
```

The next set of competitive races are open seats which lean slightly
Republican with PVIs R+1 or R+2.

```{r}
comp_table(comp_districts %>% filter(incumbent_status == "No_Incumbent"))
```

The majority of the model training comes from head-to-head
Democratic/Republican matchups, so we should treat the four districts
that are not D/R head-to-heads as being even more unpredictable than
the +/- 20 or so points that the model has for its 95% confidence
interval.

- The First Worcester and Twenty-Second Middlesex State Representative matchups both have a Republican incumbent facing an Unenrolled candidate, rather than a Democrat.
- The Eighth Bristol State Representative race has Democrat Steven Ouellette, Republican Christopher Thrasher, and *three* Unenrolled candidates in the five candidate mix.
- The only State Senate race on our list, the Third Bristol and Plymouth, features Democrat Joseph Pacheco (no relation to the retiring Senator Marc Pacheco), Republican Kelly Dooner, and independent James Dupont.

```{r fig.height=6}
district_muni_map(district_muni_pvi("State Representative",
                                    "Eighth Bristol",
                                    district_pcts,
                                    pct_geometry),
                  "State Representative - Eighth Bristol (R+2)")
```

